---
title: "Ch2P2_trees"
output: html_document
date: "2025-11-12"
---


## Load Packages 
```{r }

#https://hyphy.org/resources/json-fields.pdf
library(ape)
library(ggtree)
library(tidyverse)
library(tidytext)
library(dplyr)
library(purrr)
library(stringr)
library(scales)
library(glue)
library(tidytext)
library(ggtext)
library(KEGGREST)
#BiocManager::install("sangerseqR")
#devtools::install_version("bold", version="0.9.0", dependencies = FALSE)
#install.packages('datelife', repos = c('https://phylotastic.r-universe.dev', 'https://cloud.r-project.org'))
library(datelife)
library(cowplot)
library(tidyr)
library(ggplot2)
library(viridis)
library(ggrepel)
library(patchwork)
library(forcats)
library(scales)
library(tidyverse)
#install.packages("eulerr")
library(eulerr) 
#install.packages("UpSetR")
library(UpSetR)    
library(phylocomr)
# install.packages("rotl")
# install.packages("ape") 
# install.packages("devtools")
# install.packages("stringr")
# install.packages("datelife")
library(rotl)
library(ape)
library(devtools)
library(rlang)
library(ggpubr)
#install.packages("ggridges")
library(ggridges)
library(gridExtra)
library(gridExtra)
library(grid)


```


## Set filter
```{r, fig 1a}
#set filter

demons <- c(
  "ABAT", "ACOT2", "ADH1A", "ADH1B", "ADH1C", "ADH6", "APOA2", "ATP1A4", "CETP",
  "CH25H", "CTH", "CYP2A6", "CYP2A7", "CYP2C8", "CYP2C9", "CYP3A4", "CYP3A7",
  "DHFR2", "DHRS4L2", "GPAT2", "LDHA", "LDHAL6B", "LPA", "PGAM4", "PLA2G2C",
  "PLA2G4C", "PRKACG", "SLCO1B1", "UGT1A10", "UGT1A3", "UGT1A5", "UGT1A6",
  "UGT1A7", "UGT1A8", "UGT1A9", "UGT2A2", "UGT2B10", "UGT2B11", "UGT2B15",
  "UGT2B28", "UGT2B4", "UGT2B7"
)

exorcise <- function(df, gene_col) {
  df %>%
    mutate(.gene_tmp = toupper(.data[[gene_col]])) %>%
    filter(!.gene_tmp %in% demons) %>%
    select(-.gene_tmp)
}
 


```


## FIGURE 1: Tree Metrics (from iqtree3)
any reference to pathway is a reference to functional category
```{r, tree metrics, COMPLETE}

set.seed(100)


Category_order <- c(
  "Taurine and cysteine metabolism",
  "Retinoid / vitamin A metabolism",
  "Glycine metabolism",
  "Fatty acid & cholesterol metabolism",
  "Bile acid biosynthesis & metabolism"
)

theme_bold <- theme(
  axis.title.x = element_text(size = 15, face = "bold"),
  axis.title.y = element_text(size = 15, face = "bold"),
  axis.text.x  = element_text(face = "bold", size = 13, color = "black"),
  axis.text.y  = element_text(size = 13, face = "bold", color = "black"),
  legend.title = element_text(size = 14, face = "bold", margin = margin(b = 8)), 
  legend.text  = element_text(size = 12, face = "bold"),
  plot.title   = element_text(size = 20, face = "bold", hjust = .5),
  axis.line    = element_line(size = 1.6, color = "black"),
  legend.position = "top",
 panel.grid = element_blank(),
 plot.margin = margin(15, 160, 15, 15),
 plot.subtitle = element_text(hjust = 0.5)
)


use_per_site <- FALSE
min_genes <- 0
max_genes <- Inf
fdr_cutoff <- 0.10
hometree <- read_csv("iqtree_metrics.csv")
hometree <- exorcise(hometree, "gene_id")
gene_metadata <- read_csv("gene_sets_union.csv")
gene_metadata <- exorcise(gene_metadata, "symbol")

df <- left_join(hometree, gene_metadata, by = c("gene_id" = "symbol")) %>%
  mutate(
    across(
      c(
        raw_unconstrained_lnL, raw_consensus_lnL, rf_distance_ml_vs_consensus,
        n_sequences, n_parsimony_informative_sites, prop_invariable_sites,
        raw_internal_length, n_sites, pct_constant_sites
      ),
      as.numeric
    ),
    delta_lnL = raw_unconstrained_lnL - raw_consensus_lnL,
    D = -2 * (raw_consensus_lnL - raw_unconstrained_lnL),
    D_per_pi = if_else(n_parsimony_informative_sites > 0, D / n_parsimony_informative_sites, NA_real_),
    rf_rel = if_else(2 * n_sequences - 6 > 0, rf_distance_ml_vs_consensus / (2 * n_sequences - 6), NA_real_),
    proportion_variable = 1 - prop_invariable_sites,
    effect = if (use_per_site) D_per_pi else D
  )

df <- df %>%
  rename(functional_category = pathway_label)

df <- exorcise(df, "gene_id")

main_df <- df %>%
  filter(
    !tolower(as.character(has_high_gap_warning)) %in% c("true","t","1"),
    !is.na( functional_category ), !is.na(rf_rel), n_sequences >= 10
  ) %>%
  distinct(gene_id,  functional_category , .keep_all = TRUE) %>%
  mutate(
    delta_sign = case_when(
      effect > 0 ~ "positive",
      effect < 0 ~ "negative",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(delta_sign))




gene_sets <- main_df %>%
  distinct(gene_id,  functional_category ) %>%
  mutate(value = 1L) %>%
  pivot_wider(names_from =  functional_category , values_from = value, values_fill = 0L) %>%
  as.data.frame()

pathways <- setdiff(names(gene_sets), "gene_id")
gene_sets[pathways] <- lapply(gene_sets[pathways], as.integer)


pw <- combn(pathways, 2)
table1 <- data.frame(
  pathway1 = pw[1, ],
  pathway2 = pw[2, ],
  genes = I(apply(pw, 2, function(x) gene_sets$gene_id[gene_sets[[x[1]]] == 1 & gene_sets[[x[2]]] == 1])),
  stringsAsFactors = FALSE
) %>%
  filter(lengths(genes) > 0) %>%
  mutate(n_overlap = lengths(genes))

path_df <- main_df %>%
  select(gene_id,  functional_category , effect, rf_rel) %>%
  filter(!is.na(effect), !is.na(rf_rel)) %>%
  add_count( functional_category , name = "n") %>%
  filter(n >= min_genes, n <= max_genes) %>%
  select(-n)

ad_vs_other_categories_full <- path_df %>%
  group_by( functional_category ) %>%
  summarise(
    n_pathway = n(),
    n_other_pathways = sum(path_df$ functional_category  != first( functional_category )),
    median_effect_pathway = median(effect, na.rm = TRUE),
    median_effect_other_pathways = median(path_df$effect[path_df$ functional_category  != first( functional_category )], na.rm = TRUE),
    median_rf_pathway = median(rf_rel, na.rm = TRUE),
    median_rf_other_pathways = median(path_df$rf_rel[path_df$ functional_category  != first( functional_category )], na.rm = TRUE),
    p_ad_effect = if (n() >= 5 && n_other_pathways >= 5) {
      tryCatch(
        as.numeric(kSamples::ad.test(
          effect,
          path_df$effect[path_df$ functional_category  != first( functional_category )],
          method = "asymptotic"
        )$ad[1, " asympt. P-value"]),
        error = function(e) NA_real_
      )
    } else NA_real_,
    p_ad_rf = if (n() >= 5 && n_other_pathways >= 5) {
      tryCatch(
        as.numeric(kSamples::ad.test(
          rf_rel,
          path_df$rf_rel[path_df$ functional_category  != first( functional_category )],
          method = "asymptotic"
        )$ad[1, " asympt. P-value"]),
        error = function(e) NA_real_
      )
    } else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(
    fdr_ad_effect = p.adjust(p_ad_effect, method = "BH"),
    fdr_ad_rf = p.adjust(p_ad_rf, method = "BH")
  ) %>%
  arrange(fdr_ad_rf, fdr_ad_effect, p_ad_rf, p_ad_effect)

ad_vs_other_pathways_sig <- ad_vs_other_categories_full %>%
  filter(if_any(starts_with("fdr_ad_"), ~coalesce(.x <= fdr_cutoff, FALSE)))

plot_df_B <- path_df %>%
  mutate(
    pathway_chr=as.character( functional_category ),
    sig_path=pathway_chr %in% ad_vs_other_pathways_sig$ functional_category 
  )
path_levels_B <- plot_df_B %>%
  group_by(pathway_chr) %>%
  summarise(median_effect=median(effect,na.rm=TRUE),.groups="drop") %>%
  arrange(median_effect) %>%
  pull(pathway_chr)
plot_df_B$pathway_plot <- factor(plot_df_B$pathway_chr, levels=Category_order)

star_df_B <- plot_df_B %>%
  group_by(pathway_plot) %>%
  summarise(
    y_star = max(effect, na.rm = TRUE) + 0.05 * diff(range(plot_df_B$effect, na.rm = TRUE)),
    sig_path = first(sig_path),
    .groups = "drop"
  ) %>%
  filter(sig_path)

rf_lab <- "Relative RF distance\n(0 = identical topology, 1 = maximally different)"
ylab_txt <- if (use_per_site) {
  "Gene-tree support relative to the consensus tree\n(D per informative site)"
} else {
  "Gene-tree support relative to the consensus tree\n(D = 2 × [lnL_gene tree − lnL_consensus tree])"
}



path_stats <- path_df %>%
  group_by( functional_category ) %>%
  summarise(
    n = n(),
    median_effect = median(effect, na.rm = TRUE),
    median_rf = median(rf_rel, na.rm = TRUE),
    .groups = "drop"
  )

pw_pairs <- t(combn(sort(unique(as.character(path_df$ functional_category ))), 2)) %>%
  as.data.frame(stringsAsFactors = FALSE)
names(pw_pairs) <- c("category_1","category_2")

ad_pairwise_categories_full <- pw_pairs %>%
  rowwise() %>%
  mutate(
    n_1 = path_stats$n[match(category_1, path_stats$ functional_category )],
    n_2 = path_stats$n[match(category_2, path_stats$ functional_category )],
    median_effect_1 = path_stats$median_effect[match(category_1, path_stats$ functional_category )],
    median_effect_2 = path_stats$median_effect[match(category_2, path_stats$ functional_category )],
    median_rf_1 = path_stats$median_rf[match(category_1, path_stats$ functional_category )],
    median_rf_2 = path_stats$median_rf[match(category_2, path_stats$ functional_category )],
    p_ad_effect = {
      x <- path_df$effect[path_df$ functional_category  == category_1]
      y <- path_df$effect[path_df$ functional_category  == category_2]
      if (length(x) >= 5 && length(y) >= 5) {
        tryCatch(as.numeric(kSamples::ad.test(x, y, method = "asymptotic")$ad[1, " asympt. P-value"]),
                 error = function(e) NA_real_)
      } else NA_real_
    },
    p_ad_rf = {
      x <- path_df$rf_rel[path_df$ functional_category  == category_1]
      y <- path_df$rf_rel[path_df$ functional_category  == category_2]
      if (length(x) >= 5 && length(y) >= 5) {
        tryCatch(as.numeric(kSamples::ad.test(x, y, method = "asymptotic")$ad[1, " asympt. P-value"]),
                 error = function(e) NA_real_)
      } else NA_real_
    }
  ) %>%
  ungroup() %>%
  mutate(
    fdr_ad_effect = p.adjust(p_ad_effect, method = "BH"),
    fdr_ad_rf = p.adjust(p_ad_rf, method = "BH")
  ) %>%
  arrange(fdr_ad_rf, fdr_ad_effect, p_ad_rf, p_ad_effect)

ad_pairwise_pathways_sig <- ad_pairwise_categories_full %>%
  filter(if_any(starts_with("fdr_ad_"), ~coalesce(.x <= fdr_cutoff, FALSE)))

plot_df_C <- path_df %>%
  mutate(
    pathway_chr=as.character( functional_category ),
    sig_path=pathway_chr %in% c(ad_pairwise_pathways_sig$category_1,ad_pairwise_pathways_sig$category_2)
  )
path_levels_C <- plot_df_C %>%
  group_by(pathway_chr) %>%
  summarise(median_effect=median(effect,na.rm=TRUE),.groups="drop") %>%
  arrange(median_effect) %>%
  pull(pathway_chr)
plot_df_C$pathway_plot <- factor(plot_df_C$pathway_chr, levels=Category_order)

pair_brackets_C <- ad_pairwise_pathways_sig %>%
  mutate(
    x1 = match(category_1, levels(plot_df_C$pathway_plot)),
    x2 = match(category_2, levels(plot_df_C$pathway_plot)),
    y = max(plot_df_C$effect, na.rm = TRUE) +
      row_number() * 0.06 * diff(range(plot_df_C$effect, na.rm = TRUE)),
    x_mid = (x1 + x2) / 2
  )

# Compute max rf_rel per pathway
max_rf <- plot_df_C %>%
  group_by(pathway_plot) %>%
  summarise(max_rf = max(rf_rel, na.rm = TRUE)) %>%
  mutate(x_pos = as.numeric(factor(pathway_plot)))


# Join to bracket table
pair_brackets_C <- pair_brackets_C %>%
  left_join(max_rf, by = c("x1" = "x_pos")) %>%
  rename(max_rf_x1 = max_rf) %>%
  left_join(max_rf, by = c("x2" = "x_pos")) %>%
  rename(max_rf_x2 = max_rf)


# Set bracket height slightly above max to ensure visibility
pair_brackets_C$y_rf <- pmax(pair_brackets_C$max_rf_x1,
                             pair_brackets_C$max_rf_x2) +
  0.03 * diff(range(plot_df_C$rf_rel, na.rm = TRUE))


pBf <- ggplot(plot_df_C, aes(pathway_plot, rf_rel)) +
  geom_violin(aes(fill = sig_path), color = "grey40", scale = "width",
              alpha = 0.9, show.legend = FALSE) +
  geom_jitter(aes(color = effect), width = 0.12, size = 1.3, alpha = 0.65) +
  stat_summary(fun = median, geom = "point", color = "black", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +

  # set bracket segments to use rf_rel range instead of effect range 
  geom_segment(
    data = pair_brackets_C,
    aes(x = x1, xend = x2, y = y_rf, yend = y_rf),
    inherit.aes = FALSE, linewidth = 0.4
  ) +
  geom_segment(
    data = pair_brackets_C,
    aes(x = x1, xend = x1,
        y = y_rf,
        yend = y_rf - 0.015 * diff(range(plot_df_C$rf_rel, na.rm = TRUE))),
    inherit.aes = FALSE, linewidth = 0.4
  ) +
  geom_segment(
    data = pair_brackets_C,
    aes(x = x2, xend = x2,
        y = y_rf,
        yend = y_rf - 0.015 * diff(range(plot_df_C$rf_rel, na.rm = TRUE))),
    inherit.aes = FALSE, linewidth = 0.4
  ) +
  # geom_text(
  #   data = pair_brackets_C,
  #   aes(
  #     x = x_mid,
  #     y = y_rf + 0.015 * diff(range(plot_df_C$rf_rel, na.rm = TRUE))
  #   )
  #   label = "*", inherit.aes = FALSE, size = 5, fontface = "bold") +
  scale_fill_manual(values = c(`FALSE` = "grey92", `TRUE` = "#f4a261"), guide = "none") +
  scale_color_viridis_c(option = "C", name = "Effect size") +
  scale_x_discrete() +
  coord_flip(clip = "off") +

  annotate(
    "text", x = Inf, y = Inf,
    label = if (nrow(ad_pairwise_pathways_sig)) {
      paste0(
        ad_pairwise_pathways_sig$category_1, " vs ", ad_pairwise_pathways_sig$category_2,
        ": effect FDR=", signif(ad_pairwise_pathways_sig$fdr_ad_effect, 2),
        ", RF FDR=", signif(ad_pairwise_pathways_sig$fdr_ad_rf, 2),
        collapse = "\n"
      )
    } else {
      paste0("No pathway pairs with BH-adjusted AD p-value ≤ ", fdr_cutoff, " for effect or RF")
    },
    hjust = .735, vjust = 0.5, size = 3.8, lineheight = 1,  color = "grey15",  
  fontface = "bold"
  ) +

  labs(
    x = "Functional Category",
    y = "Relative RF distance\n(0 = identical topology, 1 = maximally different)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.text.y = element_text(size = 9),
    plot.margin = margin(5.5, 220, 5.5, 5.5)
  )


pBf <- pBf +
  # theme + spacing
  theme_bold +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15))) +
  coord_flip(clip = "off") +
  theme(legend.position = "none")


ggsave("pBViolinFlipped.tiff", width = 12, height = 8)


#Legends

# Temporarily add legend back with new styling for legend extraction


legend_title <- "Gene-tree support relative to the consensus tree\n (D = 2 × [lnL_gene tree – lnL_consensus tree])"

pBf_with_legend <- pBf +
  scale_color_viridis_c(
    option = "C",        # set colors 
    name = legend_title
  ) +
  guides(
    color = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(10, "cm"),
      barheight = unit(0.4, "cm"),
      title.theme = element_text(
        size = 10,
        face = "bold",
        hjust = 0.5
      )
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text  = element_text(size = 10, face = "bold")
  )

legend_B <- get_legend(pBf_with_legend)
legend_B_plot <- ggdraw(legend_B)

ggsave(
  "legend_panelBflipped.tiff",
  legend_B_plot,
  width = 6,
  height = 2,
  dpi = 300,
  bg = "white"
)

# Compute max rf_rel per category for star placement
max_rf_B <- plot_df_B %>%
  group_by(pathway_plot) %>%
  summarise(max_rf = max(rf_rel, na.rm = TRUE))

# Join into star_df_B
star_df_B <- star_df_B %>%
  left_join(max_rf_B, by = "pathway_plot")

# New star height in rf_rel space
star_df_B$y_star_rf <- star_df_B$max_rf +
  0.03 * diff(range(plot_df_B$rf_rel, na.rm = TRUE))



### Build flipped panel A (pAf) 
pAf <- ggplot(plot_df_B, aes(pathway_plot, rf_rel)) +
  geom_violin(aes(fill = sig_path), color = "grey40", scale = "width",
              alpha = 0.9, show.legend = FALSE) +
  geom_jitter(aes(color = effect), width = 0.12, size = 1.3, alpha = 0.65) +
  stat_summary(fun = median, geom = "point", color = "black", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +

  # stars to use y_star_rf 
  geom_text(
    data = star_df_B,
    aes(x = pathway_plot, y = y_star_rf),
    label = "*",
    inherit.aes = FALSE,
    size = 5,
    fontface = "bold"
  ) +

  scale_fill_manual(values = c(`FALSE` = "grey92", `TRUE` = "#f4a261"), guide = "none") +
  scale_color_viridis_c(option = "C", name = "Effect size") +
  scale_x_discrete() +
  coord_flip(clip = "off") +

  annotate(
    "text", x = Inf, y = Inf,
    label = if (nrow(ad_vs_other_pathways_sig)) {
      paste0(
        ad_vs_other_pathways_sig$ functional_category ,
        ": effect FDR=", signif(ad_vs_other_pathways_sig$fdr_ad_effect, 2),
        ", RF FDR=", signif(ad_vs_other_pathways_sig$fdr_ad_rf, 2),
        collapse = "\n"
      )
    } else {
      paste0("No pathways with BH-adjusted AD p-value ≤ ", fdr_cutoff, " for effect or RF")
    },
    hjust = 1, vjust = 0.3, size = 3.9, lineheight = 1, color = "grey15",  
  fontface = "bold"
  ) +

  labs(
    x = "Functional Category",
    y = "Relative RF distance\n(0 = identical topology, 1 = maximally different)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.text.y = element_text(size = 9),
    plot.margin = margin(5.5, 220, 5.5, 5.5)
  ) +

  # finalize formatting
  theme_bold +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15))) +
  coord_flip(clip = "off") +
  theme(legend.position = "none")



### Save flipped panel A 
ggsave("pAViolinFlipped.tiff", pAf, width = 12, height = 8, dpi = 300, bg = "white")


legend_title <- "Gene-tree support relative to the consensus tree\n (D = 2 × [lnL_gene tree – lnL_consensus tree])"

pAf_with_legend <- pAf +
  scale_color_viridis_c(
    option = "C",        #  set colors 
    name = legend_title
  ) +
  guides(
    color = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(10, "cm"),
      barheight = unit(0.4, "cm"),
      title.theme = element_text(
        size = 10,
        face = "bold",
        hjust = 0.5
      )
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text  = element_text(size = 10, face = "bold")
  )

legend_Af <- get_legend(pAf_with_legend)
legend_Af_plot <- ggdraw(legend_Af)

ggsave(
  "legend_panelAflipped.tiff",
  legend_Af_plot,
  width = 6,
  height = 2,
  dpi = 300,
  bg = "white"
)


ad_vs_other_pathways_sig
ad_pairwise_pathways_sig
table1








dev.off()


pA
pB

table1 <- table1 %>%
  rename(
    category1 = pathway1,
    category2 = pathway2
  )

ad_pairwise_categories_sig <- ad_pairwise_pathways_sig %>%
  rename(
    category1 = category_1,
    category2 = category_2
  )

ad_vs_other_categories_sig <- ad_vs_other_pathways_sig %>%
  rename(
    n_category = n_pathway
  )




write_tsv(table1, file = "table1.tsv")
write.csv(ad_pairwise_categories_full, file = "ad_parwise_categories.csv",row.names = FALSE)
write.csv(ad_vs_other_categories_full, file = "ad_vs_other_categories.csv", row.names = FALSE)


```






## FIGURE 2: HyPhy: GARD (Recombination/Breakpoint support analysis)
NOTE: GARD DOES NOT OFFER BRANCH level (therefore per primate species/sequence-lvl information).
It makes inferences about global breakpoints/recombination sites based on modeling global alignment pattern

```{r, GARD}

theme_bold_breakpoints <- theme(
    axis.text = element_text(size = 9, face = "bold", color = "black"),
    axis.title = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 10, hjust = 0),
    strip.text = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid = element_blank()
  )

# GARD 3-panel figure: breakpoints, breakpoint-count distribution, pathway enrichment
min_nseq <- 8
top_n_genes <- 5            # top N per pathway to show in Panel A

#####read input files (adapt filenames if needed) 
gard_res <- read.csv("tgard_summary.csv", stringsAsFactors = FALSE)
gard_res <- exorcise(gard_res, "gene")
function_metadata <- read_csv("gene_sets_union.csv")
gard_res <- merge(gard_res, function_metadata, by.x = "gene", by.y = "symbol", all.x = TRUE)
gene_length <- read_tsv("gene_lengths.tsv")
gene_length <- gene_length %>% 
  group_by(gene_dir) %>% 
  mutate(avg_length = mean(seq_length, na.rm = TRUE)) %>% 
  dplyr::select(c(gene_dir, avg_length)) %>% ungroup()

gard_res <- merge(gard_res, gene_length, by.x = "gene", by.y = "gene_dir") %>% unique() %>%
  mutate(n_seq_gard = map_int(sequence_names, ~ sum(nzchar(str_trim(
    str_split(str_replace_all(str_remove_all(as.character(.x), "^\\s*\\[|\\]\\s*$"), "[\"'\\\\]", ""), "\\s*,\\s*")[[1]]
  ))))) %>%
  filter(n_seq_gard >= 5)

gard_res <- exorcise(gard_res, "gene")

#######Parsing helpers (robust) 
parse_numvec <- function(s) {
  if (is.null(s) || is.na(s) || !nzchar(as.character(s))) return(integer(0))
  if (is.numeric(s)) return(as.numeric(s))
  s_chr <- as.character(s)
  out <- tryCatch(fromJSON(s_chr), error = function(e) NULL)
  if (!is.null(out)) {
    vals <- unlist(out)
    return(suppressWarnings(as.numeric(vals)))
  }
  s2 <- str_remove_all(s_chr, "^\\s*\\[|\\]\\s*$")
  s2 <- str_replace_all(s2, "[\\[\\]]", "")
  toks <- str_split(s2, ",")[[1]] %>% str_trim()
  nums <- suppressWarnings(as.numeric(toks))
  nums[!is.na(nums)]
}

parse_partitions <- function(s) {
  if (is.null(s) || is.na(s) || !nzchar(as.character(s))) return(NULL)
  s_chr <- as.character(s)
  out <- tryCatch(fromJSON(s_chr), error = function(e) NULL)
  if (!is.null(out)) return(out)
  m <- str_match_all(s_chr, "\\[\\s*(\\d+)\\s*,\\s*(\\d+)\\s*\\]")[[1]]
  if (nrow(m) == 0) return(NULL)
  lapply(seq_len(nrow(m)), function(i) c(as.integer(m[i,2]), as.integer(m[i,3])))
}

####Build partition aggregates 
partition_agg <- gard_res %>%
  filter(!is.na(partition_nt_range) & nzchar(as.character(partition_nt_range))) %>%
  mutate(
    .part  = suppressWarnings(as.integer(partition)),
    .start = suppressWarnings(as.integer(str_trim(sub("^\\s*(\\d+)\\s*-.*$", "\\1", as.character(partition_nt_range))))),
    .end   = suppressWarnings(as.integer(str_trim(sub("^.*-(\\s*\\d+)\\s*$", "\\1", as.character(partition_nt_range)))))
  ) %>%
  distinct(gene, .part, .start, .end) %>%
  arrange(gene, .part) %>%
  group_by(gene) %>%
  summarize(
    final_breaks = list(if (n() <= 1) integer(0) else head(.end[order(.part)], -1)),
    n_partitions = n(),
    n_final_breaks = ifelse(n_partitions <= 1, 0L, n_partitions - 1L),
    .groups = "drop"
  )

gard_res <- gard_res %>%
  rename(functional_category = pathway_label)

#####Build compact summary table
gard_compact <- gard_res %>%
  mutate(
    number_of_sequences_numeric = suppressWarnings(as.integer(number_of_sequences)),
    aln_len = suppressWarnings(as.integer(number_of_sites))
  ) %>%
  filter(is.na(number_of_sequences_numeric) | number_of_sequences_numeric >= min_nseq) %>%
  mutate(
    sum_deltaAICc = map_dbl(improvements_deltaAICc, ~ sum(parse_numvec(.x), na.rm = TRUE)),
    max_deltaAICc = map_dbl(improvements_deltaAICc, ~ { v <- parse_numvec(.x); if (length(v)>0) max(v, na.rm=TRUE) else NA_real_ }),
    partition_count_parsed = map_int(partition_bps, ~ { parts <- parse_partitions(.x); if (is.null(parts)) 0L else length(parts) }),
    site_positions_parsed = map(site_positions, parse_numvec),
    site_supports_parsed  = map(site_supports, parse_numvec)
  ) %>%
  mutate(avg_length = ifelse(is.na(as.numeric(avg_length)), suppressWarnings(as.numeric(number_of_sites)), as.numeric(avg_length))) %>%
  distinct(gene,  functional_category , .keep_all = TRUE) %>%
  left_join(partition_agg, by = "gene") %>%
  mutate(
    final_breaks = ifelse(map_lgl(final_breaks, is.null), list(integer(0)), final_breaks),
    n_final_breaks = coalesce(n_final_breaks, 0L),
    n_partitions = coalesce(n_partitions, 1L)
  ) %>%
  mutate(median_deltaAICc = map_dbl(improvements_deltaAICc, ~ { v <- parse_numvec(.x); if(length(v)==0) NA_real_ else median(v, na.rm=TRUE) }))

#Panel A: Histogram of breakpoint counts across genes 
# compute count of final breaks per gene (use gard_compact authoritative)
break_counts <- gard_compact %>%
  transmute(gene,  functional_category , n_final_breaks = as.integer(n_final_breaks)) %>%
  mutate(n_final_breaks = ifelse(is.na(n_final_breaks), 0L, n_final_breaks))

###histo barplot
break_counts_summary <- break_counts %>%
  group_by(n_final_breaks) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  arrange(n_final_breaks)

#title 

     # title = "Panel A — Distribution of final breakpoint counts across genes"

pA <- ggplot(break_counts_summary, aes(x = as.factor(n_final_breaks), y = n_genes)) +
  geom_col(fill = "firebrick", alpha = 0.9) +
  geom_text(aes(label = n_genes), hjust = -0.1, size = 3) +
  coord_flip(clip = "off") +
  labs(x = "Number of GARD final breakpoints (per gene)", y = "Number of genes") +
       # title = "Panel A — Distribution of final breakpoint counts across genes")
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank()
        )

pA

pA <- pA + theme_bold_breakpoints 

pA

ggsave("PanelA_breakpoint_counts.tiff",
  width = 7.5,
  height = 4.2,
  dpi = 300,
  units = "in",
  bg = "white",
  compression = "lzw"
)

################DATA PREP FOR ADV OF PANEL B########################
top_by_pathway <- gard_compact %>%
  filter(!is.na( functional_category )) %>%
  arrange( functional_category , desc(n_final_breaks), desc(median_deltaAICc)) %>%  # use median now
  group_by( functional_category ) %>%
  slice_head(n = top_n_genes) %>%
  ungroup() %>%
  mutate(
    gene_label = paste0(gene, " (n=", n_final_breaks, ")"),
    gene_label = ifelse(is.na(gene_label), gene, gene_label)
  ) %>%
  group_by( functional_category ) %>%
  arrange(desc(n_final_breaks), desc(median_deltaAICc)) %>%
  mutate(gene_ord = factor(gene_label, levels = rev(unique(gene_label)))) %>%
  ungroup()

bp_long <- top_by_pathway %>%
  filter(!is.na(avg_length)) %>%
  mutate(final_breaks = map(final_breaks, ~ as.integer(.x[!is.na(.x)]))) %>%
  dplyr::select(gene, gene_label,  functional_category , avg_length, final_breaks) %>%
  unnest_longer(final_breaks, values_to = "breakpoint") %>%
  mutate(breakpoint = as.integer(breakpoint)) %>%
  filter(!is.na(breakpoint) & breakpoint >= 1 & breakpoint <= avg_length)

lm_df <- top_by_pathway %>%
  filter(!is.na(avg_length)) %>%
  dplyr::select(gene, gene_label,  functional_category , avg_length, n_final_breaks) %>%
  mutate(avg_length = as.numeric(avg_length),
         n_final_breaks = as.numeric(n_final_breaks))

#fitting linear model per path
lm_results <- lm_df %>%
  group_by( functional_category ) %>%
  group_modify(~{
    df <- .x
    if(nrow(df) < 2) {
      tibble(
        slope = NA_real_,
        intercept = NA_real_,
        r2 = NA_real_,
        p_value = NA_real_,
        label = "n<2"
      )
    } else {
      fit <- lm(n_final_breaks ~ avg_length, data = df)
      s <- summary(fit)
      slope <- coef(fit)["avg_length"]
      intercept <- coef(fit)["(Intercept)"]
      r2 <- s$r.squared
      p_value <- s$coefficients["avg_length", "Pr(>|t|)"]  # <- now correct
      tibble(
        slope = slope,
        intercept = intercept,
        r2 = r2,
        p_value = p_value,
        label = sprintf("R²=%.2f\np=%.3g", r2, p_value)
      )
    }
  }, .keep = TRUE) %>%
  ungroup()



# bars top df categories
bars_df <- top_by_pathway %>%
  filter(!is.na(avg_length)) %>%
  left_join(lm_results %>% dplyr::select(functional_category , label), by = "functional_category")

#title

     # title = "Panel B — Breakpoint positions (top genes per category)"

pB <- ggplot() +
  geom_segment(data = bars_df, aes(x = 1, xend = avg_length, y = gene_label, yend = gene_label),
               colour = "grey85", size = 6, lineend = "butt") +
  geom_point(data = bp_long, aes(x = breakpoint, y = gene_label), shape = "|", size = 4, color = "firebrick") +
  geom_text(
  data = lm_results,
  aes(x = Inf, y = Inf, label = label),
  inherit.aes = FALSE,
  hjust = 1.2,   # pulls text into the margin
  vjust = 2.1,
  size = 4
) +
  facet_wrap(~  functional_category , scales = "free_y", ncol = 2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.01))) +
  labs(x = "Position along gene (codons)", y = NULL,
       # title = "Panel B — Breakpoint positions (top genes per category)"
       ) +
 theme(legend.position = "none", panel.grid = element_blank(),plot.margin = margin(10, 40, 10, 10) 
        )

pB

pB <- pB + theme_bold_breakpoints + theme(panel.grid = element_blank(),plot.margin = margin(10, 40, 10, 10)) 

pB

ggsave("PanelB_category_violin.tiff",
  width = 11,
  height = 5,
  dpi = 300,
  units = "in",
  bg = "white",
  compression = "lzw"
)


pB <- pB + theme_bold_breakpoints + theme(axis.text.x = element_text(size = 8, face = "bold", color = "black"), axis.text.y = element_text(size = 9, face = "bold", color = "black"),panel.grid = element_blank(),plot.margin = margin(10, 40, 10, 10)) 

pA <- pA + theme(axis.text = element_text(size = 11, face = "bold", color = "black"))

# Combine with existing panels if present
fig_left  <- (pA / pB)
print(fig_left)


ggsave(
  filename = "Figure2.tiff",
  plot = fig_left,
  width = 14,
  height = 8.5,
  dpi = 300,
  units = "in",
  bg = "white",
  compression = "lzw"
)



```



