---
title: "1_TreeMetrics_HyPhy_Fig3"
output: html_document
date: "2026-03-20"
---

## Load Packages 
```{r }

#https://hyphy.org/resources/json-fields.pdf
library(ape)
library(ggtree)
library(tidyverse)
library(tidytext)
library(dplyr)
library(purrr)
library(stringr)
library(scales)
library(glue)
library(tidytext)
library(ggtext)
library(KEGGREST)
#BiocManager::install("sangerseqR")
#devtools::install_version("bold", version="0.9.0", dependencies = FALSE)
#install.packages('datelife', repos = c('https://phylotastic.r-universe.dev', 'https://cloud.r-project.org'))
library(datelife)
library(cowplot)
library(tidyr)
library(ggplot2)
library(viridis)
library(ggrepel)
library(patchwork)
library(forcats)
library(scales)
library(tidyverse)
#install.packages("eulerr")
library(eulerr) 
#install.packages("UpSetR")
library(UpSetR)    

```

## FIGURE 3A: busted
any reference to pathway is a reference to functional category

Busted kind of indicates most everything would be interesting at as high a level as 'pathway'... i was surprised to find resolution improved at all as we explore some of the other methods (site/branch-site/branch)-level resolutions\n

fubar, fel, and meme (all site/branch-site) point to taurine and cysteine as well as retinoid metabolism\n

absrel (branch-level provides add'l evidence supporting those two pathways may be subject to more positive selection than others)

```{r}

#set filter
demons <- c(
  "ABAT", "ACOT2", "ADH1A", "ADH1B", "ADH1C", "ADH6", "APOA2", "ATP1A4", "CETP",
  "CH25H", "CTH", "CYP2A6", "CYP2A7", "CYP2C8", "CYP2C9", "CYP3A4", "CYP3A7",
  "DHFR2", "DHRS4L2", "GPAT2", "LDHA", "LDHAL6B", "LPA", "PGAM4", "PLA2G2C",
  "PLA2G4C", "PRKACG", "SLCO1B1", "UGT1A10", "UGT1A3", "UGT1A5", "UGT1A6",
  "UGT1A7", "UGT1A8", "UGT1A9", "UGT2A2", "UGT2B10", "UGT2B11", "UGT2B15",
  "UGT2B28", "UGT2B4", "UGT2B7"
)

exorcise <- function(df, gene_col) {
  df %>%
    mutate(.gene_tmp = toupper(.data[[gene_col]])) %>%
    filter(!.gene_tmp %in% demons) %>%
    select(-.gene_tmp)
}

#set theme to add to plots
theme_bold <- theme(
  # Base text 
  text = element_text(family = "sans"),

  # Axis titles 
  axis.title.x = element_text(size = 15, face = "bold", margin = margin(t = 8)),
  axis.title.y = element_text(size = 15, face = "bold", margin = margin(r = 8)),

  #  Axis tick labels
  axis.text.x  = element_text(size = 12, face = "bold", color = "black",
                              margin = margin(t = 4)),
  axis.text.y  = element_text(size = 12, face = "bold", color = "black",
                              margin = margin(r = 4)),

  #  Axis lines 
  axis.line = element_line(size = 0.9, color = "black"),

  # Facet strips 
  strip.text = element_text(size = 14, face = "bold", color = "black",
                            margin = margin(t = 4, b = 4)),
  strip.background = element_rect(fill = "grey92", color = "grey60", linewidth = 0.6),

  #  Legend 
  legend.title = element_text(size = 13, face = "bold"),
  legend.text  = element_text(size = 11, face = "bold"),
  legend.key.size = unit(0.9, "lines"),
  legend.position = "top",

  #  Panel 
  panel.grid = element_blank(),
  panel.background = element_rect(fill = "white", color = NA),

  #  Plot title 
  plot.title = element_text(size = 17, face = "bold", hjust = 0.5,
                            margin = margin(b = 8)),

  #  Margins 
  plot.margin = margin(10, 12, 10, 10)
)




inset_style <- function() {
  list(
    #  Dark, visible jitter points 
    scale_color_manual(values = Category),
    #  Text + theme settings 
    theme(
        text = element_text(family = "sans"),
        
      axis.text.y = element_blank(
        # size = 12, face = "bold", color = "grey15",
        # margin = margin(r = 0.1)
      ),
      axis.text.x = element_text(
        size = 14, face = "bold", color = "grey15",
        margin = margin(t = 1)
      ),
      axis.title.x = element_text(
        size = 16, face = "bold"
      ),
      legend.position = "none",
      axis.line = element_line(color = "black", linewidth = 0.5),
      plot.margin = margin(3, 6, 4, 4),
      panel.grid = element_blank(),
      panel.border = element_rect(color = "black", linewidth = 0.5)
    )
  )
}


Category_order <- c(
  "Taurine and cysteine metabolism",
  "Retinoid / vitamin A metabolism",
  "Glycine metabolism",
  "Fatty acid & cholesterol metabolism",
  "Bile acid biosynthesis & metabolism"
  
  
)



Category <- c("Glycine metabolism" = "#0072B2", "Retinoid / vitamin A metabolism" = "darkorange", "Taurine and cysteine metabolism" = "#009E73", "Bile acid biosynthesis & metabolism" = "red","Fatty acid & cholesterol metabolism" = "purple")




alpha_cut <- 0.10 #### this is alpha for fdr so okay to be 0.1 in this case...
pos_thr   <- 0.10
top_n_genes <- 25

busted_res <- read.csv("tbusted_summary.csv", stringsAsFactors = FALSE) %>%
  merge(
    read_csv("gene_sets_union.csv"),
    by.x = "gene", by.y = "symbol", all.x = TRUE
  ) %>%
  merge(
    read_tsv("gene_lengths.tsv") %>%
      group_by(gene_dir) %>%
      summarise(avg_length = mean(seq_length, na.rm = TRUE), .groups = "drop"),
    by.x = "gene", by.y = "gene_dir", all.x = TRUE
  ) %>%
  unique()

busted_res <- busted_res %>%
  rename(functional_category = pathway_label)

busted_res <- exorcise(busted_res, "gene")

# clean / derive 
busted_clean <- busted_res %>%
  mutate(
    across(
      c(
        pval, lrt,
        omega_positive, omega_neutral, omega_purifying,
        proportion_positive, proportion_neutral, proportion_purifying,
        avg_length
      ),
      as.numeric
    ),
    p_bh = p.adjust(
      ifelse(is.na(pval), NA_real_, pmax(pval, .Machine$double.xmin)),
      method = "BH"
    ),
    branches_clean = str_remove_all(coalesce(branches, ""), "^\\[|\\]$") %>%
      str_replace_all('"', ""),
    number_of_sequences_num = ifelse(
      branches_clean == "",
      NA_integer_,
      str_count(branches_clean, ",\\s*") + 1
    ),
    prop_pos_ok = !is.na(proportion_positive) & proportion_positive >= 0.05,
    prop_pur_ok = !is.na(proportion_purifying) & proportion_purifying >= 0.05,
    omega_pos_plot = pmin(omega_positive, 100),
    omega_pur_ratio = omega_neutral / pmax(omega_purifying, 1e-6),
    omega_pur_plot = pmin(omega_pur_ratio, 100)
  ) %>%
  filter(
    !is.na(functional_category),
    !is.na(gene),
    !is.na(pval),
    !is.na(p_bh),
    !is.na(omega_positive),
    !is.na(omega_neutral),
    !is.na(omega_purifying),
    omega_positive < 100,
    omega_neutral < 100,
    omega_purifying < 100,
    number_of_sequences_num >= 5
  )




# top positive genes 
top_pos <- busted_clean %>%
  filter(
    pval < 0.05,
    p_bh <= alpha_cut,
    !is.na(proportion_positive),
    !is.na(omega_pos_plot),
    omega_positive >= 1,
    omega_neutral <= 1,
    omega_purifying <= omega_neutral,
    prop_pos_ok
  ) %>%
  group_by(functional_category) %>%
  slice_max(order_by = proportion_positive, n = top_n_genes, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    subgroup = "POS",
    x = proportion_positive,
    y = omega_pos_plot
  )

# top purifying genes 
top_pur <- busted_clean %>%
  filter(
    pval < 0.05,
    p_bh <= alpha_cut,
    !is.na(proportion_purifying),
    !is.na(omega_pur_plot),
    omega_purifying <= omega_neutral,
    omega_neutral <= 1,
    omega_positive >= 1,
    prop_pur_ok
  ) %>%
  group_by(functional_category) %>%
  slice_max(order_by = proportion_purifying, n = top_n_genes, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    subgroup = "PUR",
    x = proportion_purifying,
    y = -omega_pur_plot
  )

# combine for single mirrored plot 
plot_dat <- bind_rows(top_pos, top_pur) %>%
  distinct(gene, functional_category, subgroup, .keep_all = TRUE) %>%
  filter(!is.na(x), !is.na(y), between(x, 0, 1))

# equal-space mirrored scaling 
pos_max_raw <- plot_dat %>%
  filter(subgroup == "POS") %>%
  summarise(v = max(y, na.rm = TRUE)) %>%
  pull(v)

neg_max_raw <- plot_dat %>%
  filter(subgroup == "PUR") %>%
  summarise(v = max(abs(y), na.rm = TRUE)) %>%
  pull(v)

if (!is.finite(pos_max_raw) || pos_max_raw <= 0) pos_max_raw <- 1
if (!is.finite(neg_max_raw) || neg_max_raw <= 0) neg_max_raw <- 1

plot_dat <- plot_dat %>%
  mutate(
    y_plot = case_when(
      subgroup == "POS" ~ y / pos_max_raw,
      subgroup == "PUR" ~ y / neg_max_raw,
      TRUE ~ NA_real_
    )
  )

pretty_pos <- pretty(c(0, pos_max_raw), n = 4)
pretty_neg <- pretty(c(0, neg_max_raw), n = 4)

pretty_pos <- pretty_pos[pretty_pos > 0 & pretty_pos <= pos_max_raw]
pretty_neg <- pretty_neg[pretty_neg > 0 & pretty_neg <= neg_max_raw]

neg_breaks <- -(pretty_neg / neg_max_raw)
pos_breaks <-  (pretty_pos / pos_max_raw)

y_breaks <- c(neg_breaks, 0, pos_breaks)
y_labels <- c(as.character(pretty_neg), "0", as.character(pretty_pos))

pos_one_line <- if (pos_max_raw >= 1)  1 / pos_max_raw else NA_real_
neg_one_line <- if (neg_max_raw >= 1) -1 / neg_max_raw else NA_real_

# label all genes with original intensity > 1 on either side 
label_df <- plot_dat %>%
  filter(abs(y) > 1)

#  all-gene inset data 
path_order <- busted_clean %>%
  filter(!is.na(proportion_positive)) %>%
  group_by(functional_category) %>%
  summarise(med_pos = median(proportion_positive, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(med_pos)) %>%
  pull(functional_category)

inset_dat <- busted_clean %>%
  filter(
    !is.na(functional_category),
    !is.na(proportion_positive),
    !is.na(omega_positive),
    is.finite(proportion_positive),
    is.finite(omega_positive)
  ) %>%
  transmute(
    gene,
    functional_category = factor(functional_category, levels = path_order),
    breadth = proportion_positive,
    intensity = omega_positive,
    avg_length = avg_length
  )

# one-vs-rest Wilcoxon tests on ENTIRE distributions 
wilcox_vs_rest <- function(dat, value_col) {
  vals <- dat[[value_col]]
  levs <- levels(dat$functional_category)

  out <- lapply(levs, function(pw) {
    x <- dat %>% filter(functional_category == pw) %>% pull(.data[[value_col]])
    y <- dat %>% filter(functional_category != pw) %>% pull(.data[[value_col]])

    p_raw <- if (length(x) >= 2 && length(y) >= 2) {
      tryCatch(
        wilcox.test(x, y, exact = FALSE)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    }

    tibble(
      functional_category = factor(pw, levels = levs),
      n_pathway = length(x),
      n_rest = length(y),
      med_path = ifelse(length(x) > 0, median(x, na.rm = TRUE), NA_real_),
      med_rest = ifelse(length(y) > 0, median(y, na.rm = TRUE), NA_real_),
      p_raw = p_raw
    )
  }) %>%
    bind_rows() %>%
    mutate(
      p_bh = p.adjust(p_raw, method = "BH"),
      sig_both = !is.na(p_raw) & !is.na(p_bh) & p_raw < 0.05 & p_bh <= 0.10
    )

  out
}

breadth_test <- wilcox_vs_rest(inset_dat, "breadth")
intensity_test <- wilcox_vs_rest(inset_dat, "intensity")

#  symbol placement on breadth distributions 
star_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + 0.03, .groups = "drop") %>%
  left_join(breadth_test %>% select(functional_category, p_raw, p_bh, sig_both), by = "functional_category") %>%
  filter(sig_both)

cross_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + 0.06, .groups = "drop") %>%
  left_join(intensity_test %>% select(functional_category, p_raw, p_bh, sig_both), by = "functional_category") %>%
  filter(sig_both)



inset_dat$functional_category <- factor(inset_dat$functional_category, levels = Category_order)
star_df$functional_category  <- factor(star_df$functional_category,  levels = Category_order)
cross_df$functional_category <- factor(cross_df$functional_category, levels = Category_order)



#dummy plot for legend extraction

legend_plot_obj <- ggplot(
  inset_dat,
  aes(x = functional_category, y = 1, color = functional_category)
) +
  geom_point(size = 3) +
  scale_color_manual(values = Category) +
  labs(color = "Category") +
  guides(color = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 9) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) + inset_style() + theme(legend.position = "right",
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8))



color_legend <- cowplot::get_legend(
  legend_plot_obj +
    guides(color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      ncol = 1, reverse = TRUE       
    )) +
    theme(
      legend.box = "vertical",   
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      panel.border = element_blank()
    ))

#extract and save legend

legend_only_plot <- cowplot::ggdraw(color_legend)

legend_only_plot

ggsave(
  "inset_color_legend.tiff",
  legend_only_plot,
  width = 3.5,
  height = 1.5,
  dpi = 300,
  bg = "white"
)




#title and subtitle

    # title = "Inset: pathway-level distribution tests",
    # subtitle = "* breadth distribution; \u2020 intensity distribution;\nsymbols shown only when p < 0.05 and BH-adjusted p \u2264 0.10",


# inset plot 
p_inset <- ggplot(inset_dat, aes(functional_category, breadth, color = functional_category)) +
  geom_violin(fill = "grey90", color = "grey45", scale = "width", alpha = 0.95) +
  geom_jitter(width = 0.12, size = 0.9, alpha = 0.75, stroke = 0.2,
              show.legend = FALSE) +
  geom_text(
    data = star_df,
    aes(x = functional_category, y = y),
    label = "*",
    inherit.aes = FALSE,
    color = "black",
    size = 10.0
  ) +
  geom_text(
    data = cross_df,
    aes(x = functional_category, y = y),
    label = "\u2020",
    inherit.aes = FALSE,
    color = "black",
    size = 9.0
  ) +
  coord_flip(clip = "off") +
  scale_color_manual(values = Category)+
  scale_y_continuous(
    limits = c(0, max(c(inset_dat$breadth, star_df$y, cross_df$y), na.rm = TRUE) + 0.02),
    name = "Positive-site proportion across all genes"
  ) +
  labs(
    x = NULL,
    color = "Category"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 8),
    axis.text.y = element_text(size = 7),
    axis.text.x = element_text(size = 7),
    legend.position = "none",
    plot.margin = margin(4, 10, 4, 4)
  ) + inset_style()

#title and subtitle 
    # title = "Top BH-significant genes across Categorys",
    # subtitle = paste0(
    #   "Single mirrored descriptive panel with equal visual space above and below zero. ",
    #   "Category identity is encoded by color."
    # ),


#remove pur
plot_dat <- plot_dat %>% filter(subgroup != "PUR")
label_df <- label_df %>% filter(subgroup != "PUR")

#Category <- Category[names(Category) != "PUR"]


# single mirrored main plot 
p_main <- ggplot(plot_dat, aes(x = x, y = y_plot, color = functional_category)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "grey25", linewidth = 0.45) +
  geom_hline(yintercept = pos_one_line, linetype = "dashed", color = "grey55", linewidth = 0.35) +
  geom_hline(yintercept = neg_one_line, linetype = "dashed", color = "grey55", linewidth = 0.35) +
  # geom_point(
  #   data = plot_dat %>% filter(subgroup == "PUR", y_plot > 0),
  #   size = 2.5,
  #   alpha = 0.80
  # ) +
  geom_point(
    data = plot_dat %>% filter(subgroup == "POS"),
    shape = 21,
    size = 3.1,
    stroke = 0.9,
    fill = "white",
    alpha = 1
  ) +

  geom_text_repel(
    data = label_df,
    aes(label = gene),
    size = 2.8,
    max.overlaps = Inf,
    box.padding = 0.30,
    point.padding = 0.20,
    min.segment.length = 0,
    segment.color = "grey55",
    show.legend = FALSE
  ) +
  scale_color_manual(values = Category)+
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::number_format(accuracy = 0.1),
    name = "Breadth of inferred regime (proportion of sites)"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = y_breaks,
    labels = y_labels,
    name = expression(
      atop(
        "Positive regime intensity (" * omega[3] * ")")))+
  # scale_y_continuous(
  #   limits = c(-1, 1),
  #   breaks = y_breaks,
  #   labels = y_labels,
  #   name = expression(
  #     atop(
  #       "Positive regime intensity (" * omega[3] * ")",
  #       "Purifying regime intensity (" * omega[2] / omega[1] * ")"
  #     )
  #    )
  # ) +
  labs(
    # title = "Top BH-significant genes across pathways",
    # subtitle = paste0(
    #   "Single mirrored descriptive panel with equal visual space above and below zero. ",
    #   "Pathway identity is encoded by color."
    # ),
    color = "Category"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5)
  ) 
p_main <- p_main +theme_bold 

# integrate inset plot
p_final <- p_main +
  inset_element(
    p_inset,
    left = 0.64, bottom = 0.58, right = 0.98, top = 0.96,
    align_to = "full"
  )

print(p_final)

busted_supp_main <- busted_clean %>%
  filter(pval < 0.05, p_bh <= alpha_cut) %>%
  transmute(
    gene, functional_category,
    avg_length = round(avg_length, 1),
    number_of_sequences = number_of_sequences_num,
    pval = signif(pval, 4),
    p_bh = signif(p_bh, 4),
    lrt = round(lrt, 4),
    omega_positive = round(omega_positive, 4),
    omega_neutral = round(omega_neutral, 4),
    omega_purifying = round(omega_purifying, 4),
    proportion_positive = round(proportion_positive, 4),
    proportion_neutral = round(proportion_neutral, 4),
    proportion_purifying = round(proportion_purifying, 4)
  ) %>%
  arrange(functional_category, p_bh, pval)

busted_supp_inset <- bind_rows(
  breadth_test %>% mutate(test = "breadth"),
  intensity_test %>% mutate(test = "intensity")
) %>%
  transmute(
    functional_category, test,
    n_pathway, n_rest,
    p_raw = signif(p_raw, 4),
    p_bh = signif(p_bh, 4),
    sig_both,
    med_path = round(med_path, 4),
    med_rest = round(med_rest, 4),
  ) %>%
  arrange(test, p_bh, p_raw)



ggsave("Fig3Busted.A.tiff", plot = p_main, height = 8, width = 15)


ggsave("Fig3InsetBusted.A.tiff", plot = p_inset, height = 8, width = 15)

p_final

labeled_gene_table <- label_df %>%
  transmute(
    gene,
    functional_category,
    subgroup,
    breadth = round(x, 4),
    plotted_intensity = round(abs(y), 4),
    pval = signif(pval, 4),
    p_bh = signif(p_bh, 4),
    lrt = round(lrt, 4),
    omega_positive = round(omega_positive, 4),
    omega_neutral = round(omega_neutral, 4),
    omega_purifying = round(omega_purifying, 4),
    proportion_positive = round(proportion_positive, 4),
    proportion_neutral = round(proportion_neutral, 4),
    proportion_purifying = round(proportion_purifying, 4)
  ) %>%
  arrange(subgroup, functional_category, desc(plotted_intensity), desc(breadth))

## SUPPL TABLE
labeled_gene_table #%>% dplyr::filter("POS" %in% subgroup) %>% pull(gene) %>% unique()

write.csv(labeled_gene_table, file = "labeled_gene_table.csv")


write.csv(breadth_test, file="breadth_test_BUSTED.csv")
write.csv(intensity_test, file = "intensity_test_BUSTED.csv")
write.csv(plot_dat, file = "plot_dat.csv")
write.csv(busted_supp_inset, file = "busted_supp_inset.csv")
write.csv(busted_supp_main, file = "busted_supp_main.csv")

```



## FIGURE 3D: meme
```{r}

set.seed(100)

meme_p_thresh <- 0.05
top_n_per_pathway <- 5
nominal_p_cut <- 0.05

tiny <- 1e-6
omega_cap <- 100

meme_res <- read.csv("tmeme_summary.csv", stringsAsFactors = FALSE)

#### expected MEME columns
#### gene,branch,site_positions,site_pval,site_LRT,site_alpha,site_beta_plus,site_MEMElogl,site_substitution

function_metadata <- read_csv("gene_sets_union.csv")

## merges metadata to MEME data
meme_res <- merge(
  meme_res,
  function_metadata,
  by.x = "gene",
  by.y = "symbol",
  all.x = TRUE
)

meme_res$branch <- gsub("\\|", "_", meme_res$branch)

gene_length <- read_tsv("gene_lengths.tsv") %>%
  mutate(gene_id = toupper(gene_id))

meme_res <- merge(meme_res, gene_length, by.x = "branch", by.y = "gene_id")

meme_res <- exorcise(meme_res, "gene")

## compute avg_length + number_of_sequences_num BEFORE dropping branch
meme_res <- meme_res %>%
  filter(!str_detect(branch, "Node|NODE|root")) %>%
  group_by(gene) %>%
  mutate(
    avg_length = mean(seq_length, na.rm = TRUE),
    number_of_sequences_num = n_distinct(branch) >= 5
  ) %>%
  ungroup() %>%
  mutate(
    ## if seq_length is nucleotide length instead of codon length, change to avg_length / 3
    denom_sites = avg_length
  )

safe_first <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA_real_ else x[1]
}

## ==================================================
## MEME site table
## expand branch-level list columns, then collapse to one row per gene/pathway/site
## so downstream analysis is concordant with FUBAR/FEL
## ==================================================
meme_q_thresh <- 0.10

meme_res <- meme_res %>%
  rename(functional_category = pathway_label)


meme_sites <- meme_res %>%
  mutate(
    across(
      c(
        site_positions, site_pval, site_LRT,
        site_alpha, site_beta_plus, site_MEMElogl,
        site_substitution
      ),
      ~ {
        x <- as.character(.x)
        x <- str_remove_all(x, "^\\s*\\[|\\]\\s*$")
        x <- str_replace_all(x, "\\s*'\\s*|\\s*\"\\s*", "")
        str_trim(x)
      }
    )
  ) %>%
  separate_rows(
    site_positions, site_pval, site_LRT,
    site_alpha, site_beta_plus, site_MEMElogl,
    site_substitution,
    sep = ",\\s*", convert = FALSE
  ) %>%
  mutate(
    site = as.integer(site_positions),
    alpha = as.numeric(site_alpha),
    beta_plus = as.numeric(site_beta_plus),
    pval = as.numeric(site_pval),
    lrt = as.numeric(site_LRT),
    memelogl = as.numeric(site_MEMElogl),
    avg_length = as.numeric(avg_length),
    denom_sites = as.numeric(denom_sites),
    number_of_sequences_num = as.numeric(number_of_sequences_num)
  ) %>%
  filter(!is.na(site), !is.na(functional_category), !is.na(pval)) %>%
  group_by(gene) %>%
  mutate(
    p_adj_branch = p.adjust(pval, method = "BH")
  ) %>%
  ungroup() %>%
  mutate(
    is_sig_pos_site_branch = pval < meme_p_thresh &
      p_adj_branch <= meme_q_thresh &
      beta_plus > alpha,
    omega_branch = pmin(beta_plus / pmax(alpha, tiny), omega_cap)
  ) %>%
group_by(gene, functional_category, site) %>%
  summarise(
    avg_length = safe_first(avg_length),
    denom_sites = safe_first(denom_sites),
    number_of_sequences_num = safe_first(number_of_sequences_num),
    is_sig_pos_site = any(is_sig_pos_site_branch, na.rm = TRUE),
    is_pos_site = is_sig_pos_site,
    omega_site = ifelse(
      any(is_sig_pos_site_branch, na.rm = TRUE),
      max(omega_branch[is_sig_pos_site_branch], na.rm = TRUE),
      NA_real_
    ),
    alpha_site = ifelse(   
      any(is_sig_pos_site_branch, na.rm = TRUE),
      safe_first(alpha[is_sig_pos_site_branch]),
      NA_real_
    ),
    beta_plus_site = ifelse(    
      any(is_sig_pos_site_branch, na.rm = TRUE),
      safe_first(beta_plus[is_sig_pos_site_branch]),
      NA_real_
    ),
    pval_site = ifelse(                                           
      any(is_sig_pos_site_branch, na.rm = TRUE),
      min(pval[is_sig_pos_site_branch], na.rm = TRUE),
      NA_real_
    ),
    lrt_site = ifelse(                  
      any(is_sig_pos_site_branch, na.rm = TRUE),
      max(lrt[is_sig_pos_site_branch], na.rm = TRUE),
      NA_real_
    ),
    size_metric = ifelse(
      any(is_sig_pos_site_branch, na.rm = TRUE),
      pmax(
        pmin(max(omega_branch[is_sig_pos_site_branch], na.rm = TRUE), omega_cap),
        1.01
      ),
      NA_real_
    ),
    .groups = "drop"
  )
## ==================================================
## MAIN MEME PANEL
## significant positive sites only
## top 5 genes per pathway/category by breadth of significant positive sites
## ==================================================

meme_gene_main <- meme_sites %>%
  filter(!is.na(functional_category), !is.na(site)) %>%
  distinct(gene, functional_category, site, .keep_all = TRUE) %>%
  group_by(gene, functional_category) %>%
  summarise(
    avg_length = safe_first(avg_length),
    denom_sites = safe_first(denom_sites),
    n_sites_reported = n_distinct(site),
    n_sig_pos_sites = sum(is_sig_pos_site, na.rm = TRUE),
    prop_sig_pos = ifelse(!is.na(denom_sites) & denom_sites > 0, n_sig_pos_sites / denom_sites, NA_real_),
    sig_pos_omega_intensity = ifelse(
      any(is_sig_pos_site),
      median(omega_site[is_sig_pos_site], na.rm = TRUE),
      NA_real_
    ),
    bar_length = ifelse(
      !is.na(avg_length) & avg_length > 0,
      avg_length,
      as.numeric(n_sites_reported)
    ),
    .groups = "drop"
  ) %>%
  filter(!is.na(prop_sig_pos), is.finite(prop_sig_pos), denom_sites > 0)

top_genes <- meme_gene_main %>%
  filter(n_sig_pos_sites > 0) %>%
  group_by(functional_category) %>%
  arrange(desc(prop_sig_pos), desc(sig_pos_omega_intensity), gene, .by_group = TRUE) %>%
  mutate(plot_rank = row_number()) %>%
  filter(plot_rank <= top_n_per_pathway) %>%
  ungroup()

gene_order <- top_genes %>%
  arrange(functional_category, plot_rank, gene) %>%
  pull(gene) %>%
  unique()

site_plot <- meme_sites %>%
  semi_join(top_genes %>% select(gene, functional_category), by = c("gene", "functional_category")) %>%
  filter(is_sig_pos_site) %>%
  mutate(gene = factor(gene, levels = rev(gene_order)))

gene_bg <- top_genes %>%
  transmute(
    gene = factor(gene, levels = rev(gene_order)),
    functional_category,
    bar_length
  )

#title and subtitle 
    # title = "Significant episodic positive-selection sites across top genes (MEME)",
    # subtitle = paste0(
    #   "Only sites with p \u2264 ", meme_p_thresh,
    #   " and \u03b2+ > \u03b1 are shown. Top 5 genes per pathway are ranked by the breadth of significant positive sites."


  # facet_wrap(~ functional_category, scales = "free_y", ncol = 1) +
  # scale_size_continuous(
  #   name = "Positive-site ω",
  #   range = c(1.2, 4.5)

pMEME <- ggplot() +
  geom_segment(
    data = gene_bg,
    aes(x = 1, xend = bar_length, y = gene, yend = gene),
    color = "grey90",
    linewidth = 2.2
  ) +
  geom_point(
    data = site_plot,
    aes(x = site, y = gene, size = size_metric),
    color = "#D55E00",
    alpha = 0.85
  ) +
  facet_wrap(~ functional_category, scales = "free_y", ncol = 1) +
  scale_size_continuous(
  name = expression(bold("Positive-site strength (" * beta^ "+"/ alpha * ")")),
  range = c(1.2, 4.5)
) +
    scale_color_manual(values = Category)+
  labs(
    x = "Codon position",
    y = NULL,
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(face = "bold")
  )

pMEME <- pMEME + theme_bold 



pMEME

## --------------------------------------------------
## pathway-level inset: full-distribution tests, consistent with FUBAR/FEL
## * = breadth distribution differs from all other pathways/categories
## † = intensity distribution differs from all other pathways/categories
## symbols shown only when raw p < 0.05 AND BH-adjusted p <= 0.10
## --------------------------------------------------

inset_dat <- meme_gene_main %>%
  transmute(
    gene,
    functional_category = as.character(functional_category),
    breadth = prop_sig_pos,
    intensity = sig_pos_omega_intensity
  ) %>%
  filter(!is.na(breadth), is.finite(breadth))

path_order <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(med = median(breadth, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(functional_category)

inset_dat$functional_category <- factor(inset_dat$functional_category, levels = path_order)

wilcox_vs_rest <- function(dat, value_col) {
  levs <- levels(dat$functional_category)

  bind_rows(lapply(levs, function(pw) {
    x <- dat %>%
      filter(functional_category == pw) %>%
      pull(.data[[value_col]])

    y <- dat %>%
      filter(functional_category != pw) %>%
      pull(.data[[value_col]])

    x <- x[is.finite(x)]
    y <- y[is.finite(y)]

    p_raw <- if (length(x) >= 2 && length(y) >= 2) {
      tryCatch(
        wilcox.test(x, y, exact = FALSE)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    }

    tibble(
      functional_category = factor(pw, levels = levs),
      n_path = length(x),
      n_rest = length(y),
      med_path = ifelse(length(x) > 0, median(x, na.rm = TRUE), NA_real_),
      med_rest = ifelse(length(y) > 0, median(y, na.rm = TRUE), NA_real_),
      p_raw = p_raw
    )
  })) %>%
    mutate(
      p_bh = p.adjust(p_raw, method = "BH"),
      sig_both = !is.na(p_raw) & !is.na(p_bh) & p_raw < 0.05 & p_bh <= 0.10
    )
}

breadth_test <- wilcox_vs_rest(inset_dat, "breadth")
intensity_test <- wilcox_vs_rest(inset_dat, "intensity")

pad <- 0.03

star_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + pad, .groups = "drop") %>%
  left_join(
    breadth_test %>% select(functional_category, p_raw, p_bh, sig_both),
    by = "functional_category"
  ) %>%
  filter(sig_both)

cross_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + 2 * pad, .groups = "drop") %>%
  left_join(
    intensity_test %>% select(functional_category, p_raw, p_bh, sig_both),
    by = "functional_category"
  ) %>%
  filter(sig_both)


inset_dat$functional_category <- factor(inset_dat$functional_category, levels = Category_order)
star_df$functional_category  <- factor(star_df$functional_category,  levels = Category_order)
cross_df$functional_category <- factor(cross_df$functional_category, levels = Category_order)

#title and subtitle 
    # title = "Gene-level MEME breadth distributions with pathway-level breadth/intensity tests",
    # subtitle = "* breadth distribution; \u2020 intensity distribution; symbols shown only when raw p < 0.05 and BH-adjusted p \u2264 0.10",

p_inset <- ggplot(inset_dat, aes(functional_category, breadth, color = functional_category)) +
  geom_violin(fill = "grey90", color = "grey45", scale = "width", alpha = 0.95) +
geom_jitter(width = 0.12, size = 0.9, alpha = 0.75, stroke = 0.2,
              show.legend = FALSE) +
  geom_text(
    data = star_df,
    aes(x = functional_category, y = y),
    label = "*",
    inherit.aes = FALSE,
    color = "black",
    size = 10.0
  ) +
  geom_text(
    data = cross_df,
    aes(x = functional_category, y = y),
    label = "\u2020",
    inherit.aes = FALSE,
    color = "black",
    size = 9.0
  ) +
  coord_flip(clip = "off") +
  scale_color_manual(values = Category)+
  scale_y_continuous(
    limits = c(0, max(c(inset_dat$breadth, star_df$y, cross_df$y), na.rm = TRUE) + 0.02),
    name = "Positive-site proportion across all genes"
  ) +
  labs(
    x = NULL,
    color = "Category"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 8),
    axis.text.y = element_text(size = 7),
    axis.text.x = element_text(size = 7),
    legend.position = "none",
    plot.margin = margin(4, 10, 4, 4)
  ) +inset_style()

p_inset_meme <- p_inset +
  theme(
    plot.background = element_rect(fill = "white", color = "grey70", linewidth = 0.4),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(3, 3, 3, 3),
    plot.title = element_text(size = 8, face = "bold"),
    plot.subtitle = element_text(size = 6),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.position = "none"
  )

pMEME_final <- pMEME + inset_element(
  p_inset_meme,
  left = 0.66,
  right = 0.995,
  bottom = 0.08,
  top = 0.44,
  align_to = "panel"
)


ggsave("Fig3MEME.D.tiff",plot = pMEME, height = 8, width = 15)



ggsave("Fig3inset.D.tiff",plot = p_inset, height = 8, width = 15)

pMEME_final

# breadth_test %>% arrange(p_bh, p_raw)
# intensity_test %>% arrange(p_bh, p_raw)

meme_supp_main <- meme_sites %>%
  filter(is_sig_pos_site) %>%
  group_by(gene, functional_category) %>%
  summarise(
    number_of_sequences = safe_first(number_of_sequences_num),
    avg_length          = safe_first(avg_length),
    n_sig_pos_sites     = sum(is_sig_pos_site, na.rm = TRUE),
    prop_sig_pos        = safe_first(n_sig_pos_sites) / safe_first(denom_sites),
    sig_pos_omega_intensity = median(omega_site, na.rm = TRUE),
    n_sites_reported    = safe_first(denom_sites),
    sig_site_positions  = paste(sort(unique(site)), collapse = ", "),
    sig_site_alpha      = paste(round(alpha_site[order(site)], 4), collapse = ", "),
    sig_site_beta_plus  = paste(round(beta_plus_site[order(site)], 4), collapse = ", "),
    sig_site_omegas     = paste(round(omega_site[order(site)], 4), collapse = ", "),
    sig_site_lrts       = paste(round(lrt_site[order(site)], 4), collapse = ", "),
    sig_site_pvalues    = paste(signif(pval_site[order(site)], 4), collapse = ", "),
    .groups = "drop"
  ) %>%
  transmute(
    gene,
    functional_category,
    avg_length          = round(avg_length, 1),
    number_of_sequences,
    n_sites_reported,
    n_sig_pos_sites,
    prop_sig_pos        = round(prop_sig_pos, 4),
    sig_pos_omega_intensity = round(sig_pos_omega_intensity, 4),
    sig_site_positions,
    sig_site_alpha,
    sig_site_beta_plus,
    sig_site_omegas,
    sig_site_lrts,
    sig_site_pvalues
  ) %>%
  arrange(functional_category, desc(prop_sig_pos), desc(sig_pos_omega_intensity))

meme_supp_inset <- bind_rows(
  breadth_test %>% mutate(test = "breadth"),
  intensity_test %>% mutate(test = "intensity")
) %>%
  transmute(
    functional_category, test,
    n_path, n_rest,
    med_path = round(med_path, 4),
    med_rest = round(med_rest, 4),
    p_raw = signif(p_raw, 4),
    p_bh = signif(p_bh, 4),
    sig_both
  ) %>%
  arrange(test, p_bh, p_raw)


write.csv(breadth_test, file="breadth_test_FEL.csv")
write.csv(intensity_test, file = "intensity_test_FEL.csv")
write.csv(meme_supp_inset, file = "meme_supp_inset.csv")
write.csv(meme_supp_main, file = "meme_supp_main.csv")


```







## FIGURE 3C: fel
```{r}


set.seed(100)

fel_p_thresh <- 0.05
fel_q_thresh <- 0.10
alpha_cut <- 0.10
omega_cap <- 100
top_n_per_pathway <- 5

## helpers
safe_first <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA_real_ else x[1]
}

clean_hyphy_vec <- function(x) {
  x <- as.character(x)
  x[is.na(x)] <- ""
  x <- str_remove_all(x, "^\\s*\\[|\\]\\s*$")
  x <- str_replace_all(x, "\\s*'\\s*|\\s*\"\\s*", "")
  x <- str_trim(x)
  x[x == "" | x == "NA"] <- NA_character_
  x
}

clean_num <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("\u2212", "-") %>%
    str_replace_all("\\s+", "") %>%
    str_replace_all("(,)(?=\\d{3}(?:[^\\d]|$))", "") %>%
    str_replace_all("[^0-9eE+\\-\\.]", "") %>%
    na_if("") %>%
    as.numeric()
}

parse_partition_range <- function(x) {
  x <- as.character(x)
  m <- str_match(x, ".*?(\\d+)\\D+(\\d+).*")
  tibble(
    partition_start = suppressWarnings(as.integer(m[, 2])),
    partition_end   = suppressWarnings(as.integer(m[, 3]))
  )
}

## conservative BH when only reported-site p-values are available:
## pad missing unreported codons with p = 1
bh_with_missing_as_one <- function(p_sig, m_total) {
  out <- rep(NA_real_, length(p_sig))
  idx <- which(!is.na(p_sig))

  if (length(idx) == 0) return(out)

  m_total <- as.integer(m_total[1])

  if (is.na(m_total) || m_total < length(idx)) {
    return(out)
  }

  padded <- c(p_sig[idx], rep(1, m_total - length(idx)))
  out[idx] <- p.adjust(padded, method = "BH")[seq_along(idx)]
  out
}

## =========================================================
## read + merge metadata
## =========================================================

fel_res <- read.csv("tfel_summary.csv", stringsAsFactors = FALSE)

function_metadata <- read_csv("gene_sets_union.csv")
fel_res <- merge(
  fel_res,
  function_metadata,
  by.x = "gene_id",
  by.y = "symbol",
  all.x = TRUE
)


fel_res <- fel_res %>%
  rename(functional_category = pathway_label)

gene_length <- read_tsv("gene_lengths.tsv") %>%
  mutate(gene_id = toupper(gene_id))

fel_res <- merge(
  fel_res,
  gene_length,
  by.x = "branch_id",
  by.y = "gene_id"
)




fel_res <- fel_res %>%
  filter(!str_detect(branch_id, regex("node|root", ignore_case = TRUE))) %>%   
  dplyr::select(-any_of("fasta_file")) %>%
  distinct() %>%
  group_by(gene_id) %>%
  mutate(n_seqs_gene = n_distinct(branch_id)) %>%
  ungroup() %>%
  filter(n_seqs_gene >= 5)  

fel_res <- exorcise(fel_res, "gene_id")

## gene-wide average length from branch-level seq_length if present
if ("seq_length" %in% names(fel_res)) {
  fel_res <- fel_res %>%
    group_by(gene_id) %>%
    mutate(avg_length = mean(seq_length, na.rm = TRUE)) %>%
    ungroup() %>%
    dplyr::select(-seq_length) %>%
    distinct()
} else if (!"avg_length" %in% names(fel_res)) {
  fel_res$avg_length <- NA_real_
}

## keep pathway map separate so pathway duplication does not happen
## before site-level q adjustment / collapsing
pathway_map <- fel_res %>%
  dplyr::select(gene_id, functional_category) %>%
  filter(!is.na(functional_category)) %>%
  distinct()

## =========================================================
## reconstruct gene-wide denominators from distinct partitions
## =========================================================

gene_denoms <- fel_res %>%
  dplyr::select(
    gene_id,
    partition,
    partition_codon_range,
    FEL_n_sites,
    avg_length
  ) %>%
  distinct()

range_tbl <- parse_partition_range(gene_denoms$partition_codon_range)

gene_denoms <- bind_cols(gene_denoms, range_tbl) %>%
  mutate(
    partition_width = case_when(
      !is.na(partition_start) & !is.na(partition_end) ~ partition_end - partition_start + 1L,
      TRUE ~ as.numeric(FEL_n_sites)
    )
  ) %>%
  group_by(gene_id) %>%
  summarise(
    gene_n_sites = sum(unique(partition_width[!is.na(partition_width)]), na.rm = TRUE),
    avg_length = safe_first(as.numeric(avg_length)),
    .groups = "drop"
  )

## =========================================================
## site-level parsing from reported FEL hit vectors
## this summary table contains only reported FEL hits,
## not all tested site p-values, so BH is conservative
## =========================================================

site_cols <- c(
  "FEL_sig_sites",
  "FEL_site_pvalues",
  "FEL_site_omegas",
  "FEL_site_LRTs",
  "FEL_site_classes",
  "FEL_omega_coord",
  "FEL_LRTs_site"
)
site_cols <- site_cols[site_cols %in% colnames(fel_res)]

site_core <- fel_res %>%
  dplyr::select(
    gene_id,
    branch_id,
    partition,
    partition_codon_range,
    all_of(site_cols)
  ) %>%
  distinct()

site_core <- bind_cols(site_core, parse_partition_range(site_core$partition_codon_range))

fel_sites_raw <- site_core %>%
  mutate(across(all_of(site_cols), clean_hyphy_vec)) %>%
  separate_rows(all_of(site_cols), sep = ",\\s*", convert = FALSE) %>%
  mutate(
    site_local = clean_num(FEL_sig_sites),
    site_from_coord = coalesce(
      if ("FEL_omega_coord" %in% names(.)) clean_num(FEL_omega_coord) else rep(NA_real_, n()),
      if ("FEL_LRTs_site" %in% names(.)) clean_num(FEL_LRTs_site) else rep(NA_real_, n())
    ),
    site = case_when(
      !is.na(site_local) & !is.na(partition_start) ~ as.integer(partition_start + site_local - 1L),
      !is.na(site_local) ~ as.integer(site_local),
      !is.na(site_from_coord) ~ as.integer(site_from_coord),
      TRUE ~ NA_integer_
    ),
    pvalue = clean_num(FEL_site_pvalues),
    omega = clean_num(FEL_site_omegas),
    lrt = clean_num(FEL_site_LRTs),
    site_class = str_to_lower(str_trim(FEL_site_classes))
  ) %>%
  filter(!is.na(site), !is.na(pvalue))

## collapse to unique gene-wide codons
## when a codon is reported more than once, keep the strongest evidence
fel_sites_gene <- fel_sites_raw %>%
  arrange(gene_id, site, pvalue, desc(omega), desc(lrt)) %>%
  group_by(gene_id, site) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  left_join(gene_denoms, by = "gene_id") %>%
  group_by(gene_id) %>%
  mutate(
    p_adj = bh_with_missing_as_one(pvalue, gene_n_sites)
  ) %>%
  ungroup() %>%
  mutate(
    is_sig_site = !is.na(pvalue) & pvalue < fel_p_thresh &
      !is.na(p_adj) & p_adj <= fel_q_thresh,
    is_pos_site = is_sig_site &
      str_detect(site_class, "positive|divers") &
      !is.na(omega) & omega <= omega_cap,
    is_neg_site = is_sig_site &
      str_detect(site_class, "negative|purif"),
    size_metric = pmax(pmin(omega, omega_cap), 1.01)
  )

## =========================================================
## gene-level breadth summaries
## =========================================================

fel_gene_core <- fel_sites_gene %>%
  group_by(gene_id) %>%
  summarise(
    gene_n_sites = safe_first(gene_n_sites),
    avg_length = safe_first(avg_length),
    n_sig_pos_sites = sum(is_pos_site, na.rm = TRUE),
    n_sig_neg_sites = sum(is_neg_site, na.rm = TRUE),
    prop_sig_pos = ifelse(gene_n_sites > 0, n_sig_pos_sites / gene_n_sites, NA_real_),
    median_pos_omega = ifelse(
      n_sig_pos_sites > 0,
      median(omega[is_pos_site], na.rm = TRUE),
      NA_real_
    ),
    bar_length = ifelse(!is.na(avg_length) & avg_length > 0, avg_length, gene_n_sites),
    .groups = "drop"
  ) %>%
  filter(!is.na(prop_sig_pos), gene_n_sites > 0)

fel_gene <- pathway_map %>%
  inner_join(fel_gene_core, by = "gene_id")

## =========================================================
## top genes per pathway by breadth
## =========================================================



top_by_breadth <- fel_gene %>%
  filter(n_sig_pos_sites > 0) %>%
  group_by(functional_category) %>%
  arrange(
    desc(prop_sig_pos),
    desc(n_sig_pos_sites),
    desc(median_pos_omega),
    gene_id,
    .by_group = TRUE
  ) %>%
  mutate(plot_rank = row_number()) %>%
  filter(plot_rank <= top_n_per_pathway) %>%
  ungroup()

## =========================================================
## main breadth panel
## =========================================================

gene_order <- top_by_breadth %>%
  arrange(functional_category, plot_rank, gene_id) %>%
  pull(gene_id) %>%
  unique()

site_plot <- fel_sites_gene %>%
  filter(is_pos_site) %>%
  inner_join(pathway_map, by = "gene_id") %>%
  semi_join(
    top_by_breadth %>% dplyr::select(gene_id, functional_category),
    by = c("gene_id", "functional_category")
  ) %>%
  mutate(gene_id = factor(gene_id, levels = rev(gene_order)))

gene_bg <- top_by_breadth %>%
  transmute(
    gene_id = factor(gene_id, levels = rev(gene_order)),
    functional_category,
    bar_length
  )

#title and subtitle 
    # # title = "FEL positive-selection codons across genes ranked by breadth",
    # # subtitle = paste0(
    # #   "Genes are ranked within pathway by fraction of gene codons that are significant positive FEL sites. ",
    # #   "Shown codons satisfy raw p < ", fel_p_thresh,
    # #   ", BH q ≤ ", fel_q_thresh,
    # #   ", positive class, and ω ≤ ", omega_cap, "."
    # )

pFEL_breadth <- ggplot() +
  geom_segment(
    data = gene_bg,
    aes(x = 1, xend = bar_length, y = gene_id, yend = gene_id),
    color = "grey90",
    linewidth = 2.2
  ) +
  geom_point(
    data = site_plot,
    aes(x = site, y = gene_id, size = size_metric),
    color = "#D55E00",
    alpha = 0.85
  ) +
  facet_wrap(~ functional_category, scales = "free_y", ncol = 1) +
  scale_size_continuous(
    name = "Positive-site ω",
    range = c(1.2, 4.5)
  ) +
    scale_color_manual(values = Category)+
  labs(
    x = "Gene-wide codon position",
    y = NULL,
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(face = "bold")
  ) 

pFEL_breadth <- pFEL_breadth + theme_bold

## =========================================================
## pathway-level inset: full-distribution tests, consistent with FUBAR
## * = breadth distribution differs from all other pathways
## † = intensity distribution differs from all other pathways
## symbols shown only when raw p < 0.05 AND BH-adjusted p <= 0.10
## =========================================================

inset_dat <- fel_gene %>%
  transmute(
    gene_id,
    functional_category = as.character(functional_category),
    breadth = prop_sig_pos,
    intensity = median_pos_omega
  ) %>%
  filter(!is.na(breadth), is.finite(breadth))

path_order <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(med = median(breadth, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(functional_category)

inset_dat$functional_category <- factor(inset_dat$functional_category, levels = Category_order)

wilcox_vs_rest <- function(dat, value_col) {
  levs <- levels(dat$functional_category)

  bind_rows(lapply(levs, function(pw) {
    x <- dat %>%
      filter(functional_category == pw) %>%
      pull(.data[[value_col]])

    y <- dat %>%
      filter(functional_category != pw) %>%
      pull(.data[[value_col]])

    x <- x[is.finite(x)]
    y <- y[is.finite(y)]

    p_raw <- if (length(x) >= 2 && length(y) >= 2) {
      tryCatch(
        wilcox.test(x, y, exact = FALSE)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    }

    tibble(
      functional_category = factor(pw, levels = levs),
      n_path = length(x),
      n_rest = length(y),
      med_path = ifelse(length(x) > 0, median(x, na.rm = TRUE), NA_real_),
      med_rest = ifelse(length(y) > 0, median(y, na.rm = TRUE), NA_real_),
      p_raw = p_raw
    )
  })) %>%
    mutate(
      p_bh = p.adjust(p_raw, method = "BH"),
      sig_both = !is.na(p_raw) & !is.na(p_bh) & p_raw < 0.05 & p_bh <= 0.10
    )
}

breadth_test <- wilcox_vs_rest(inset_dat, "breadth")
intensity_test <- wilcox_vs_rest(inset_dat, "intensity")

pad <- 0.03

star_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + pad, .groups = "drop") %>%
  left_join(
    breadth_test %>% select(functional_category, p_raw, p_bh, sig_both),
    by = "functional_category"
  ) %>%
  filter(sig_both)

cross_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(breadth, na.rm = TRUE) + 2 * pad, .groups = "drop") %>%
  left_join(
    intensity_test %>% select(functional_category, p_raw, p_bh, sig_both),
    by = "functional_category"
  ) %>%
  filter(sig_both)

upper_lim <- max(c(inset_dat$breadth, star_df$y, cross_df$y), na.rm = TRUE) + 0.02


inset_dat$functional_category <- factor(inset_dat$functional_category, levels = Category_order)
star_df$functional_category  <- factor(star_df$functional_category,  levels = Category_order)
cross_df$functional_category <- factor(cross_df$functional_category, levels = Category_order)

#title and subtitle
    # title = "Gene-level FEL breadth distributions with pathway-level breadth/intensity tests",
    # subtitle = "* breadth distribution; \u2020 intensity distribution; symbols shown only when raw p < 0.05 and BH-adjusted p \u2264 0.10",

p_inset <- ggplot(inset_dat, aes(functional_category, breadth, color = functional_category)) +
  geom_violin(fill = "grey90", color = "grey45", scale = "width", alpha = 0.95) +
geom_jitter(width = 0.12, size = 0.9, alpha = 0.75, stroke = 0.2,
              show.legend = FALSE) +
geom_text(
  data = star_df,
  aes(x = functional_category, y = y),
  label = "*",
  inherit.aes = FALSE,
  color = "black",
  size = 10.0
) +
geom_text(
  data = cross_df,
  aes(x = functional_category, y = y),
  label = "\u2020",
  inherit.aes = FALSE,
  color = "black",
  size = 9.0
) +
  coord_flip(clip = "off") +
  scale_color_manual(values = Category)+
  scale_y_continuous(
  limits = c(0, 0.25),
  breaks = seq(0, 0.25, by = 0.05),
  name = "Positive-site proportion across all genes"
)+
  labs(
    x = NULL,
    color = "Category"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 8),
    axis.text.y = element_text(size = 7),
    axis.text.x = element_text(size = 7),
    legend.position = "none",
    plot.margin = margin(4, 10, 4, 4)
  )+inset_style()

p_inset_fel <- p_inset +
  theme(
    plot.background = element_rect(fill = "white", color = "grey70", linewidth = 0.4),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(3, 3, 3, 3),
    plot.title = element_text(size = 8, face = "bold"),
    plot.subtitle = element_text(size = 6),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.position = "none"
  )

pFEL_final <- pFEL_breadth + inset_element(
  p_inset_fel,
  left = 0.66,
  right = 0.995,
  bottom = 0.08,
  top = 0.44,
  align_to = "panel"
)

pFEL_final

ggsave("Fig3FEL.C.tiff",plot = pFEL_breadth, height = 8, width = 15)

ggsave("Fig3FELinset.C.tiff",plot = p_inset, height = 8, width = 15)

fel_supp_table <- fel_gene %>%
  filter(n_sig_pos_sites > 0) %>%
  left_join(
    fel_sites_gene %>%
      filter(is_pos_site) %>%
      inner_join(pathway_map, by = "gene_id") %>%
      group_by(gene_id, functional_category) %>%
      summarise(
        sig_site_positions = paste(sort(unique(site)), collapse = ", "),
        sig_site_omegas    = paste(round(omega[order(site)], 4), collapse = ", "),
        sig_site_lrts      = paste(round(lrt[order(site)], 4), collapse = ", "),
        sig_site_pvalues   = paste(signif(pvalue[order(site)], 4), collapse = ", "),
        sig_site_padj      = paste(signif(p_adj[order(site)], 4), collapse = ", "),
        .groups = "drop"
      ),
    by = c("gene_id", "functional_category")
  ) %>%
  transmute(
    gene = gene_id,
    functional_category,
    avg_length = round(avg_length, 1),
    gene_n_sites,
    n_sig_pos_sites,
    n_sig_neg_sites,
    prop_sig_pos = round(prop_sig_pos, 4),
    median_pos_omega = round(median_pos_omega, 4),
    sig_site_positions,
    sig_site_omegas,
    sig_site_lrts,
    sig_site_pvalues,
    sig_site_padj
  ) %>%
  arrange(functional_category, desc(prop_sig_pos), desc(median_pos_omega))

fel_supp_inset <- bind_rows(
  breadth_test %>% mutate(test = "breadth"),
  intensity_test %>% mutate(test = "intensity")
) %>%
  transmute(
    functional_category, test,
    n_path, n_rest,
    med_path = round(med_path, 4),
    med_rest = round(med_rest, 4),
    p_raw = signif(p_raw, 4),
    p_bh = signif(p_bh, 4),
    sig_both
  ) %>%
  arrange(test, p_bh, p_raw)

write.csv(breadth_test, file="breadth_test_FEL.csv")
write.csv(intensity_test, file = "intensity_test_FEL.csv")
write.csv(fel_supp_inset, file = "fel_supp_inset.csv")
write.csv(fel_supp_table, file = "fel_supp_main.csv")




```



## FIGURE 3B: fubar

```{r, fig 3 panel B}

set.seed(100)

fubar_pp_thresh <- 0.90
top_n_per_pathway <- 5
alpha_cut <- 0.10
pos_thr <- 0.10

tiny <- 1e-6
omega_cap <- 100

fubar_res <- read.csv("tfubar_summary.csv", stringsAsFactors = FALSE)

#### original
#### program,gene,branch,GTR,sites,alpha,beta,beta-alpha,Prob[alpha>beta],Prob[alpha<beta],BayesFactor[alpha<beta]


#### let's adjust
fubar_res <- rename(
  fubar_res,
  beta_minus_alpha = beta.alpha,
  P_alphaGTbeta = Prob.alpha.beta.,
  P_betaGTalpha = Prob.alpha.beta..1,
  BayesFactor_betaGTalpha = BayesFactor.alpha.beta.
)



function_metadata <- read_csv("gene_sets_union.csv")

## merges metadata to FUBAR data
fubar_res <- merge(
  fubar_res,
  function_metadata,
  by.x = "gene",
  by.y = "symbol",
  all.x = TRUE
)

fubar_res$branch <- gsub("\\|", "_", fubar_res$branch)

gene_length <- read_tsv("gene_lengths.tsv") %>%
  mutate(gene_id = toupper(gene_id))

fubar_res <- merge(fubar_res, gene_length, by.x = "branch", by.y = "gene_id")

## compute avg_length + number_of_sequences_num BEFORE dropping branch
fubar_res <- fubar_res %>%
  filter(!str_detect(branch, "Node|root")) %>%
  group_by(gene) %>%
  mutate(
    avg_length = mean(seq_length, na.rm = TRUE),
    number_of_sequences_num = n_distinct(branch) >= 5
  ) %>%
  ungroup() %>%
  dplyr::select(-c(branch, fasta_file, GTR, seq_length)) %>%
  unique()


fubar_res <- exorcise(fubar_res, "gene")

fubar_res <- fubar_res %>%
  rename(functional_category = pathway_label)

fubar_sites <- fubar_res %>%
  mutate(
    across(
      c(
        sites, alpha, beta, beta_minus_alpha,
        P_alphaGTbeta, P_betaGTalpha, BayesFactor_betaGTalpha
      ),
      ~ {
        x <- as.character(.x)
        x <- str_remove_all(x, "^\\s*\\[|\\]\\s*$")
        x <- str_replace_all(x, "\\s*'\\s*|\\s*\"\\s*", "")
        str_trim(x)
      }
    )
  ) %>%
  separate_rows(
    sites, alpha, beta, beta_minus_alpha,
    P_alphaGTbeta, P_betaGTalpha, BayesFactor_betaGTalpha,
    sep = ",\\s*", convert = FALSE
  ) %>%
  mutate(
    site = as.integer(sites),
    alpha = as.numeric(alpha),
    beta = as.numeric(beta),
    beta_minus_alpha2 = beta - alpha,
    pp_pos = as.numeric(P_betaGTalpha),
    pp_neg = as.numeric(P_alphaGTbeta),
    bf_pos = as.numeric(BayesFactor_betaGTalpha),
    avg_length = as.numeric(avg_length),
    number_of_sequences_num = as.numeric(number_of_sequences_num),

    ## explicit significant-site definitions used everywhere downstream
    is_sig_pos_site = !is.na(pp_pos) & pp_pos >= fubar_pp_thresh & beta_minus_alpha2 > 1,
    is_sig_neg_site = !is.na(pp_neg) & pp_neg >= fubar_pp_thresh & beta_minus_alpha2 < -1,

    ## aliases for backward compatibility
    is_pos_site = is_sig_pos_site,
    is_neg_site = is_sig_neg_site,

    omega_site = pmin(beta / pmax(alpha, tiny), omega_cap),
    size_metric = pmax(pmin(beta / pmax(alpha, tiny), omega_cap), 1.01)
  ) %>%
  filter(!is.na(site), !is.na(functional_category))

safe_first <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA_real_ else x[1]
}

## ==================================================
## MAIN FUBAR PANEL
## significant positive sites only
## top 5 genes per pathway by breadth of significant positive sites
## ==================================================

fubar_gene_main <- fubar_sites %>%
  filter(!is.na(functional_category), !is.na(site)) %>%
  distinct(gene, functional_category, site, .keep_all = TRUE) %>%
  group_by(gene, functional_category) %>%
  summarise(
    avg_length = safe_first(avg_length),
    n_sites_reported = n_distinct(site),
    n_sig_pos_sites = sum(is_sig_pos_site, na.rm = TRUE),
    prop_sig_pos = ifelse(n_sites_reported > 0, n_sig_pos_sites / n_sites_reported, NA_real_),
    sig_pos_omega_intensity = ifelse(
      any(is_sig_pos_site),
      median(omega_site[is_sig_pos_site], na.rm = TRUE),
      NA_real_
    ),
    bar_length = ifelse(
      !is.na(avg_length) & avg_length > 0,
      avg_length,
      as.numeric(n_sites_reported)
    ),
    .groups = "drop"
  ) %>%
  filter(!is.na(prop_sig_pos), is.finite(prop_sig_pos), n_sites_reported > 0)

top_genes <- fubar_gene_main %>%
  filter(n_sig_pos_sites > 0) %>%
  group_by(functional_category) %>%
  arrange(desc(prop_sig_pos), desc(sig_pos_omega_intensity), gene, .by_group = TRUE) %>%
  mutate(plot_rank = row_number()) %>%
  filter(plot_rank <= top_n_per_pathway) %>%
  ungroup()

gene_order <- top_genes %>%
  arrange(functional_category, plot_rank, gene) %>%
  pull(gene) %>%
  unique()

site_plot <- fubar_sites %>%
  semi_join(top_genes %>% select(gene, functional_category), by = c("gene", "functional_category")) %>%
  filter(is_sig_pos_site) %>%
  mutate(gene = factor(gene, levels = rev(gene_order)))

gene_bg <- top_genes %>%
  transmute(
    gene = factor(gene, levels = rev(gene_order)),
    functional_category,
    bar_length
  )

#title and subtitle
    # title = "Significant pervasive positive-selection sites across top genes (FUBAR)",
    # subtitle = paste0(
    #   "Only sites with P(β>α) ≥ ", fubar_pp_thresh,
    #   " and β−α > 1 are shown. Top 5 genes per pathway are ranked by the breadth of significant positive sites."
    # )


pFUBAR <- ggplot() +
  geom_segment(
    data = gene_bg,
    aes(x = 1, xend = bar_length, y = gene, yend = gene),
    color = "grey90",
    linewidth = 2.2
  ) +
  geom_point(
    data = site_plot,
    aes(x = site, y = gene, size = size_metric),
    color = "#D55E00",
    alpha = 0.85
  ) +
    scale_color_manual(values = Category)+
  facet_wrap(~ functional_category, scales = "free_y", ncol = 1) +
  scale_size_continuous(name = "Positive-site ω (β/α)", range = c(1.2, 4.5)) +
  labs(
    x = "Codon position",
    y = NULL,
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(face = "bold")
  )
 
pFUBAR <- pFUBAR + theme_bold

## --------------------------------------------------
## BREADTH / FREQUENCY INSET ONLY
## --------------------------------------------------
## --------------------------------------------------
## INSET: pathway-level full-distribution tests
## breadth = prop_sig_pos
## intensity = sig_pos_omega_intensity
## mark only if raw p < 0.05 AND BH-adjusted p <= 0.10
## * = breadth
## † = intensity
## --------------------------------------------------

fubar_gene_inset <- fubar_sites %>%
  filter(!is.na(functional_category), !is.na(site)) %>%
  distinct(gene, functional_category, site, .keep_all = TRUE) %>%
  group_by(gene, functional_category) %>%
  summarise(
    n_sites_reported = n_distinct(site),
    n_sig_pos_sites = sum(is_sig_pos_site, na.rm = TRUE),
    avg_length = safe_first(avg_length),
    prop_sig_pos = ifelse(n_sites_reported > 0, n_sig_pos_sites / n_sites_reported, NA_real_),
    sig_pos_omega_intensity = ifelse(
      any(is_sig_pos_site),
      median(omega_site[is_sig_pos_site], na.rm = TRUE),
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  filter(!is.na(prop_sig_pos), is.finite(prop_sig_pos), n_sites_reported > 0)

inset_dat <- fubar_gene_inset %>%
  transmute(
    gene,
    functional_category = as.character(functional_category),
    breadth = prop_sig_pos,
    intensity = sig_pos_omega_intensity
  )

path_order <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(med = median(breadth, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(functional_category)

inset_dat$functional_category <- factor(inset_dat$functional_category, levels = path_order)

wilcox_vs_rest <- function(dat, value_col) {
  levs <- levels(dat$functional_category)

  bind_rows(lapply(levs, function(pw) {
    x <- dat %>% filter(functional_category == pw) %>% pull(.data[[value_col]])
    y <- dat %>% filter(functional_category != pw) %>% pull(.data[[value_col]])

    x <- x[is.finite(x)]
    y <- y[is.finite(y)]

    p_raw <- if (length(x) >= 2 && length(y) >= 2) {
      tryCatch(
        wilcox.test(x, y, exact = FALSE)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    }

    tibble(
      functional_category = factor(pw, levels = levs),
      n_path = length(x),
      n_rest = length(y),
      med_path = ifelse(length(x) > 0, median(x, na.rm = TRUE), NA_real_),
      med_rest = ifelse(length(y) > 0, median(y, na.rm = TRUE), NA_real_),
      p_raw = p_raw
    )
  })) %>%
    mutate(
      p_bh = p.adjust(p_raw, method = "BH"),
      sig_both = !is.na(p_raw) & !is.na(p_bh) & p_raw < 0.05 & p_bh <= 0.10
    )
}

breadth_test <- wilcox_vs_rest(inset_dat, "breadth")
intensity_test <- wilcox_vs_rest(inset_dat, "intensity")

pad <- max(100 * inset_dat$breadth, na.rm = TRUE) * 0.08 + 0.5

star_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(100 * breadth, na.rm = TRUE) + pad, .groups = "drop") %>%
  left_join(breadth_test %>% select(functional_category, p_raw, p_bh, sig_both), by = "functional_category") %>%
  filter(sig_both)

cross_df <- inset_dat %>%
  group_by(functional_category) %>%
  summarise(y = max(100 * breadth, na.rm = TRUE) + 2 * pad, .groups = "drop") %>%
  left_join(intensity_test %>% select(functional_category, p_raw, p_bh, sig_both), by = "functional_category") %>%
  filter(sig_both)

upper_lim <- max(c(100 * inset_dat$breadth, star_df$y, cross_df$y, 1), na.rm = TRUE)

inset_dat$functional_category <- factor(inset_dat$functional_category, levels = Category_order)
star_df$functional_category  <- factor(star_df$functional_category,  levels = Category_order)
cross_df$functional_category <- factor(cross_df$functional_category, levels = Category_order)


#title and subtitle 
    # title = "Distributions of significant positive-selection breadth by pathway",
    # subtitle = "* breadth distribution; \u2020 intensity distribution; symbols shown only when raw p < 0.05 and BH-adjusted p \u2264 0.10",

p_inset <- ggplot(inset_dat, aes(functional_category, 100 * breadth, color = functional_category)) +
  geom_violin(fill = "grey88", color = "grey40", scale = "width", alpha = 0.9) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    fill = NA,
    color = "grey20",
    linewidth = 0.35
  ) +
  geom_jitter(width = 0.12, size = 0.9, alpha = 0.75, stroke = 0.2,
              show.legend = FALSE) +
  geom_text(
    data = star_df,
    aes(functional_category, y),
    label = "*",
    inherit.aes = FALSE,
    color = "black",
    size = 10.0
  ) +
  geom_text(
    data = cross_df,
    aes(functional_category, y),
    label = "\u2020",
    inherit.aes = FALSE,
    color = "black",
    size = 9.0
  ) +
  coord_flip(clip = "off") +
  scale_color_manual(values = Category) +
  scale_y_continuous(
    name = "Significant positive sites (% of evaluated codon sites)",
    limits = c(0, upper_lim),
    expand = expansion(mult = c(0.02, 0.18))
  ) +
  labs(
    x = NULL,
    color = "Category"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(size = 9, face = "bold"),
    plot.subtitle = element_text(size = 7),
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 6),
    plot.margin = margin(4, 14, 4, 4),
    legend.position = "none"
  ) +inset_style()

pFUBAR_final <- pFUBAR + inset_element(
  p_inset,
  left = 0.58, right = 0.98,
  bottom = 0.61, top = 0.90,
  align_to = "panel"
)

pFUBAR_final

### SUPPL TABLEs
fubar_supp_table <- fubar_gene_main %>%
  filter(n_sig_pos_sites > 0) %>%
  left_join(
    fubar_sites %>%
      filter(is_sig_pos_site) %>%
      group_by(gene, functional_category) %>%
      summarise(
        number_of_sequences = safe_first(number_of_sequences_num),
        sig_site_positions = paste(sort(unique(site)), collapse = ", "),
        sig_site_alpha     = paste(round(alpha[order(site)], 4), collapse = ", "),
        sig_site_beta      = paste(round(beta[order(site)], 4), collapse = ", "),
        sig_site_omegas    = paste(round(omega_site[order(site)], 4), collapse = ", "),
        sig_site_pp_pos    = paste(round(pp_pos[order(site)], 4), collapse = ", "),
        sig_site_bf_pos    = paste(round(bf_pos[order(site)], 4), collapse = ", "),
        .groups = "drop"
      ),
    by = c("gene", "functional_category")
  ) %>%
  left_join(
    fubar_sites %>%
      group_by(gene, functional_category) %>%
      summarise(
        n_sig_neg_sites = sum(is_sig_neg_site, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("gene", "functional_category")
  ) %>%
  transmute(
    gene, functional_category,
    avg_length = round(avg_length, 1),
    number_of_sequences,
    n_sites_reported,
    n_sig_pos_sites,
    n_sig_neg_sites,
    prop_sig_pos = round(prop_sig_pos, 4),
    sig_pos_omega_intensity = round(sig_pos_omega_intensity, 4),
    sig_site_positions,
    sig_site_alpha,
    sig_site_beta,
    sig_site_omegas,
    sig_site_pp_pos,
    sig_site_bf_pos
  ) %>%
  arrange(functional_category, desc(prop_sig_pos), desc(sig_pos_omega_intensity))

fubar_supp_inset <- bind_rows(
  breadth_test %>% mutate(test = "breadth"),
  intensity_test %>% mutate(test = "intensity")
) %>%
  transmute(
    functional_category, test,
    n_path, n_rest,
    med_path = round(med_path, 4),
    med_rest = round(med_rest, 4),
    p_raw = signif(p_raw, 4),
    p_bh = signif(p_bh, 4),
    sig_both
  ) %>%
  arrange(test, p_bh, p_raw)

ggsave("Fig3FUBAR.B.tiff",plot = pFUBAR, height = 8, width = 15)


ggsave("Fig3FUBARinset.B.tiff",plot = p_inset, height = 8, width = 15)

# write.csv(full_gene_data_for_plotted_pathways, file = "full_gene_data_for_plotted_pathways.csv")


write.csv(breadth_test, file="breadth_test_FUBAR.csv")
write.csv(intensity_test, file = "intensity_test_FUBAR.csv")
write.csv(fubar_supp_inset,file = "fubar_supp_inset.csv")
write.csv(fubar_supp_table, file = "fubar_supp_table.csv")


```



## sure UGT1s are postively selected... we already knew that, but what about the places where pathways overlap... we know they do in this dataset, and it's also part of the reason we chose some of these genes by path in the first place anyways

## now the results here show that some of the biggest significant overlap is between pathways where bile acid path features are major partners...

## retinoid path where UGT1s are taking center stage of course we show as the driver, but we also show significant overlap between this path and fatty acid metabolism...

## many of the top 10 drivers are not reported as positively selected in the primate literature yet with the exception of CD36,OSBPL5,LPL---- the other 7 are generally under-reported LRPAP1, STARD3, ABCB11, LCAT, and also not a strong primate-specific one for SCARB1, LIPA, or ABCG8... and none of these are UGT genes already canonically discussed in the primate lit

```{r}

## =========================================================
## BRIDGE / PATHWAY-PAIR OVERLAP ANALYSIS
## - tests whether specific pathway pairs share more
##   selection-supported genes than expected
## - BH-corrects permutation p-values across all pathway pairs
## - extracts the actual genes driving significant pairs
## =========================================================

set.seed(100)

## -----------------------------
## user-tunable settings
## -----------------------------
min_methods_support <- 2   # gene must show positive-site support in at least this many methods
pair_alpha_bh <- 0.10      # BH cutoff for declaring a significant pathway pair
n_perm <- 5000             # increase if needed

## -----------------------------
## helper
## -----------------------------
collapse_unique_chr <- function(x, sep = " | ") {
  x <- sort(unique(as.character(x[!is.na(x)])))
  if (length(x) == 0) NA_character_ else paste(x, collapse = sep)
}

## -----------------------------
##  pathway membership across methods
##    keep all observed gene-pathway memberships
## -----------------------------
pathway_membership <- bind_rows(
  fubar_gene_main %>%
    transmute(
      gene = as.character(gene),
      functional_category = as.character(functional_category)
    ),
  fel_gene %>%
    transmute(
      gene = as.character(gene_id),
      functional_category = as.character(functional_category)
    ),
  meme_gene_main %>%
    transmute(
      gene = as.character(gene),
      functional_category = as.character(functional_category)
    )
) %>%
  filter(!is.na(gene), !is.na(functional_category), gene != "", functional_category != "") %>%
  distinct()



## -----------------------------
##  method-level support
##    "support" = at least one positive site in that method
## -----------------------------
fubar_support <- fubar_gene_main %>%
  group_by(gene) %>%
  summarise(
    has_fubar = any(n_sig_pos_sites > 0, na.rm = TRUE),
    fubar_n_sig = max(n_sig_pos_sites, na.rm = TRUE),
    fubar_breadth = max(prop_sig_pos, na.rm = TRUE),
    .groups = "drop"
  )

fel_support <- fel_gene %>%
  group_by(gene_id) %>%
  summarise(
    has_fel = any(n_sig_pos_sites > 0, na.rm = TRUE),
    fel_n_sig = max(n_sig_pos_sites, na.rm = TRUE),
    fel_breadth = max(prop_sig_pos, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(gene = gene_id)

meme_support <- meme_gene_main %>%
  group_by(gene) %>%
  summarise(
    has_meme = any(n_sig_pos_sites > 0, na.rm = TRUE),
    meme_n_sig = max(n_sig_pos_sites, na.rm = TRUE),
    meme_breadth = max(prop_sig_pos, na.rm = TRUE),
    .groups = "drop"
  )

gene_support <- pathway_membership %>%
  distinct(gene) %>%
  left_join(fubar_support, by = "gene") %>%
  left_join(fel_support,   by = "gene") %>%
  left_join(meme_support,  by = "gene") %>%
  mutate(
    has_fubar = coalesce(has_fubar, FALSE),
    has_fel   = coalesce(has_fel, FALSE),
    has_meme  = coalesce(has_meme, FALSE),
    fubar_n_sig = coalesce(fubar_n_sig, 0),
    fel_n_sig   = coalesce(fel_n_sig, 0),
    meme_n_sig  = coalesce(meme_n_sig, 0),
    support_n = rowSums(across(c(has_fubar, has_fel, has_meme))),
    support_methods = pmap_chr(
      list(has_fubar, has_fel, has_meme),
      ~ {
        out <- c()
        if (..1) out <- c(out, "FUBAR")
        if (..2) out <- c(out, "FEL")
        if (..3) out <- c(out, "MEME")
        if (length(out) == 0) NA_character_ else paste(out, collapse = " | ")
      }
    )
  )

## -----------------------------
##  restrict to genes with recurrent support
## -----------------------------
supported_genes <- gene_support %>%
  filter(support_n >= min_methods_support)

pm_sup <- pathway_membership %>%
  semi_join(supported_genes, by = "gene") %>%
  distinct()

## gene-level pathway degree among supported genes
gene_degree <- pm_sup %>%
  count(gene, name = "n_pathways") %>%
  arrange(desc(n_pathways), gene)

## -----------------------------
## enumerate category pairs
## -----------------------------
path_levels <- sort(unique(pm_sup$functional_category))

if (length(path_levels) < 2) {
  stop("Fewer than 2 pathway labels remain after support filtering; no pairwise test can be run.")
}

path_pairs <- as_tibble(t(combn(path_levels, 2)), .name_repair = "minimal")
colnames(path_pairs) <- c("path1", "path2")

## -----------------------------
##  observed overlap counts
##    count genes that belong to BOTH pathways in the pair
## -----------------------------
pair_overlap_results <- path_pairs %>%
  rowwise() %>%
  mutate(
    obs = pm_sup %>%
      group_by(gene) %>%
      summarise(
        has1 = any(functional_category == path1),
        has2 = any(functional_category == path2),
        .groups = "drop"
      ) %>%
      filter(has1, has2) %>%
      nrow()
  ) %>%
  ungroup()

## -----------------------------
##  degree-preserving permutation null
##    preserves how many pathway labels each gene has
## -----------------------------
degree_tbl <- pm_sup %>%
  count(gene, name = "deg")

perm_counts <- replicate(n_perm, {
  perm_pm <- degree_tbl %>%
    rowwise() %>%
    mutate(
      functional_category = list(sample(path_levels, size = deg, replace = FALSE))
    ) %>%
    unnest(functional_category) %>%
    ungroup()

  path_pairs %>%
    rowwise() %>%
    mutate(
      n_shared = perm_pm %>%
        group_by(gene) %>%
        summarise(
          has1 = any(functional_category == path1),
          has2 = any(functional_category == path2),
          .groups = "drop"
        ) %>%
        filter(has1, has2) %>%
        nrow()
    ) %>%
    pull(n_shared)
})

## -----------------------------
## 7) permutation p-values + BH correction
## -----------------------------
pair_overlap_results <- pair_overlap_results %>%
  mutate(
    p_perm = vapply(seq_len(n()), function(i) {
      mean(perm_counts[i, ] >= obs[i])
    }, numeric(1)),
    p_bh = p.adjust(p_perm, method = "BH"),
    sig_bh = p_bh <= pair_alpha_bh
  ) %>%
  arrange(p_bh, desc(obs), path1, path2)

pair_overlap_results

## -----------------------------
##  extract the actual genes driving each pair
##    for ALL pairs
## -----------------------------
pair_driver_genes <- pair_overlap_results %>%
  select(path1, path2, obs, p_perm, p_bh, sig_bh) %>%
  rowwise() %>%
  do({
    pair_row <- .
    driver_tbl <- pm_sup %>%
      group_by(gene) %>%
      summarise(
        has1 = any(functional_category == pair_row$path1),
        has2 = any(functional_category == pair_row$path2),
        pathway_list = collapse_unique_chr(functional_category),
        .groups = "drop"
      ) %>%
      filter(has1, has2) %>%
      select(-has1, -has2) %>%
      left_join(supported_genes, by = "gene") %>%
      left_join(gene_degree, by = "gene") %>%
      mutate(
        path1 = pair_row$path1,
        path2 = pair_row$path2,
        obs = pair_row$obs,
        p_perm = pair_row$p_perm,
        p_bh = pair_row$p_bh,
        sig_bh = pair_row$sig_bh
      ) %>%
      arrange(
        desc(support_n),
        desc(n_pathways),
        desc(fel_n_sig),
        desc(fubar_n_sig),
        desc(meme_n_sig),
        gene
      )

    driver_tbl
  }) %>%
  ungroup()

pair_driver_genes

## -----------------------------
##  keep only significant pairs
## -----------------------------
significant_pairs <- pair_overlap_results %>%
  filter(sig_bh)

significant_pair_driver_genes <- pair_driver_genes %>%
  semi_join(significant_pairs, by = c("path1", "path2"))

significant_pair_driver_genes


## =========================================================
##  rank bridge genes within each significant pair
##     using observed quantities only
##
##     Formal significance remains at the category-pair level:
##       - pair_overlap_results$p_bh
##
##     Gene ordering within each significant pair describes:
##       1) support_n
##       2) median FUBAR/FEL positive-site breadth
##       3) n_pathways
##       4) FEL / FUBAR / MEME significant-site counts
## =========================================================

significant_pair_driver_genes_ranked <- significant_pair_driver_genes %>%
  rowwise() %>%
  mutate(
    median_breadth_ff = {
      x <- c(
        if (isTRUE(has_fubar)) fubar_breadth else NA_real_,
        if (isTRUE(has_fel))   fel_breadth   else NA_real_
      )
      x <- x[is.finite(x)]
      if (length(x) == 0) NA_real_ else median(x)
    }
  ) %>%
  ungroup() %>%
  group_by(path1, path2) %>%
  arrange(
    desc(support_n),
    desc(median_breadth_ff),
    desc(n_pathways),
    desc(fel_n_sig),
    desc(fubar_n_sig),
    desc(meme_n_sig),
    gene,
    .by_group = TRUE
  ) %>%
  mutate(driver_rank_within_pair = row_number()) %>%
  ungroup()

significant_pair_driver_genes_ranked

## =========================================================
##  compact summary:
##     one row per significant pair + ranked gene list
## =========================================================

significant_pair_driver_summary <- significant_pair_driver_genes_ranked %>%
  group_by(path1, path2, obs, p_perm, p_bh) %>%
  summarise(
    n_driver_genes = n(),
    driver_genes = paste(gene, collapse = ", "),
    top_driver_genes = paste(gene[order(driver_rank_within_pair)][1:min(10, n())], collapse = ", "),
    support_methods_seen = collapse_unique_chr(unlist(strsplit(
      paste(support_methods[!is.na(support_methods)], collapse = " | "),
      " \\| "
    ))),
    .groups = "drop"
  ) %>%
  arrange(p_bh, desc(obs), path1, path2)

significant_pair_driver_summary

## =========================================================
##  quick view
##     top ranked bridge genes + high-breadth candidates
## =========================================================

label_breadth_cut <- 0.10
n_label_per_pair <- 8

top_drivers_per_significant_pair <- significant_pair_driver_genes_ranked %>%
  filter(driver_rank_within_pair <= 10) %>%
  mutate(
    high_breadth_candidate = median_breadth_ff > label_breadth_cut
  ) %>%
  select(
    path1, path2, driver_rank_within_pair,
    gene, support_n, support_methods, n_pathways, pathway_list,
    median_breadth_ff, high_breadth_candidate,
    fubar_n_sig, fel_n_sig, meme_n_sig,
    fubar_breadth, fel_breadth, meme_breadth,
    p_bh
  ) %>%
  arrange(p_bh, path1, path2, driver_rank_within_pair)

top_drivers_per_significant_pair %>% View()

## =========================================================
##   bridge figure
##     - only BH-significant pathway pairs are shown
##     - only genes with support in >= min_methods_support methods
##     - x-axis = median FUBAR/FEL breadth
##     - labels show top-ranked genes and any genes with
##       median_breadth_ff > 10%
## =========================================================

pair_label_df <- significant_pairs %>%
  mutate(
    pair_label = paste0(
      path1, " \u2194 ", path2,
      "\nobs = ", obs,
      ", BH-adjusted p = ", formatC(p_bh, format = "f", digits = 3)
    )
  ) %>%
  select(path1, path2, pair_label)

bridge_fig_df <- significant_pair_driver_genes_ranked %>%
  left_join(pair_label_df, by = c("path1", "path2")) %>%
  mutate(
    support_n = factor(support_n, levels = sort(unique(support_n))),
    high_breadth_flag = median_breadth_ff > label_breadth_cut,
    top_flag = driver_rank_within_pair <= n_label_per_pair,
    label_flag = top_flag | high_breadth_flag,
    label_reason = case_when(
      top_flag & high_breadth_flag ~ "top_rank_and_high_breadth",
      top_flag ~ "top_rank",
      high_breadth_flag ~ "high_breadth_only",
      TRUE ~ "unlabeled"
    )
  ) %>%
  arrange(p_bh, path1, path2, driver_rank_within_pair) %>%
  mutate(
    pair_gene = paste(pair_label, gene, sep = "___"),
    pair_gene = factor(pair_gene, levels = rev(unique(pair_gene)))
  )

x_max_bridge <- max(bridge_fig_df$median_breadth_ff, na.rm = TRUE)
if (!is.finite(x_max_bridge) || x_max_bridge <= 0) {
  x_max_bridge <- 0.05
}

#title and subtitle 

    # title = "Bridge genes contributing to significant pathway-pair overlaps",
    # subtitle = paste0(
    #   "Only pathway pairs with BH-adjusted permutation significance are shown. ",
    #   "All displayed genes were supported in >= ", min_methods_support,
    #   " methods and annotated to both pathways in the pair. ",
    #   "Within each pair, genes are ordered by tool support first, then by median FUBAR/FEL breadth. ",
    #   "Top ", n_label_per_pair, " genes per pair are labeled, and any additional genes with median FUBAR/FEL breadth > ",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " are also labeled."
    # ),



#theme for suppl figure

theme_bold1 <-  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    #  Panel 
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),

    #  Axis titles 
    axis.title.x = element_text(
      size = 16, face = "bold",
      margin = margin(t = 10)
    ),
    axis.title.y = element_blank(),
    
    #  Axis text  
    axis.text.x = element_text(
      size = 14, face = "bold", color = "black",
      margin = margin(t = 4)
    ),
    axis.text.y = element_text(size = 12, face = "bold", color = "black"
    ),
    #  Axis lines 
    axis.line = element_line(size = 0.9, color = "black"),

    # Facet strips 
    strip.text = element_text(
      size = 16, face = "bold",
      margin = margin(t = 4, b = 4)
    ),
    strip.background = element_rect(
      fill = "grey94",
      color = "grey55",
      linewidth = 0.7
    ),

    # Legend 
    legend.position = "none",
    legend.title = element_blank(),
    legend.text  = element_text(),
    
    #  Plot title & caption 
    plot.title = element_text(
      size = 18, face = "bold", hjust = 0.5,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      size = 8.5, hjust = 0,
      margin = margin(t = 10)
    )
)



p_bridge_clean <- ggplot(
  bridge_fig_df,
  aes(x = median_breadth_ff, y = pair_gene)
) +
  geom_segment(
    aes(x = 0, xend = median_breadth_ff, yend = pair_gene),
    color = "grey88",
    linewidth = 0.6
  ) +
  geom_point(
    aes(fill = support_n, alpha = label_flag),
    shape = 21,
    size = 3.3,
    color = "black",
    stroke = 0.3
  ) +
geom_point(
  aes(fill = support_n, alpha = label_flag),
  shape = 21,
  size = 3.3,
  color = "black",
  stroke = 0.3
) +
# ggrepel::geom_text_repel(
#   data = bridge_fig_df %>% filter(!label_flag),
#   aes(
#     label = gene),
#   size = 4.0,
#   color = "black",
#   segment.color = "grey20",
#   segment.size = 0.3,
#   force = 0.25,                
#   max.overlaps = Inf,
#   direction = "y",
#   hjust = 0,
#   nudge_x = x_max_bridge * 0.25) +
ggrepel::geom_text_repel(
  data = bridge_fig_df %>% filter(label_flag),
  aes(label = gene),
  size = 5,
  color = "grey20",
  segment.color = "grey35",
  segment.size = 0.4,
  direction = "y",
  hjust = 0,
  nudge_x = x_max_bridge * 0.025,
  box.padding = 0.2,
  point.padding = 0.15,
  max.overlaps = Inf,
  min.segment.length = 0
)+
  facet_wrap(~ pair_label, scales = "free_y", ncol = 1) +
scale_y_discrete(
  labels = function(x) sub("^.*___", "", x)
) +
  scale_x_continuous(
    limits = c(0, x_max_bridge * 1.08),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  scale_fill_grey(
    start = 0.75,
    end = 0.25,
    name = "Methods supporting\npositive selection"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.35),
    guide = "none"
  ) +
  labs(
    x = "Median gene-level positive-site breadth across FUBAR and FEL\n(proportion of codon sites inferred as positively selected; descriptive ordering only)",
    y = NULL,
    # title = "Bridge genes contributing to significant pathway-pair overlaps",
    # subtitle = paste0(
    #   "Only pathway pairs with BH-adjusted permutation significance are shown. ",
    #   "All displayed genes were supported in >= ", min_methods_support,
    #   " methods and annotated to both pathways in the pair. ",
    #   "Within each pair, genes are ordered by tool support first, then by median FUBAR/FEL breadth. ",
    #   "Top ", n_label_per_pair, " genes per pair are labeled, and any additional genes with median FUBAR/FEL breadth > ",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " are also labeled."
    # ),
    # caption = paste0(
    #   "Pair-level significance comes from the BH-corrected permutation test; the x-axis is a descriptive gene-ordering quantity, not a significance threshold. ",
    #   "The >",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " label rule is used only to highlight candidate genes for literature follow-up."
    # )
  ) +
  theme_minimal(base_size = 11
  ) 

p_bridge_clean

p_bridge_clean <- p_bridge_clean + theme_bold1

p_bridge_clean

ggsave("supple_lollipop.tiff",plot = p_bridge_clean, height = 12, width = 18)

bridge_fig_df$pair_label

#subsets for plotting individually

bile_fatty_df <- bridge_fig_df %>%
  filter(pair_label == "Bile acid biosynthesis & metabolism ↔ Fatty acid & cholesterol metabolism\nobs = 32, BH-adjusted p = 0.000")

bile_retinoid_df <- bridge_fig_df %>%
  filter(pair_label == "Bile acid biosynthesis & metabolism ↔ Retinoid / vitamin A metabolism\nobs = 17, BH-adjusted p = 0.007")

glycine_df <- bridge_fig_df %>%
  filter(pair_label == "Glycine metabolism ↔ Taurine and cysteine metabolism\nobs = 15, BH-adjusted p = 0.005")


#new themes for each plot 


theme_bold2 <-  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    #  Panel 
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),

    #  Axis titles 
    axis.title.x = element_blank(
    ),
    axis.title.y = element_blank(),
    
    #  Axis text  
    axis.text.x = element_text(
      size = 18, face = "bold", color = "black"
    ),
    axis.text.y = element_text(size = 15, face = "bold", color = "black"
    ),
    #  Axis lines 
    axis.line = element_line(size = 0.9, color = "black"),

    #  Facet strips 
    strip.text = element_text(
      size = 24, face = "bold",
      margin = margin(t = 4, b = 4)
    ),
    strip.background = element_rect(
      fill = "grey94",
      color = "grey55",
      linewidth = 0.7
    ),

    #  Legend 
    legend.position = "none",
    legend.title = element_blank(),
    legend.text  = element_text(),
    
    #  Plot title & caption 
    plot.title = element_text(
      size = 18, face = "bold", hjust = 0.5,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      size = 8.5, hjust = 0,
      margin = margin(t = 10)
    )
)


theme_bold3 <-  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    #  Panel 
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),

    # Axis titles
    axis.title.x = element_text(
      size = 24, face = "bold"
    ),
    axis.title.y = element_blank(),
    
    # Axis text 
    axis.text.x = element_text(
      size = 18, face = "bold", color = "black"
    ),
    axis.text.y = element_text(size = 15, face = "bold", color = "black"
    ),
    # Axis lines 
    axis.line = element_line(size = 0.9, color = "black"),

    # Facet strips 
    strip.text = element_text(
      size = 24, face = "bold",
      margin = margin(t = 4, b = 4)
    ),
    strip.background = element_rect(
      fill = "grey94",
      color = "grey55",
      linewidth = 0.7
    ),

    #  Legend
    legend.position = "none",
    legend.title = element_blank(),
    legend.text  = element_text(),
    
    #  Plot title & caption
    plot.title = element_text(
      size = 18, face = "bold", hjust = 0.5,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      size = 8.5, hjust = 0,
      margin = margin(t = 10)
    )
)
  
  
#plot

p_bridge_BileFatty <- ggplot(
  bile_fatty_df,
  aes(x = median_breadth_ff, y = pair_gene)
) +
  geom_segment(
    aes(x = 0, xend = median_breadth_ff, yend = pair_gene),
    color = "grey88",
    linewidth = 1.6
  ) +
  geom_point(
    aes(fill = support_n, alpha = label_flag),
    shape = 21,
    size = 4,
    color = "black",
    stroke = 0.3
  ) +
geom_point(
  aes(fill = support_n, alpha = label_flag),
  shape = 21,
  size = 4,
  color = "black",
  stroke = 0.3
) +
# ggrepel::geom_text_repel(
#   data = bridge_fig_df %>% filter(!label_flag),
#   aes(
#     label = gene),
#   size = 4.0,
#   color = "black",
#   segment.color = "grey20",
#   segment.size = 0.3,
#   force = 0.25,                
#   max.overlaps = Inf,
#   direction = "y",
#   hjust = 0,
#   nudge_x = x_max_bridge * 0.25) +
ggrepel::geom_text_repel(
  data = bile_fatty_df %>% filter(label_flag),
  aes(label = gene),
  size = 6,
  color = "grey20",
  segment.color = "grey35",
  segment.size = 0.4,
  direction = "y",
  hjust = 0,
  nudge_x = x_max_bridge * 0.025,
  box.padding = 0.2,
  point.padding = 0.15,
  max.overlaps = Inf,
  min.segment.length = 0
)+
  facet_wrap(~ pair_label, scales = "free_y", ncol = 1, labeller = labeller(
    pair_label = function(x) sub("\n.*$", "", x)
  )) +
scale_y_discrete(
  labels = function(x) sub("^.*___", "", x)
) +
  scale_x_continuous(
    limits = c(0, x_max_bridge * 1.08),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  scale_fill_grey(
    start = 0.75,
    end = 0.25,
    name = "Methods supporting\npositive selection"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.35),
    guide = "none"
  ) +
  labs(
    # x = "Median gene-level positive-site breadth across FUBAR and FEL\n(proportion of codon sites inferred as positively selected; descriptive ordering only)",
    # y = NULL,
    # title = "Bridge genes contributing to significant pathway-pair overlaps",
    # subtitle = paste0(
    #   "Only pathway pairs with BH-adjusted permutation significance are shown. ",
    #   "All displayed genes were supported in >= ", min_methods_support,
    #   " methods and annotated to both pathways in the pair. ",
    #   "Within each pair, genes are ordered by tool support first, then by median FUBAR/FEL breadth. ",
    #   "Top ", n_label_per_pair, " genes per pair are labeled, and any additional genes with median FUBAR/FEL breadth > ",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " are also labeled."
    # ),
    # caption = paste0(
    #   "Pair-level significance comes from the BH-corrected permutation test; the x-axis is a descriptive gene-ordering quantity, not a significance threshold. ",
    #   "The >",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " label rule is used only to highlight candidate genes for literature follow-up."
    # )
  ) +
  theme_minimal(base_size = 11
  ) 

p_bridge_BileFatty

p_bridge_BileFatty <- p_bridge_BileFatty + theme_bold2

p_bridge_BileFatty

ggsave("BILEFattysupple_lollipop.tiff",plot = p_bridge_BileFatty, height = 8, width = 18)



p_bridge_Bile_retinoid <- ggplot(
  bile_retinoid_df,
  aes(x = median_breadth_ff, y = pair_gene)
) +
  geom_segment(
    aes(x = 0, xend = median_breadth_ff, yend = pair_gene),
    color = "grey88",
    linewidth = 1.6
  ) +
  geom_point(
    aes(fill = support_n, alpha = label_flag),
    shape = 21,
    size = 4,
    color = "black",
    stroke = 0.3
  ) +
geom_point(
  aes(fill = support_n, alpha = label_flag),
  shape = 21,
  size = 4,
  color = "black",
  stroke = 0.3
) +
# ggrepel::geom_text_repel(
#   data = bridge_fig_df %>% filter(!label_flag),
#   aes(
#     label = gene),
#   size = 4.0,
#   color = "black",
#   segment.color = "grey20",
#   segment.size = 0.3,
#   force = 0.25,                
#   max.overlaps = Inf,
#   direction = "y",
#   hjust = 0,
#   nudge_x = x_max_bridge * 0.25) +
ggrepel::geom_text_repel(
  data = bile_retinoid_df %>% filter(label_flag),
  aes(label = gene),
  size = 6,
  color = "grey20",
  segment.color = "grey35",
  segment.size = 0.4,
  direction = "y",
  hjust = 0,
  nudge_x = x_max_bridge * 0.025,
  box.padding = 0.2,
  point.padding = 0.15,
  max.overlaps = Inf,
  min.segment.length = 0
)+
  facet_wrap(~ pair_label, scales = "free_y", ncol = 1, labeller = labeller(
    pair_label = function(x) sub("\n.*$", "", x)
  ) ) +
scale_y_discrete(
  labels = function(x) sub("^.*___", "", x)
) +
  scale_x_continuous(
    limits = c(0, x_max_bridge * 1.08),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  scale_fill_grey(
    start = 0.75,
    end = 0.25,
    name = "Methods supporting\npositive selection"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.35),
    guide = "none"
  ) +
  # labs(
  #   x = "Median gene-level positive-site breadth across FUBAR and FEL\n(proportion of codon sites inferred as positively selected; descriptive ordering only)",
    # y = NULL,
    # title = "Bridge genes contributing to significant pathway-pair overlaps",
    # subtitle = paste0(
    #   "Only pathway pairs with BH-adjusted permutation significance are shown. ",
    #   "All displayed genes were supported in >= ", min_methods_support,
    #   " methods and annotated to both pathways in the pair. ",
    #   "Within each pair, genes are ordered by tool support first, then by median FUBAR/FEL breadth. ",
    #   "Top ", n_label_per_pair, " genes per pair are labeled, and any additional genes with median FUBAR/FEL breadth > ",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " are also labeled."
    # ),
    # caption = paste0(
    #   "Pair-level significance comes from the BH-corrected permutation test; the x-axis is a descriptive gene-ordering quantity, not a significance threshold. ",
    #   "The >",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " label rule is used only to highlight candidate genes for literature follow-up."
    # )
  theme_minimal(base_size = 11
  ) 

# p_bridge_Bile_retinoid
# 
# p_bridge_Bile_retinoid <- p_bridge_Bile_retinoid + theme_bold2
# 
# p_bridge_Bile_retinoid
# 
# ggsave("BILERetnoidsupple_lollipop.tiff",plot = p_bridge_Bile_retinoid, height = 5, width = 18)



p_bridge_Glycine <- ggplot(
  glycine_df,
  aes(x = median_breadth_ff, y = pair_gene)
) +
  geom_segment(
    aes(x = 0, xend = median_breadth_ff, yend = pair_gene),
    color = "grey88",
    linewidth = 1.6
  ) +
  geom_point(
    aes(fill = support_n, alpha = label_flag),
    shape = 21,
    size = 4,
    color = "black",
    stroke = 0.3
  ) +
geom_point(
  aes(fill = support_n, alpha = label_flag),
  shape = 21,
  size = 4,
  color = "black",
  stroke = 0.3
) +
# ggrepel::geom_text_repel(
#   data = bridge_fig_df %>% filter(!label_flag),
#   aes(
#     label = gene),
#   size = 4.0,
#   color = "black",
#   segment.color = "grey20",
#   segment.size = 0.3,
#   force = 0.25,                
#   max.overlaps = Inf,
#   direction = "y",
#   hjust = 0,
#   nudge_x = x_max_bridge * 0.25) +
ggrepel::geom_text_repel(
  data = glycine_df %>% filter(label_flag),
  aes(label = gene),
  size = 6,
  color = "grey20",
  segment.color = "grey35",
  segment.size = 0.4,
  direction = "y",
  hjust = 0,
  nudge_x = x_max_bridge * 0.025,
  box.padding = 0.2,
  point.padding = 0.15,
  max.overlaps = Inf,
  min.segment.length = 0
)+
  facet_wrap(~ pair_label, scales = "free_y", ncol = 1, labeller = labeller(
    pair_label = function(x) sub("\n.*$", "", x)
  )) +
scale_y_discrete(
  labels = function(x) sub("^.*___", "", x)
) +
  scale_x_continuous(
    limits = c(0, x_max_bridge * 1.08),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  scale_fill_grey(
    start = 0.75,
    end = 0.25,
    name = "Methods supporting\npositive selection"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.35),
    guide = "none"
  ) +
  labs(
    x = "Median gene-level positive-site breadth across FUBAR and FEL\n(proportion of codon sites inferred as positively selected; descriptive ordering only)",
    y = NULL,
    # title = "Bridge genes contributing to significant pathway-pair overlaps",
    # subtitle = paste0(
    #   "Only pathway pairs with BH-adjusted permutation significance are shown. ",
    #   "All displayed genes were supported in >= ", min_methods_support,
    #   " methods and annotated to both pathways in the pair. ",
    #   "Within each pair, genes are ordered by tool support first, then by median FUBAR/FEL breadth. ",
    #   "Top ", n_label_per_pair, " genes per pair are labeled, and any additional genes with median FUBAR/FEL breadth > ",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " are also labeled."
    # ),
    # caption = paste0(
    #   "Pair-level significance comes from the BH-corrected permutation test; the x-axis is a descriptive gene-ordering quantity, not a significance threshold. ",
    #   "The >",
    #   scales::percent(label_breadth_cut, accuracy = 1),
    #   " label rule is used only to highlight candidate genes for literature follow-up."
    # )
  ) +
  theme_minimal(base_size = 11
  ) 

p_bridge_Glycine

p_bridge_Glycine <- p_bridge_Glycine + theme_bold3

p_bridge_Glycine

ggsave("Glycinesupple_lollipop.tiff",plot = p_bridge_Glycine, height = 5, width = 18)



## =========================================================
##  suppl data 
## =========================================================

bridge_rank_qc <- significant_pair_driver_genes_ranked %>%
  mutate(
    high_breadth_candidate = median_breadth_ff > label_breadth_cut,
    top_flag = driver_rank_within_pair <= n_label_per_pair,
    label_flag = top_flag | high_breadth_candidate,
    label_reason = case_when(
      top_flag & high_breadth_candidate ~ "top_rank_and_high_breadth",
      top_flag ~ "top_rank",
      high_breadth_candidate ~ "high_breadth_only",
      TRUE ~ "unlabeled"
    )
  ) %>%
  select(
    path1, path2, gene,
    driver_rank_within_pair,
    support_n, support_methods,
    median_breadth_ff,
    high_breadth_candidate, top_flag, label_flag, label_reason,
    n_pathways, pathway_list,
    fubar_breadth, fel_breadth, meme_breadth,
    fubar_n_sig, fel_n_sig, meme_n_sig,
    p_bh
  ) %>%
  arrange(p_bh, path1, path2, driver_rank_within_pair)

bridge_rank_qc <- bridge_rank_qc %>%
  rename(
    category1 = path1,
    category2 = path2,
    n_functional_categories = n_pathways,
    functional_category_list = pathway_list
  )


bridge_rank_qc #%>% View()

write.csv(bridge_rank_qc, file = "bridge_rank_QC.csv")

```






## FIGURE 3E: absrel
genes farther right = supported on more distinct terminal branches
bigger points = support appears in more partitions
darker-filled points = higher median supported omega
colored outlines = which pathway dominates
triangles + grey side text = multi-pathway genes

....Which genes show the most recurrent terminal-branch (ones w/full annotation) aBSREL support, and are those genes concentrated in particular pathway contexts?

inset is answering a different question. It does not summarize genes. It uses all branch-level baseline omega values and compares their distributions across functional_category categories with one-vs-rest Wilcoxon tests, then BH-corrects those p-values. A star appears only when both raw p < 0.05 and BH-adjusted p <= 0.10

Main panel = recurrent gene-level accumulation of supported terminal-branch signal
Inset = pathway-level distributional context for branch-level baseline omega
```{r}

OMEGA_CAP <- 100

# -----------------------------
# user options
# -----------------------------
ABSREL_P_THRESH <- 0.05
ABSREL_Q_THRESH <- 0.10
MIN_BRANCHES_MAIN <- 2
MIN_PARTITIONS_MAIN <- 0
TOP_GENES_MAIN <- 25

# -----------------------------
# helpers
# -----------------------------
extract_species <- function(x) {
  str_match(x, "^[^_]+_([A-Z]+_[A-Z]+(?:_[A-Z]+)?)_ENS")[, 2] |>
    str_replace_all("_", " ")
}

# -----------------------------
# input
# -----------------------------
function_metadata <- read_csv("gene_sets_union.csv")

absrel_res <- read.csv("tabsrel_summary.csv") %>%
  tidyr::extract(
    col = gene_id,
    into = c("gene_id", "partition"),
    regex = "^(.*?)(part.*)$",
    remove = FALSE
  )

function_metadata <- function_metadata %>%
  rename(functional_category = pathway_label)

absrel <- absrel_res %>%
  left_join(function_metadata, by = c("gene_id" = "symbol")) %>%
  filter(!str_detect(branch_id, "^NODE")) %>%
  mutate(
    functional_category  = coalesce(functional_category, pathway_key, "unknown"),
    baseline_omega = pmin(aBSREL_baseline_omega, OMEGA_CAP),
    p_raw = pmin(
      coalesce(aBSREL_pvalue_uncorrected,
               aBSREL_pvalue_corrected,
               aBSREL_pvalue),
      1
    ),
    lrt            = aBSREL_LRT,    
    positive_omega_classes  = aBSREL_positive_omega_classes,  
    positive_omega_weights  = aBSREL_positive_omega_weights,
    gene          = gene_id,
    species_label = extract_species(branch_id),
    label_text    = paste(gene, species_label, sep = ", ")
  ) %>%
  filter(
    !is.na(branch_id),
    !is.na(gene),
    !is.na(species_label),
    !is.na(partition),
    !is.na(baseline_omega),
    !is.na(functional_category),
    !is.na(p_raw)
  ) %>%
  mutate(
    p_adj_dataset = p.adjust(p_raw, method = "BH"),
    supported_branch_global = p_raw < ABSREL_P_THRESH & p_adj_dataset <= ABSREL_Q_THRESH
  )

# -----------------------------
# partition ordering
# -----------------------------
partition_levels <- absrel %>%
  distinct(partition) %>%
  mutate(part_num = suppressWarnings(as.numeric(str_extract(partition, "\\d+")))) %>%
  arrange(part_num, partition) %>%
  pull(partition)

absrel <- absrel %>%
  filter(!str_detect(branch_id, regex("NODE", ignore_case = TRUE))) %>%  
  group_by(gene) %>%
  mutate(n_seqs_gene = n_distinct(branch_id)) %>%
  ungroup() %>%
  filter(n_seqs_gene >= 5) %>% 
  mutate(partition = factor(partition, levels = partition_levels))

absrel <- exorcise(absrel, "gene_id")

# =========================================================
#  globally supported terminal branches
# =========================================================
absrel_supported <- absrel %>%
  filter(supported_branch_global)

# supported branches with omega > 1 for positive-branch omega summaries
absrel_supported_pos <- absrel_supported %>%
  filter(baseline_omega > 1)

# =========================================================
#  summarize supported branches by gene
#    omega summaries use only globally supported branches with omega > 1
# =========================================================
absrel_gene_summary <- absrel_supported %>%
  group_by(gene) %>%
  summarise(
    n_supported_branch_obs = n(),
    n_unique_supported_terminal_branches = n_distinct(branch_id),

    n_partitions_supported = n_distinct(as.character(partition)),
    partitions = paste(sort(unique(as.character(partition))), collapse = "; "),

    n_species_observed = n_distinct(species_label),
    species = paste(sort(unique(species_label)), collapse = "; "),

    n_pathways_supported = n_distinct(functional_category),
    pathways = paste(sort(unique(functional_category)), collapse = " | "),
    dominant_pathway = names(sort(table(functional_category), decreasing = TRUE))[1],

    n_supported_terminal_branches_omega_gt1 = n_distinct(branch_id[baseline_omega > 1]),

    median_supported_omega_pos = {
      x <- baseline_omega[baseline_omega > 1]
      if (length(x) == 0) NA_real_ else median(x, na.rm = TRUE)
    },
    q90_supported_omega_pos = {
      x <- baseline_omega[baseline_omega > 1]
      if (length(x) == 0) NA_real_ else quantile(x, 0.90, na.rm = TRUE)
    },
    max_supported_omega_pos = {
      x <- baseline_omega[baseline_omega > 1]
      if (length(x) == 0) NA_real_ else max(x, na.rm = TRUE)
    },

    min_p_raw = min(p_raw, na.rm = TRUE),
    min_p_adj_dataset = min(p_adj_dataset, na.rm = TRUE),
    exemplar_species = species_label[which.max(baseline_omega)],
    exemplar_label = label_text[which.max(baseline_omega)],
    .groups = "drop"
  ) %>%
  mutate(
    n_supported_branches = n_unique_supported_terminal_branches
  ) %>%
  filter(
    n_unique_supported_terminal_branches >= MIN_BRANCHES_MAIN |
      n_partitions_supported >= MIN_PARTITIONS_MAIN |
      n_pathways_supported > 1
  ) %>%
  arrange(
    desc(n_unique_supported_terminal_branches),
    desc(n_partitions_supported),
    desc(n_supported_terminal_branches_omega_gt1),
    desc(median_supported_omega_pos),
    gene
  )

# keep only top recurrent genes for the main panel
absrel_top_genes <- absrel_gene_summary %>%
  slice_head(n = TOP_GENES_MAIN)



gene_levels <- absrel_top_genes %>%
  arrange(
    n_supported_branches,
    n_partitions_supported,
    median_supported_omega_pos,
    gene
  ) %>%
  pull(gene)

absrel_top_genes <- absrel_top_genes %>%
  mutate(
    gene = factor(gene, levels = gene_levels),
    dominant_pathway = factor(dominant_pathway),
    pathway_note = if_else(n_pathways_supported > 1, pathways, NA_character_)
  )

# =========================================================
#  main panel
# =========================================================
x_max_absrel <- max(absrel_top_genes$n_supported_branches, na.rm = TRUE)
if (!is.finite(x_max_absrel) || x_max_absrel <= 0) {
  x_max_absrel <- 1
}

#title and subtitle

    # title = "Recurrent aBSREL-supported signal across top genes",
    # subtitle = paste0(
    #   "Main panel summarizes genes with globally supported terminal branches ",
    #   "(raw p < ", ABSREL_P_THRESH,
    #   ", dataset-wide BH-adjusted q <= ", ABSREL_Q_THRESH,
    #   "). X = number of unique supported terminal branches; point size = number of supported partitions. ",
    #   "Point fill shows the median baseline omega among supported branches with omega > 1. ",
    #   "Small grey text marks pathway categories for multi-pathway genes."


pABSREL_main <- ggplot(
  absrel_top_genes,
  aes(x = n_supported_branches, y = gene)
) +
  geom_segment(
    aes(
      x = 0,
      xend = n_supported_branches,
      yend = gene,
      color = dominant_pathway
    ),
    linewidth = 0.8,
    alpha = 0.55,
    show.legend = FALSE
  ) +
  geom_point(
    aes(
      size = n_partitions_supported,
      fill = median_supported_omega_pos,
      color = dominant_pathway,
      shape = n_pathways_supported > 1
    ),
    stroke = 0.5,
    alpha = 0.95
  ) +
      scale_color_manual(values = Category)+
  geom_text_repel(
    aes(label = as.character(gene)),
    size = 0,
    box.padding = 0.22,
    point.padding = 0.14,
    min.segment.length = 0,
    max.overlaps = Inf,
    seed = 1,
    color = "black",
    show.legend = FALSE
  ) +
  geom_text(
    data = absrel_top_genes %>% filter(n_pathways_supported > 1, !is.na(pathway_note)),
    aes(
      x = n_supported_branches + max(0.18, 0.03 * x_max_absrel),
      y = gene,
      label = pathway_note
    ),
    inherit.aes = FALSE,
    vjust = 0.35,
    hjust = -0.03,
    size = 2.85,
    color = "grey15",
    lineheight = 0.95
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 21, `TRUE` = 24),
    labels = c(`FALSE` = "single-Category gene", `TRUE` = "multi-Category gene"),
    name = "Category context"
  ) +
  scale_size_continuous(
    range = c(2.5, 6.5),
    breaks = sort(unique(absrel_top_genes$n_partitions_supported)),
    name = "Supported partitions"
  ) +
  scale_fill_gradient(
    low = "grey85",
    high = "grey15",
    na.value = "grey92",
    name = "Median baseline omega\namong supported branches\nwith omega > 1"
  ) +
  scale_x_continuous(
    breaks = pretty_breaks(),
    limits = c(0, x_max_absrel * 1.55),
    expand = expansion(mult = c(0.01, 0.04))
  ) +
  labs(
    # title = "Recurrent aBSREL-supported signal across top genes",
    # subtitle = paste0(
    #   "Main panel summarizes genes with globally supported terminal branches ",
    #   "(raw p < ", ABSREL_P_THRESH,
    #   ", dataset-wide BH-adjusted q <= ", ABSREL_Q_THRESH,
    #   "). X = number of unique supported terminal branches; point size = number of supported partitions. ",
    #   "Point fill shows the median baseline omega among supported branches with omega > 1. ",
    #   "Small grey text marks pathway categories for multi-pathway genes."
    # ),
    x = "Unique supported terminal branches",
    y = NULL,
    color = "Dominant Category"
  ) +
  guides(color = "none")+
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 8, color = "grey20"),
    axis.text.x = element_text(color = "grey15"),
    axis.title = element_text(color = "grey10"),
    plot.title = element_text(face = "bold", size = 13, color = "grey10"),
    plot.subtitle = element_text(size = 10, color = "grey35"),
    legend.position = "right"
  ) 
pABSREL_main <- pABSREL_main + theme_bold

# =========================================================
# branch-level companion table for displayed genes
# =========================================================
absrel_branch_table <- absrel_supported %>%
  semi_join(absrel_top_genes %>% select(gene), by = "gene") %>%
  left_join(
    absrel_top_genes %>%
      select(
        gene,
        n_supported_branch_obs,
        n_unique_supported_terminal_branches,
        n_supported_branches,
        n_partitions_supported,
        n_pathways_supported,
        dominant_pathway,
        pathways,
        n_supported_terminal_branches_omega_gt1,
        median_supported_omega_pos,
        q90_supported_omega_pos,
        max_supported_omega_pos
      ),
    by = "gene"
  ) %>%
  arrange(
    desc(n_unique_supported_terminal_branches),
    desc(n_partitions_supported),
    gene,
    desc(baseline_omega),
    p_raw,
    species_label
  ) %>%
  select(
    gene,
    species_label,
    functional_category,
    partition,
    baseline_omega,
    p_raw,
    p_adj_dataset,
    supported_branch_global,
    n_supported_branch_obs,
    n_unique_supported_terminal_branches,
    n_supported_branches,
    n_partitions_supported,
    n_pathways_supported,
    dominant_pathway,
    pathways,
    n_supported_terminal_branches_omega_gt1,
    median_supported_omega_pos,
    q90_supported_omega_pos,
    max_supported_omega_pos
  )

absrel_branch_table

# compact gene-level table for writeup
absrel_gene_table <- absrel_top_genes %>%
  select(
    gene,
    dominant_pathway,
    n_supported_branch_obs,
    n_unique_supported_terminal_branches,
    n_supported_branches,
    n_partitions_supported,
    n_pathways_supported,
    pathways,
    n_supported_terminal_branches_omega_gt1,
    median_supported_omega_pos,
    q90_supported_omega_pos,
    max_supported_omega_pos,
    min_p_raw,
    min_p_adj_dataset,
    exemplar_species
  )

absrel_gene_table

# =========================================================
# pathway inset
# revert to all branch-level baseline omega values by pathway
# sideways, consistent with MEME / FEL / FUBAR
# =========================================================
path_order <- absrel %>%
  group_by(functional_category) %>%
  summarise(med = median(baseline_omega, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(functional_category)

pathway_plot_df <- absrel %>%
  mutate(
    functional_category = factor(functional_category, levels = path_order)
  )

sig_tbl <- tibble(functional_category = levels(pathway_plot_df$functional_category)) %>%
  mutate(
    p_value = purrr::map_dbl(functional_category, \(pw) {
      x <- pathway_plot_df$baseline_omega[pathway_plot_df$functional_category == pw]
      y <- pathway_plot_df$baseline_omega[pathway_plot_df$functional_category != pw]
      if (length(x) < 2 || length(y) < 2) NA_real_
      else wilcox.test(x, y, exact = FALSE)$p.value
    }),
    median_pathway = purrr::map_dbl(functional_category, \(pw)
      median(pathway_plot_df$baseline_omega[pathway_plot_df$functional_category == pw], na.rm = TRUE)
    ),
    median_rest = purrr::map_dbl(functional_category, \(pw)
      median(pathway_plot_df$baseline_omega[pathway_plot_df$functional_category != pw], na.rm = TRUE)
    )
  ) %>%
  mutate(
    delta_median = median_pathway - median_rest,
    p_adj = p.adjust(p_value, method = "BH"),
    sig_both = !is.na(p_value) & !is.na(p_adj) & p_value < 0.05 & p_adj <= 0.10
  )

star_df_absrel <- pathway_plot_df %>%
  group_by(functional_category) %>%
  summarise(
    y_max = max(baseline_omega, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y = pmax(y_max * 1.35, y_max + 0.05)
  ) %>%
  left_join(
    sig_tbl %>% select(functional_category, sig_both),
    by = "functional_category"
  ) %>%
  filter(sig_both)

upper_lim_absrel <- max(
  c(pathway_plot_df$baseline_omega, star_df_absrel$y, 0.05),
  na.rm = TRUE
) * 1.05



pathway_plot_df$functional_category <- factor(
  pathway_plot_df$functional_category,
  levels = Category_order
)

star_df_absrel$functional_category <- factor(
  star_df_absrel$functional_category,
  levels = Category_order
)

#title and subtitle 
    # title = "Distributions of branch-level baseline omega by pathway",
    # subtitle = "* shown only when raw p < 0.05 and BH-adjusted p <= 0.10 in one-vs-rest Wilcoxon tests",


p_inset_absrel <- ggplot(
  pathway_plot_df,
  aes(functional_category, baseline_omega, color = functional_category)
) +
  geom_violin(
    fill = "grey90",
    color = "grey45",
    scale = "width",
    alpha = 0.55,
    linewidth = 0.35
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    fill = NA,
    color = "grey20",
    linewidth = 0.30
  ) +
  geom_jitter(width = 0.12, size = 0.9, alpha = 0.75, stroke = 0.2,
              show.legend = FALSE) +
  geom_label(
    data = star_df_absrel,
    aes(functional_category, y),
    label = "*",
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    label.size = 0,
    size = 10.0,
    label.padding = grid::unit(0.05, "lines")
  ) +
  coord_flip(clip = "off") +
  scale_color_manual(values = Category) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 0.01),
    limits = c(0, upper_lim_absrel),
    expand = expansion(mult = c(0.02, 0.20)),
    name = "Baseline omega across all terminal branches"
  ) +
  labs(
    x = NULL,
    # title = "Distributions of branch-level baseline omega by Category",
    # subtitle = "* shown only when raw p < 0.05 and BH-adjusted p <= 0.10 in one-vs-rest Wilcoxon tests",
    color = "Category"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 9, face = "bold"),
    plot.subtitle = element_text(size = 7),
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 6),
    legend.position = "none",
    plot.margin = margin(4, 18, 4, 4),
    plot.background = element_rect(fill = "white", color = "white", linewidth = 0.4),  
    panel.background = element_rect(fill = "white", color = NA),
  )+inset_style()

p_inset_absrel

# =========================================================
#  final composed figure
# =========================================================
pABSREL_final <- pABSREL_main +
  patchwork::inset_element(
    p_inset_absrel,
    left = 0.60,
    right = 0.98,
    bottom = 0.42,
    top = 0.72,
    align_to = "panel"
  )

pABSREL_final

absrel_supp_main <- absrel_supported %>%
  semi_join(absrel_top_genes %>% select(gene), by = "gene") %>%
  transmute(
    gene,
    species_label,
    functional_category,
    partition,
    baseline_omega          = round(baseline_omega, 4),
    lrt                     = round(lrt, 4),
    p_raw                   = signif(p_raw, 4),
    p_adj_dataset           = signif(p_adj_dataset, 4),
    supported_branch_global,
    positive_omega_classes,
    positive_omega_weights
  ) %>%
  arrange(gene, desc(baseline_omega), p_raw, species_label)

absrel_supp_inset <- sig_tbl %>%
  transmute(
    functional_category,
    n_path = purrr::map_int(functional_category, ~ sum(pathway_plot_df$functional_category == .x)),
    n_rest = purrr::map_int(functional_category, ~ sum(pathway_plot_df$functional_category != .x)),
    med_path = round(median_pathway, 4),
    med_rest = round(median_rest, 4),
    delta_median = round(delta_median, 4),
    p_raw = signif(p_value, 4),
    p_bh = signif(p_adj, 4),
    sig_both
  ) %>%
  arrange(p_bh, p_raw)


ggsave("Fig3ABSREL.E.tiff",plot = pABSREL_main, height = 8, width = 15)


ggsave("Fig3ABSRELinset.E.tiff",plot = p_inset_absrel, height = 8, width = 15)


absrel_top_genes <- absrel_top_genes %>%
  rename(
    n_functional_categories_supported = n_pathways_supported,
    functional_categories = pathways,
    dominant_functional_category = dominant_pathway,
    functional_category_note = pathway_note
  )

# write.csv(breadth_test, file="breadth_test_ABSREL.csv")
# write.csv(intensity_test, file = "intensity_test_ABSREL.csv")
write.csv(absrel_top_genes, file = "absrel_top_genes.csv" )
write.csv(absrel_supp_inset, file = "absrel_supp_inset.csv")
write.csv(absrel_supp_main, file = "absrel_supp_main.csv" )


```

#SupplTable
```{r}


# 
# 
# absrel_gene_summary_full <- absrel_supported %>%
#   group_by(gene) %>%
#   summarise(
#     n_supported_branch_obs = n(),
#     n_unique_supported_terminal_branches = n_distinct(branch_id),
# 
#     n_partitions_supported = n_distinct(as.character(partition)),
#     partitions = paste(sort(unique(as.character(partition))), collapse = "; "),
# 
#     n_species_observed = n_distinct(species_label),
#     species = paste(sort(unique(species_label)), collapse = "; "),
# 
#     n_categories_supported = n_distinct(functional_category),
#     categories = paste(sort(unique(functional_category)), collapse = " | "),
#     dominant_category = names(sort(table(functional_category), decreasing = TRUE))[1],
# 
#     n_supported_terminal_branches_omega_gt1 =
#       n_distinct(branch_id[baseline_omega > 1]),
# 
#     median_supported_omega_pos = {
#       x <- baseline_omega[baseline_omega > 1]
#       if (length(x) == 0) NA_real_ else median(x, na.rm = TRUE)
#     },
#     q90_supported_omega_pos = {
#       x <- baseline_omega[baseline_omega > 1]
#       if (length(x) == 0) NA_real_ else quantile(x, 0.90, na.rm = TRUE)
#     },
#     max_supported_omega_pos = {
#       x <- baseline_omega[baseline_omega > 1]
#       if (length(x) == 0) NA_real_ else max(x, na.rm = TRUE)
#     },
# 
#     min_p_raw = min(p_raw, na.rm = TRUE),
#     min_p_adj_dataset = min(p_adj_dataset, na.rm = TRUE),
#     exemplar_species = species_label[which.max(baseline_omega)],
#     exemplar_label = label_text[which.max(baseline_omega)],
#     .groups = "drop"
#   ) %>%
#   mutate(
#     n_supported_branches = n_unique_supported_terminal_branches,
#     category_note = if_else(n_categories_supported > 1, categories, NA_character_)
#   ) %>%
#   arrange(
#     gene,
#     desc(n_unique_supported_terminal_branches),
#     desc(n_partitions_supported)
#   )
# 
# # absrel_gene_summary_full <- absrel_gene_summary_full %>%
# #   rename(
# #     dominant_category = dominant_pathway,
# #     categories = pathways,
# #     n_categories_supported = n_pathways_supported
# #   )
# 
# absrel_full_primategene_table <- absrel_supported %>%
#   select(
#     species_label,
#     gene,
#     functional_category,
#     partition,
#     baseline_omega,
#     p_raw,
#     p_adj_dataset,
#     supported_branch_global
#   ) %>%
#   left_join(
#     absrel_gene_summary_full %>%
#       select(
#         gene,
#         n_supported_branch_obs,
#         n_unique_supported_terminal_branches,
#         n_supported_branches,
#         n_partitions_supported,
#         n_categories_supported,
#         dominant_category,
#         categories,
#         n_supported_terminal_branches_omega_gt1,
#         median_supported_omega_pos,
#         q90_supported_omega_pos,
#         max_supported_omega_pos,
#         min_p_raw,
#         min_p_adj_dataset,
#         exemplar_species
#       ),
#     by = "gene"
#   ) %>%
#   arrange(
#     gene,
#     species_label,
#     desc(baseline_omega),
#     p_raw
#   )
# 
# 
# absrel_full_primategene_table <- absrel_full_primategene_table %>%
#   rename(
#  functional_categories =  categories
#   )
# 
# 
# absrel_full_primategene_table
# 
# write.csv(absrel_full_primategene_table, file = "absrel_full_primategene_table.csv")

bile_fat_genes <- unique(bile_fatty_df$gene)


bile_fat_branches <- absrel_supported %>%
  filter(
    gene %in% bile_fat_genes,
    baseline_omega > 1,
    supported_branch_global == TRUE
  ) %>%
  count(species_label, sort = TRUE)



library(purrr)

species_enrichment <- map_df(unique(absrel_supported$species_label), function(sp) {
  df <- absrel_supported %>%
    filter(species_label == sp) %>%
    distinct(gene, functional_category, hit = (baseline_omega > 1 & supported_branch_global))

  k <- sum(df$hit & df$functional_category %in% c("Bile acid biosynthesis & metabolism",
                                                  "Fatty acid & cholesterol metabolism"))
  K <- sum(df$functional_category %in% c("Bile acid biosynthesis & metabolism",
                                         "Fatty acid & cholesterol metabolism"))
  n <- sum(df$hit)
  N <- nrow(df)

  tibble(
    species = sp,
    p_hyper = phyper(k - 1, K, N - K, n, lower.tail = FALSE),
    k = k, K = K, n = n, N = N
  )
})

set.seed(123)

all_genes <- unique(absrel_supported$gene)

# observed recurrence
obs <- absrel_supported %>%
  filter(
    gene %in% bile_fat_genes,
    baseline_omega > 1,
    supported_branch_global
  ) %>%
  count(species_label, name = "obs_hits")

# permutation
perm_results <- replicate(5000, {
  perm_genes <- sample(all_genes, length(bile_fat_genes))
  absrel_supported %>%
    filter(
      gene %in% perm_genes,
      baseline_omega > 1,
      supported_branch_global
    ) %>%
    count(species_label) %>%
    pull(n)
}, simplify = FALSE)

perm_df <- tibble(
  species_label = obs$species_label,
  obs_hits = obs$obs_hits,
  p_perm = map_dbl(obs$obs_hits, function(x) {
    mean(unlist(perm_results) >= x)
  })
)


gene_branch_counts <- absrel_supported %>%
  filter(baseline_omega > 1, supported_branch_global) %>%
  count(gene, name = "n_supported_branches")

bile_fat_genes <- unique(bile_fatty_df$gene)

bile_fat_branch_counts <- gene_branch_counts %>%
  filter(gene %in% bile_fat_genes)

k <- sum(gene_branch_counts$gene %in% bile_fat_genes & gene_branch_counts$n_supported_branches > 0)
K <- sum(gene_branch_counts$gene %in% bile_fat_genes)
n <- sum(gene_branch_counts$n_supported_branches > 0)
N <- nrow(gene_branch_counts)

p_hyper_gene <- phyper(k - 1, K, N - K, n, lower.tail = FALSE)
p_hyper_gene

obs_total_branches <- sum(bile_fat_branch_counts$n_supported_branches)

set.seed(123)

all_genes <- unique(gene_branch_counts$gene)
n_bile <- length(bile_fat_genes)

perm_totals <- replicate(5000, {
  perm_genes <- sample(all_genes, n_bile)
  sum(gene_branch_counts$n_supported_branches[gene_branch_counts$gene %in% perm_genes])
})

p_perm_gene <- mean(perm_totals >= obs_total_branches)
p_perm_gene

gene_species_matrix <- absrel_supported %>%
  mutate(hit = as.integer(baseline_omega > 1 & supported_branch_global)) %>%
  group_by(gene, species_label) %>%
  summarise(hit = max(hit), .groups = "drop") %>%   # collapse duplicates
  pivot_wider(
    names_from = species_label,
    values_from = hit,
    values_fill = 0
  )


library(vegan)

bile_fat_matrix <- gene_species_matrix %>%
  filter(gene %in% bile_fat_genes) %>%
  column_to_rownames("gene")

jaccard_matrix <- vegdist(bile_fat_matrix, method = "jaccard")

jaccard_matrix

row_sums <- rowSums(bile_fat_matrix)
row_sums

bile_fat_matrix_filtered <- bile_fat_matrix[rowSums(bile_fat_matrix) > 0, ]

jaccard_matrix1 <- vegdist(bile_fat_matrix_filtered, method = "jaccard")
 

hc <- hclust(jaccard_matrix1, method = "average")
plot(hc)

tiff("bilefat_jaccard_dendrogram.tiff", width = 2000, height = 1600, res = 300, compression = "lzw")
plot(hc, main = "Bile-Fat Gene Convergence Dendrogram", xlab = "", sub = "")
dev.off()


library(pheatmap)

Heatmap_Bile <- pheatmap(as.matrix(bile_fat_matrix),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = c("white", "black"))

Heatmap_Bile

write.csv(
  bile_fat_branches,
  file = "bile_fat_branch_recurrence.csv",
  row.names = FALSE
)


write.csv(
  species_enrichment,
  file = "species_enrichment_bilefat_genedrivers.csv",
  row.names = FALSE
)


write.csv(
  perm_df,
  file = "species_permutation_bilefat_genedrivers.csv",
  row.names = FALSE
)


write.csv(
  gene_branch_counts,
  file = "gene_branch_counts_allgenes.csv",
  row.names = FALSE
)

write.csv(
  bile_fat_branch_counts,
  file = "gene_branch_counts_bilefat_only.csv",
  row.names = FALSE
)


write.table(
  data.frame(p_hyper_gene = p_hyper_gene),
  file = "genelevel_hypergeometric_bilefat.txt",
  row.names = FALSE,
  quote = FALSE
)


write.table(
  data.frame(
    obs_total_branches = obs_total_branches,
    p_perm_gene = p_perm_gene
  ),
  file = "genelevel_permutation_bilefat.txt",
  row.names = FALSE,
  quote = FALSE
)


write.csv(
  gene_species_matrix,
  file = "gene_species_hit_matrix.csv",
  row.names = FALSE
)


write.csv(
  bile_fat_matrix,
  file = "bilefat_gene_species_matrix_raw.csv"
)

write.csv(
  bile_fat_matrix_filtered,
  file = "bilefat_gene_species_matrix_filtered.csv"
)


write.csv(
  as.matrix(jaccard_matrix),
  file = "jaccard_matrix_raw.csv"
)


write.csv(
  as.matrix(jaccard_matrix1),
  file = "jaccard_matrix_filtered.csv"
)


write.csv(
  as.matrix(bile_fat_matrix),
  file = "bilefat_heatmap_matrix.csv"
)

library(grid)

png("Heatmap_Bilefat.tiff", width = 2000, height = 2400, res = 300)
grid::grid.newpage()
grid::grid.draw(Heatmap_Bile$gtable)
dev.off()



```

