---
title: "3One2Many"
author: "Will Quinn"
date: "2026-05-19"
output: html_document
---

## Load Packages 
```{r }

#https://hyphy.org/resources/json-fields.pdf
library(ape)
library(ggtree)
library(tidyverse)
library(tidytext)
library(dplyr)
library(purrr)
library(stringr)
library(scales)
library(glue)
library(tidytext)
library(ggtext)
library(KEGGREST)
#BiocManager::install("sangerseqR")
#devtools::install_version("bold", version="0.9.0", dependencies = FALSE)
#install.packages('datelife', repos = c('https://phylotastic.r-universe.dev', 'https://cloud.r-project.org'))
library(datelife)
library(cowplot)
library(tidyr)
library(ggplot2)
library(viridis)
library(ggrepel)
library(patchwork)
library(forcats)
library(scales)
library(tidyverse)
#install.packages("eulerr")
library(eulerr) 
#install.packages("UpSetR")
library(UpSetR)    
library(phylocomr)
# install.packages("rotl")
# install.packages("ape") 
# install.packages("devtools")
# install.packages("stringr")
# install.packages("datelife")
library(rotl)
library(ape)
library(devtools)
library(rlang)
library(ggpubr)
#install.packages("ggridges")
library(ggridges)
library(gridExtra)
library(gridExtra)
library(grid)


```







#Ortholog detect
```{r}
library(httr)
library(jsonlite)
library(dplyr)
library(readr)
library(furrr)
# install.packages("RMariaDB")
# install.packages("DBI")
#install.packages("RMySQL")
library(RMySQL)
library(DBI)
library(RMariaDB)

output_dir <- "OrthologTester"
dir.create(output_dir, showWarnings = FALSE)

#########ONE2ONE ORTHOLOG FILTER #######
manifest <- read_csv("manifest.csv")
genes <- unique(na.omit(manifest$gene))
con <- dbConnect(MySQL(), host = "ensembldb.ensembl.org", port = 3306,
                 user = "anonymous", dbname = "ensembl_compara_113")
species_set <- c("pan_troglodytes","pan_paniscus","gorilla_gorilla","pongo_abelii",
  "nomascus_leucogenys","chlorocebus_sabaeus","macaca_mulatta","macaca_fascicularis",
  "macaca_nemestrina","papio_anubis","callithrix_jacchus","saimiri_boliviensis_boliviensis",
  "aotus_nancymaae","microcebus_murinus","colobus_angolensis_palliatus","cercocebus_atys",
  "theropithecus_gelada","prolemur_simus","carlito_syrichta","propithecus_coquereli")
q <- sprintf("SELECT gm1.display_label AS gene, gdb2.name AS species, h.description AS homology_type
  FROM homology h JOIN homology_member hm1 ON h.homology_id=hm1.homology_id
  JOIN homology_member hm2 ON h.homology_id=hm2.homology_id AND hm1.gene_member_id<>hm2.gene_member_id
  JOIN gene_member gm1 ON hm1.gene_member_id=gm1.gene_member_id
  JOIN gene_member gm2 ON hm2.gene_member_id=gm2.gene_member_id
  JOIN genome_db gdb1 ON gm1.genome_db_id=gdb1.genome_db_id
  JOIN genome_db gdb2 ON gm2.genome_db_id=gdb2.genome_db_id
  WHERE gdb1.name='homo_sapiens' AND gdb2.name IN ('%s') AND gm1.display_label IN ('%s')
  AND h.description='ortholog_one2one'",
  paste(species_set, collapse="','"), paste(genes, collapse="','"))
tags <- dbGetQuery(con, q) %>% distinct()
dbDisconnect(con)

MIN_ONE2ONE_SPECIES <- 15

tagged <- manifest %>% left_join(tags, by = c("gene", "species")) %>%
  filter(species != "homo_sapiens" & status == "alternative")
tagged2 <- manifest %>% left_join(tags, by = c("gene", "species")) %>%
  filter(species != "homo_sapiens" & status == "canonical")
exclusively_one2one_genes <- tagged %>%
  group_by(gene) %>%
  filter( n_distinct(species[homology_type == "ortholog_one2one" & !is.na(homology_type)]) >= MIN_ONE2ONE_SPECIES ) %>%
  ungroup() %>%
  pull(gene) %>%
  unique() %>%
  sort()
exclusively_one2one_genesV2 <- tagged2 %>%
  group_by(gene) %>%
  filter( n_distinct(species[homology_type == "ortholog_one2one" & !is.na(homology_type)]) >= MIN_ONE2ONE_SPECIES ) %>%
  ungroup() %>%
  pull(gene) %>%
  unique() %>%
  sort()
one2one_filter <- c(exclusively_one2one_genesV2,exclusively_one2one_genes) %>% sort() %>% unique()
##############################################
  
length(one2one_filter)

filter_one2manyID <- function(df, gene_col) {
  df %>%
    mutate(.gene_tmp = toupper(.data[[gene_col]])) %>%
    filter(.gene_tmp %in% one2one_filter) %>%
    select(-.gene_tmp)
}

 

demons <- c(
  "ABAT", "ACOT2", "ADH1A", "ADH1B", "ADH1C", "ADH6", "APOA2", "ATP1A4", "CETP",
  "CH25H", "CTH", "CYP2A6", "CYP2A7", "CYP2C8", "CYP2C9", "CYP3A4", "CYP3A7",
  "DHFR2", "DHRS4L2", "GPAT2", "LDHA", "LDHAL6B", "LPA", "PGAM4", "PLA2G2C",
  "PLA2G4C", "PRKACG", "SLCO1B1", "UGT1A10", "UGT1A3", "UGT1A5", "UGT1A6",
  "UGT1A7", "UGT1A8", "UGT1A9", "UGT2A2", "UGT2B10", "UGT2B11", "UGT2B15",
  "UGT2B28", "UGT2B4", "UGT2B7"
)

exorcise <- function(df, gene_col) {
  df %>%
    mutate(.gene_tmp = toupper(.data[[gene_col]])) %>%
    filter(!.gene_tmp %in% demons) %>%
    select(-.gene_tmp)
}
 

```


## Setup

any reference to pathway is a reference to functional category

thinking we want to do two panels here one with add'l primate "lineage" category to see if we can do some clustering that way or something?

then if nothing else: panel A primate diet category view, 
                      panel B primate EQ category view

```{r setup, include=FALSE}

absrel_res <- read.csv("tabsrel_summary.csv")
absrel_res <- absrel_res %>% extract( col = gene_id, into = c("gene_id", "partition"), regex = "^(.*)(part.*)$" )
absrel_res <- exorcise(absrel_res, "gene_id")
absrel_res <- filter_one2manyID(absrel_res, "gene_id")
function_metadata <- read_csv("gene_sets_union.csv")
function_metadata <- exorcise(function_metadata, "symbol")
function_metadata <- filter_one2manyID(function_metadata, "symbol")
absrel_res <- merge(absrel_res, function_metadata, by.x = "gene_id", by.y = "symbol")
#### remove internal/root branches... for clarity/interpretations..
absrel_res <- absrel_res %>% filter(!str_detect(branch_id, "NODE"))
#see python aux script for which we used to handle gene length consideration...
#/projects/b1057/apulvino/Chapter2/OrthologExplorer/results_sequences... rest of work for the entire proj is in v4 subdir...
#used after so we have clean merge with fel_res df: sed -i 's/|/_/g' gene_lengths.tsv
gene_length <- read_tsv("gene_lengths.tsv") %>% mutate(gene_id = str_to_upper(gene_id) )
gene_length <- filter_one2manyID(gene_length, "gene_dir")
gene_length <- exorcise(gene_length, "gene_dir")
absrel_res <- merge(absrel_res, gene_length, by.x = "branch_id" , by.y = "gene_id") %>% unique()
absrel_res$species <- sub(".*_([A-Za-z]+_[A-Za-z]+)_.*", "\\1", absrel_res$branch_id)

absrel_res <- filter_one2manyID(absrel_res, "gene_id")

absrel_res <- exorcise(absrel_res, "gene_id")

# ---- user config ----
PVAL_THRESH <- 0.05
OMEGA_CAP <- 100             # cap huge omega values for plotting
# ----------------------

# make p-value column robust (prefer corrected if available)
absrel <- absrel_res %>%
  mutate(
    p_raw = coalesce(aBSREL_pvalue_uncorrected, aBSREL_pvalue),
    p_corr = coalesce(aBSREL_pvalue_corrected, aBSREL_pvalue, p_raw),
    p_use = if_else(!is.na(p_corr), p_corr, p_raw),
    is_sig = if_else(!is.na(p_use) & p_use <= PVAL_THRESH, TRUE, FALSE),
    branch_omega = coalesce(aBSREL_branch_omega, aBSREL_max_omega, NA_real_),
    omega_plot = pmin(replace_na(branch_omega, 0), OMEGA_CAP),  # capped for plotting
    lrt = coalesce(aBSREL_LRT, 0),
    gene = gene_id,  # unify name
    pathway_label = coalesce(pathway_label, pathway_key, "unknown")
  ) %>%
  dplyr::select(branch_id, gene, pathway_label, species, everything())

############# CONSOLIDATION; DEFINING SIGNIFICANT FEATURES ######################
absrel_clean <- absrel %>%
  filter(!str_detect(branch_id, regex("NODE", ignore_case = TRUE))) %>%
  # group_by(gene) %>%
  # filter(n() >= 5) %>%
  # ungroup() %>%
  mutate(
    p_raw       = coalesce(aBSREL_pvalue_uncorrected, aBSREL_pvalue),
    p_corr      = coalesce(aBSREL_pvalue_corrected, aBSREL_pvalue, p_raw),
    p_use       = if_else(!is.na(p_corr), p_corr, p_raw),
    q_global    = p.adjust(p_use, method = "BH"),
    is_sig_glob = !is.na(q_global) & q_global <= 0.1,
    is_sig_raw  = !is.na(p_use) & p_use <= PVAL_THRESH,
    branch_omega = coalesce(aBSREL_branch_omega, aBSREL_max_omega, NA_real_),
    omega_plot   = pmin(replace_na(branch_omega, 0), OMEGA_CAP)
    )

```


## Branch Explorer
panel A asks which group contributes the most total positive-branch burden ("burden" question) in a pathway. Using summed  branch omega

Panel B asks which group accounts for the largest share of positive branches for each gene (representation question),... more fractional share of positive branches with an tie-breaking whole number of branches winner if needed (e.g. highest raw count of pos branches wins)
### FIGURE 4A: primate species comparisons

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(scales)
library(grid)



pathway_order <- c(
  "Taurine and cysteine metabolism",
  "Retinoid / vitamin A metabolism",
  "Glycine metabolism",
  "Fatty acid & cholesterol metabolism",
  "Bile acid biosynthesis & metabolism"
)

pathway_orderb <- c(
  "Bile acid biosynthesis & metabolism",
  "Fatty acid & cholesterol metabolism",
  "Glycine metabolism",
  "Retinoid / vitamin A metabolism",
  "Taurine and cysteine metabolism"
)


species_palette <- c(
  "AOTUS_NANCYMAAE"                 = "#0072B2",
  "COLOBUS_ANGOLENSIS_PALLIATUS"    = "brown",
  "PAN_TROGLODYTES"                 = "#009E73",
  "MACACA_FASCICULARIS"             = "purple",
  "CALLITHRIX_JACCHUS"              = "black",
  "SAIMIRI_BOLIVIENSIS_BOLIVIENSIS" = "#56B4E9",
  "HOMO_SAPIENS"                    = "red",
  "MICROCEBUS_MURINUS"              = "darkgrey",
  "PONGO_ABELII"                    = "#CC79A7",
  "PAPIO_ANUBIS"                    = "#0072B2",
  "GORILLA_GORILLA"                 = "#D55E00",
  "PROLEMUR_SIMUS"                  = "#F0E442",
  "MACACA_NEMESTRINA"               = "#56B4E9",
  "CARLITO_SYRICHTA"                = "#999999",
  "MACACA_MULATTA"                  = "#E69F00",
  "CHLOROCEBUS_SABAEUS"             = "forestgreen",
  "CERCOCEBUS_ATYS"                 = "#CC79A7"
)

theme_bold_large <- theme(
  plot.title = element_text(size = 13, face = "bold", hjust = 0.1),
  plot.subtitle = element_text(size = 14, face = "bold"),

  axis.title.x = element_text(size = 12, face = "bold"),
  axis.title.y = element_text(size = 12, face = "bold"),
  axis.text.x  = element_text(size = 12, face = "bold", color = "black"),
  axis.text.y  = element_text(size = 9.75, face = "bold", color = "black"),

  legend.title = element_text(size = 12, face = "bold"),
  legend.text  = element_text(size = 10, face = "bold"),
  legend.key   = element_blank(),

  strip.text.y = element_text(size = 8, face = "bold", angle = 0),
  strip.text.y.left = element_text(margin = margin(1,1,1,1)),

  panel.grid = element_blank(),
  panel.border = element_rect(color = "black", size = .5, fill = NA),
  axis.line = element_blank(),,

  plot.margin = margin(10, 18, 10, 10)
)


# theme_bold_large <- theme(
#   # Titles
#   plot.title = element_text(size = 14, face = "bold", hjust = 0.1),
#   plot.subtitle = element_text(size = 14, face = "bold"),
# 
#   # Axes
#   axis.title.x = element_text(size = 13, face = "bold"),
#   axis.title.y = element_text(size = 13, face = "bold"),
#   axis.text.x  = element_text(size = 10, face = "bold", color = "black"),
#   axis.text.y  = element_text(size = 10, face = "bold", color = "black"),
# 
#   # Legend
#   legend.title = element_text(size = 14, face = "bold"),
#   legend.text  = element_text(size = 12, face = "bold"),
#   legend.key   = element_blank(),
# 
#   # Facet strips (if present)
#   strip.text.y = element_text(size = 6, face = "bold", angle =90),
#     strip.text.y.left = element_text(margin = margin(1,1,1,1)),
# 
#   # Panel + background
#   panel.grid = element_blank(),
#   panel.border = element_rect(color = "black", size = 1.1),
#   axis.line = element_line(color = "black", linewidth = 1.2),
# 
#   # Margins (keeps q-values visible)
#   plot.margin = margin(10, 140, 1, 10),
# )

theme_bold_large1 <- theme(
  # Titles
  plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 14, face = "bold"),

  # Axes
  axis.title.x = element_text(size = 16, face = "bold"),
  axis.title.y = element_text(size = 16, face = "bold"),
  axis.text.x  = element_text(size = 13, face = "bold", color = "black"),
  axis.text.y  = element_text(size = 13, face = "bold", color = "black"),

  # Legend
  legend.title = element_text(size = 14, face = "bold"),
  legend.text  = element_text(size = 12, face = "bold"),
  legend.key   = element_blank(),

  # Facet strips (if present)
  strip.text = element_text(size = 12, face = "bold"),

  # Panel + background
  panel.grid = element_blank(),
  panel.border = element_rect(color = "black", size = .5, fill = NA),
  axis.line = element_blank(),

  # Margins (keeps q-values visible)
  plot.margin = margin(10, 10, 10, 10)
)

absrel_clean <- filter_one2manyID(absrel_clean, "gene")

absrel_clean <- exorcise(absrel_clean, "gene")

# ---- user knobs ----
BASELINE_CAP   <- 100
SIG_Q_THRESH   <- 0.10
TOP_N_PER_PATH <- 10
# --------------------

# 0) choose the corrected branch-level q column once
absrel_work <- absrel_clean

if ("q_global" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$q_global
} else if ("aBSREL_pvalue_corrected" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$aBSREL_pvalue_corrected
} else {
  stop("Need either q_global or aBSREL_pvalue_corrected in absrel_clean.")
}

# 1) parse + clean terminal branches, then keep positive branches only
branch_baseline <- absrel_work %>%
  filter(
    !is.na(branch_id),
    !str_detect(branch_id, "^NODE"),
    !is.na(aBSREL_baseline_omega),
    !is.na(pathway_label)
  ) %>%
  extract(
    col   = branch_id,
    into  = c("branch_gene", "species", "ensembl_tx", "ensembl_pep", "isoform_class"),
    regex = "^([^_]+)_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_([^_]+)_([^_]+)_(CANONICAL|ALTERNATIVE)$",
    remove = FALSE
  ) %>%
  mutate(
    species = coalesce(
      species,
      sub(".*_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_.*", "\\1", branch_id)
    ),
    species = toupper(species),
    sig = !is.na(q_use) & q_use <= SIG_Q_THRESH,
    omega_pos = aBSREL_baseline_omega,
    omega_cap = pmin(omega_pos, BASELINE_CAP),
    pos_branch = sig & !is.na(omega_pos) & omega_pos > 1
  ) %>%
  filter(pos_branch)

# optional sanity check
branch_baseline %>%
  summarise(
    n_positive_branches   = n(),
    min_positive_omega    = min(omega_pos, na.rm = TRUE),
    median_positive_omega = median(omega_pos, na.rm = TRUE),
    max_positive_omega    = max(omega_pos, na.rm = TRUE)
  )

# ============================================================
# PANEL A
# winning species per pathway
# ============================================================

pathway_species_sum <- branch_baseline %>%
  group_by(pathway_label, species) %>%
  summarise(
    species_contrib = sum(omega_cap, na.rm = TRUE),
    median_omega_cap = median(omega_cap, na.rm = TRUE),
    max_omega_cap = max(omega_cap, na.rm = TRUE),
    n_pos_branches = n(),
    min_q_species = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_species = ifelse(is.infinite(min_q_species), NA_real_, min_q_species)
  )

winner_species_by_pathway <- pathway_species_sum %>%
  group_by(pathway_label) %>%
  slice_max(order_by = species_contrib, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(species_contrib) %>%
  mutate(
    pathway_label = factor(pathway_label, levels = pathway_label),
    species_label = if_else(
      !is.na(min_q_species) & min_q_species < 0.05,
      paste0(species, " q=", signif(min_q_species, 2)),
      species
    )
  )

#removes label next to q value

# winner_species_by_pathway <- winner_species_by_pathway %>%
#   mutate(
#     species_label = if_else(
#       !is.na(min_q_species) & min_q_species < 0.05,
#       paste0("q=", signif(min_q_species, 2)),
#       ""
#     )
#   )


# remove q scores

winner_species_by_pathway <- winner_species_by_pathway %>%
  mutate(
    species_label = species   
  )

#reorder labels

winner_species_by_pathway$pathway_label <- factor(winner_species_by_pathway$pathway_label, levels = pathway_order)


# title = "A. Winning species per pathway"


p_species_A <- ggplot(
  winner_species_by_pathway,
  aes(x = species_contrib, y = pathway_label, color = species, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = species_contrib, yend = pathway_label),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = species_label),
    size = 3.2,
    color = "#3C3C3C",
    hjust= - 0.12,
    fontface = "bold"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    x = paste0("Summed capped branch \u03C9 by species (positive branches only; cap = ", BASELINE_CAP, ")"),
    y = "Functional Category",
    color = "Winning species",
    size = "Positive\nbranches",
    # title = "A. Winning species per pathway"
  ) +
  scale_color_manual(values = species_palette)  +
  scale_x_continuous(
  limits = c(0, max(winner_species_by_pathway$species_contrib) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
    #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    plot.margin = margin(5.5, 120, 5.5, 5.5),
    legend.position = "right"
  )

p_species_A <- p_species_A +
  theme_bold_large1

# ============================================================
# PANEL B
# winning species for each top-branch gene within each pathway
# simplified rule:
#   1) highest represented fraction among positive branches (prop_obs)
#   2) tie-break by n_pos_branches
# integrated concentration summary in facet labels
# ============================================================

# top genes per pathway based on positive branches only
top_genes_species <- branch_baseline %>%
  group_by(pathway_label) %>%
  slice_max(order_by = omega_cap, n = TOP_N_PER_PATH, with_ties = TRUE) %>%
  ungroup() %>%
  distinct(pathway_label, gene)

# species-level collapse within each gene
gene_species_pos <- branch_baseline %>%
  semi_join(top_genes_species, by = c("pathway_label", "gene")) %>%
  group_by(pathway_label, gene, species) %>%
  summarise(
    n_pos_branches = n(),
    median_omega_cap = median(omega_cap, na.rm = TRUE),
    max_omega_cap = max(omega_cap, na.rm = TRUE),
    min_q_species = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_species = ifelse(is.infinite(min_q_species), NA_real_, min_q_species)
  )

# represented fraction within each pathway × gene
gene_species_norm <- gene_species_pos %>%
  group_by(pathway_label, gene) %>%
  mutate(
    total_pos_branches_gene = sum(n_pos_branches, na.rm = TRUE),
    prop_obs = n_pos_branches / total_pos_branches_gene,
    gene_pathway = paste(pathway_label, gene, sep = "___")
  ) %>%
  ungroup()

# winning species per gene
winner_species_per_gene <- gene_species_norm %>%
  group_by(pathway_label, gene, gene_pathway) %>%
  arrange(
    desc(prop_obs),
    desc(n_pos_branches),
    species,
    .by_group = TRUE
  ) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    species_label = if_else(
      !is.na(min_q_species) & min_q_species < 0.05,
      paste0(species, " q=", signif(min_q_species, 2)),
      species
    )
  )

# pathway-level positive-branch opportunity by species
pathway_species_opportunity <- branch_baseline %>%
  group_by(pathway_label, species) %>%
  summarise(
    n_pos_branches_path_species = n(),
    .groups = "drop"
  ) %>%
  group_by(pathway_label) %>%
  mutate(
    expected_share = n_pos_branches_path_species / sum(n_pos_branches_path_species)
  ) %>%
  ungroup()

# concentration summary for winner assignments within each pathway
pathway_winner_summary <- winner_species_per_gene %>%
  count(pathway_label, species, name = "winner_n") %>%
  group_by(pathway_label) %>%
  mutate(
    winner_share = winner_n / sum(winner_n)
  ) %>%
  ungroup() %>%
  left_join(
    pathway_species_opportunity,
    by = c("pathway_label", "species")
  ) %>%
  mutate(
    winner_enrichment = winner_share / expected_share
  )

pathway_panel_labels <- pathway_winner_summary %>%
  group_by(pathway_label) %>%
  arrange(desc(winner_share), species, .by_group = TRUE) %>%
  summarise(
    top_winner = first(species),
    top_winner_share = first(winner_share),
    top_winner_enrichment = first(winner_enrichment),
    top2_winner_share = sum(head(winner_share, 2)),
    effective_n_winners = 1 / sum(winner_share^2),
    .groups = "drop"
  ) %>%
  mutate(
    pathway_panel = paste0(
      pathway_label,
      "\nTop winner: ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity)",
      "\nTop2 share: ", scales::percent(top2_winner_share, accuracy = 1),
      "; Effective winners: ", round(effective_n_winners, 1)
    )
  )

winner_species_per_gene <- winner_species_per_gene %>%
  left_join(
    pathway_panel_labels %>% select(pathway_label, pathway_panel),
    by = "pathway_label"
  )

gene_levels_species <- winner_species_per_gene %>%
  group_by(pathway_label) %>%
  arrange(prop_obs, .by_group = TRUE) %>%
  ungroup() %>%
  pull(gene_pathway)

winner_species_per_gene <- winner_species_per_gene %>%
  mutate(
    gene_pathway = factor(gene_pathway, levels = unique(gene_levels_species)),
    pathway_panel = factor(pathway_panel, levels = unique(pathway_panel_labels$pathway_panel))
  )

#this removes species label by q value

# winner_species_per_gene <- winner_species_per_gene %>%
#   mutate(
#     species_label = if_else(
#       !is.na(min_q_species) & min_q_species < 0.05,
#       paste0("q=", signif(min_q_species, 2)),
#       ""
#     )
#   )





# remove q scores

winner_species_per_gene <- winner_species_per_gene %>%
  mutate(
    species_label = species   
  )


#reorder labels

winner_species_per_gene$pathway_label <- factor(winner_species_per_gene$pathway_label, levels = pathway_orderb)


#title


    # title = paste0(
    #   "B. Winning species for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-\u03C9 positive branch)"
    # )

p_species_B <- ggplot(
  winner_species_per_gene,
  aes(x = prop_obs, y = gene_pathway, color = species, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = prop_obs, yend = gene_pathway),
    color = "grey70",
    linewidth = 0.4
  ) +
  scale_color_manual(values = species_palette) +
  geom_point() +
  geom_text(
    aes(label = species_label),
    size = 2.9,
    color = "#3C3C3C",
    hjust= - 0.085,
    fontface = "bold"
  ) +
  facet_grid(pathway_label ~ ., scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Winning represented fraction among positive branches",
    y = "Gene",
    color = "Winning species",
    size = "Positive\nbranches",
    # title = paste0(
    #   "B. Winning species for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-\u03C9 positive branch)"
    # )
  ) +
  scale_x_continuous(
  limits = c(0, max(winner_species_per_gene$prop_obs) * 1.75),
  expand = expansion(mult = c(0, 0.15))
) +
    #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    strip.text.y = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    panel.spacing.y = unit(0.45, "lines"),
    plot.margin = margin(5.5, 140, 5.5, 5.5),
    legend.position = "right"
  )

p_species_B <- p_species_B +
theme_bold_large


ggsave(file.path(output_dir,"Fig4Ortho.A.P1_sp.tiff"), plot = p_species_A, height = 8, width = 12)
ggsave(file.path(output_dir,"Fig4Ortho.B.P2_sp.tiff"), plot = p_species_B, height = 8, width = 12)

p_species_A / p_species_B + plot_layout(heights = c(1, 2))

pathway_panel_labels


#plain-text quick-read version of the facet summaries
panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")

panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs Functional Category opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")


summary_text_plot <- patchwork::wrap_elements(
  grid::textGrob(
    panel_B_summary_text,
    x = 0, hjust = 0,
    y = 1, vjust = 1,
    gp = grid::gpar(fontsize = 9)
  )
)
summary_text_plot

writeLines(
  panel_B_summary_text,
  con = file.path(output_dir, "panel_B_summarySpecies.txt"),
  useBytes = TRUE
)


ggplot2::ggsave(file.path(output_dir,
  filename = "panel_B_summary_plot_species.png"),
  plot = summary_text_plot,
  width = 9,
  height = 1,
  dpi = 300
)


```




### FIGURE 4B: primate lineage comparisons


```{r}

clade_palette <- c(
  "Cercopithecoidea" = "#0072B2",  # deep blue
  "Hominoidea"        = "#D55E00", # vermillion
  "Platyrrhini"       = "#009E73", # bluish green
  "Strepsirrhini"     = "#CC79A7", # reddish purple
  "Tarsiiformes"      = "#F0E442"  # yellow
)


###clade-first companion, updated to match species view
# positive-branch-only version

BASELINE_CAP   <- 100
SIG_Q_THRESH   <- 0.10
TOP_N_PER_PATH <- 10

hominoidea <- c(
  "homo_sapiens","pan_troglodytes","pan_paniscus",
  "gorilla_gorilla","pongo_abelii","nomascus_leucogenys"
)

cercopithecoidea <- c(
  "chlorocebus_sabaeus","macaca_mulatta","macaca_fascicularis",
  "macaca_nemestrina","papio_anubis","colobus_angolensis_palliatus",
  "cercocebus_atys","theropithecus_gelada"
)

platyrrhini <- c(
  "callithrix_jacchus","saimiri_boliviensis_boliviensis","aotus_nancymaae"
)

strepsirrhini <- c(
  "microcebus_murinus","prolemur_simus","propithecus_coquereli"
)

tarsiiformes <- c("carlito_syrichta")

clade_sizes <- tibble(
  clade = c("Hominoidea","Cercopithecoidea","Platyrrhini","Strepsirrhini","Tarsiiformes"),
  n_species_clade = c(
    length(hominoidea),
    length(cercopithecoidea),
    length(platyrrhini),
    length(strepsirrhini),
    length(tarsiiformes)
  )
)

absrel_clean <- filter_one2manyID(absrel_clean, "gene")

absrel_clean <- exorcise(absrel_clean, "gene")

####parse retained absrel observations ONCE
#    and define POSITIVE branches only
branch_dat <- absrel_clean %>%
  filter(
    !is.na(branch_id),
    !str_detect(branch_id, "^NODE"),
    !is.na(aBSREL_baseline_omega),
    !is.na(pathway_label)
  ) %>%
  extract(
    col   = branch_id,
    into  = c("branch_gene", "species", "ensembl_tx", "ensembl_pep", "isoform_class"),
    regex = "^([^_]+)_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_([^_]+)_([^_]+)_(CANONICAL|ALTERNATIVE)$",
    remove = FALSE
  ) %>%
  mutate(
    species = coalesce(
      species,
      sub(".*_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_.*", "\\1", branch_id)
    ),
    species = tolower(species),
    clade = case_when(
      species %in% hominoidea ~ "Hominoidea",
      species %in% cercopithecoidea ~ "Cercopithecoidea",
      species %in% platyrrhini ~ "Platyrrhini",
      species %in% strepsirrhini ~ "Strepsirrhini",
      species %in% tarsiiformes ~ "Tarsiiformes",
      TRUE ~ "Other"
    ),
    q_use = dplyr::coalesce(q_global, aBSREL_pvalue_corrected),
    sig = !is.na(q_use) & q_use <= SIG_Q_THRESH,
    omega_pos = aBSREL_baseline_omega,
    omega_cap = pmin(omega_pos, BASELINE_CAP),
    pos_branch = sig & !is.na(omega_pos) & omega_pos > 1
  ) %>%
  filter(
    clade != "Other",
    pos_branch
  )

# PANEL A
# pathway-level winner using SUMMED capped branch omega by clade among pos branches
pathway_clade_sum <- branch_dat %>%
  group_by(pathway_label, clade) %>%
  summarise(
    clade_contrib = sum(omega_cap, na.rm = TRUE),
    median_omega_cap = median(omega_cap, na.rm = TRUE),
    max_omega_cap = max(omega_cap, na.rm = TRUE),
    n_pos_branches = n(),
    min_q_clade = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_clade = ifelse(is.infinite(min_q_clade), NA_real_, min_q_clade)
  )

winner_clade_by_pathway <- pathway_clade_sum %>%
  group_by(pathway_label) %>%
  slice_max(order_by = clade_contrib, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(clade_contrib) %>%
  mutate(
    pathway_label = factor(pathway_label, levels = pathway_label),
    label = if_else(
      !is.na(min_q_clade) & min_q_clade < 0.05,
      paste0(clade, " q=", signif(min_q_clade, 2)),
      clade
    )
  )

#removes label next to q value

# winner_clade_by_pathway <- winner_clade_by_pathway %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_clade) & min_q_clade < 0.05,
#       paste0("q=", signif(min_q_clade, 2)),
#       ""
#     )
#   )



# remove q scores

winner_clade_by_pathway <- winner_clade_by_pathway %>%
  mutate(
    label = clade 
  )
 
#reorder labels

winner_clade_by_pathway$pathway_label <- factor(winner_clade_by_pathway$pathway_label, levels = pathway_order)

#title

    # title = "A. Winning clade per pathway"



p1 <- ggplot(
  winner_clade_by_pathway,
  aes(x = clade_contrib, y = pathway_label, color = clade, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = clade_contrib, yend = pathway_label),
    color = "grey70", linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 3.4,
    color = "#3C3C3C",
    hjust= - 0.18,
    fontface = "bold"
  ) +
    scale_color_manual(values = clade_palette) +
  coord_cartesian(clip = "off") +
  labs(
    x = paste0("Summed capped branch ω by clade (positive branches only; cap = ", BASELINE_CAP, ")"),
    y = "Functional Category",
    color = "Winning clade",
    size = "Positive\nbranches",
    # title = "A. Winning clade per pathway"
  ) +
  scale_x_continuous(
  limits = c(0, max(winner_clade_by_pathway$clade_contrib) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
   #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    plot.margin = margin(5.5, 120, 5.5, 5.5),
    legend.position = "right"
  )

p1 <- p1 +
  theme_bold_large1
# PANEL B
# winning clade for each top-branch gene within each pathway
# integrated with pathway-level concentration summary


# top genes per pathway based on positive branches only
top_genes <- branch_dat %>%
  group_by(pathway_label) %>%
  slice_max(order_by = omega_cap, n = TOP_N_PER_PATH, with_ties = TRUE) %>%
  ungroup() %>%
  distinct(pathway_label, gene)

# species-level collapse within gene first
gene_species_clade <- branch_dat %>%
  semi_join(top_genes, by = c("pathway_label", "gene")) %>%
  group_by(pathway_label, gene, clade, species) %>%
  summarise(
    species_max_omega = max(omega_cap, na.rm = TRUE),
    n_pos_branches_species = n(),
    min_q_species = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_species = ifelse(is.infinite(min_q_species), NA_real_, min_q_species)
  )

# normalized clade score within gene
gene_clade_norm <- gene_species_clade %>%
  group_by(pathway_label, gene, clade) %>%
  summarise(
    n_species_obs = n_distinct(species),
    n_pos_branches = sum(n_pos_branches_species, na.rm = TRUE),
    median_omega_cap = median(species_max_omega, na.rm = TRUE),
    max_omega_cap = max(species_max_omega, na.rm = TRUE),
    min_q_clade = suppressWarnings(min(min_q_species, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_clade = ifelse(is.infinite(min_q_clade), NA_real_, min_q_clade)
  ) %>%
  left_join(clade_sizes, by = "clade") %>%
  mutate(
    prop_obs = n_species_obs / n_species_clade,
    gene_pathway = paste(pathway_label, gene, sep = "___")
  )

# winning clade per gene
winner_clade_per_gene <- gene_clade_norm %>%
  group_by(pathway_label, gene, gene_pathway) %>%
  arrange(
    desc(prop_obs),
    desc(n_pos_branches),
    clade,
    .by_group = TRUE
  ) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    label = if_else(
      !is.na(min_q_clade) & min_q_clade < 0.05,
      paste0(clade, " q=", signif(min_q_clade, 2)),
      clade
    )
  )

# pathway-level positive-branch opportunity by clade
pathway_clade_opportunity <- branch_dat %>%
  group_by(pathway_label, clade) %>%
  summarise(
    n_pos_branches_path_clade = n(),
    .groups = "drop"
  ) %>%
  group_by(pathway_label) %>%
  mutate(
    expected_share = n_pos_branches_path_clade / sum(n_pos_branches_path_clade)
  ) %>%
  ungroup()

# concentration summary for winner assignments within each pathway
pathway_winner_summary <- winner_clade_per_gene %>%
  count(pathway_label, clade, name = "winner_n") %>%
  group_by(pathway_label) %>%
  mutate(
    winner_share = winner_n / sum(winner_n)
  ) %>%
  ungroup() %>%
  left_join(
    pathway_clade_opportunity,
    by = c("pathway_label", "clade")
  ) %>%
  mutate(
    winner_enrichment = winner_share / expected_share
  )

pathway_panel_labels <- pathway_winner_summary %>%
  group_by(pathway_label) %>%
  arrange(desc(winner_share), clade, .by_group = TRUE) %>%
  summarise(
    top_winner = first(clade),
    top_winner_share = first(winner_share),
    top_winner_enrichment = first(winner_enrichment),
    top2_winner_share = sum(head(winner_share, 2)),
    effective_n_winners = 1 / sum(winner_share^2),
    .groups = "drop"
  ) %>%
  mutate(
    pathway_panel = paste0(
      pathway_label,
      "\nTop winner: ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity)",
      "\nTop2 share: ", scales::percent(top2_winner_share, accuracy = 1),
      "; Effective winners: ", round(effective_n_winners, 1)
    )
  )

winner_clade_per_gene <- winner_clade_per_gene %>%
  left_join(
    pathway_panel_labels %>% select(pathway_label, pathway_panel),
    by = "pathway_label"
  )

gene_levels <- winner_clade_per_gene %>%
  group_by(pathway_label) %>%
  arrange(prop_obs, .by_group = TRUE) %>%
  ungroup() %>%
  pull(gene_pathway)

winner_clade_per_gene <- winner_clade_per_gene %>%
  mutate(
    gene_pathway = factor(gene_pathway, levels = unique(gene_levels)),
    pathway_panel = factor(pathway_panel, levels = unique(pathway_panel_labels$pathway_panel))
  )

#this removes clade name from q value label

# winner_clade_per_gene <- winner_clade_per_gene %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_clade) & min_q_clade < 0.05,
#       paste0("q=", signif(min_q_clade, 2)),
#       ""
#     )
#   )


# remove q scores

winner_clade_per_gene <- winner_clade_per_gene %>%
  mutate(
    label = clade 
  )
 
#reorder labels

winner_clade_per_gene$pathway_label <- factor(winner_clade_per_gene$pathway_label, levels = pathway_orderb)

#title

    # title = paste0(
    #   "B. Winning clade for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-ω positive branch)"
    # )



p2 <- ggplot(
  winner_clade_per_gene,
  aes(x = prop_obs, y = gene_pathway, color = clade, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = prop_obs, yend = gene_pathway),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 2.9,
    color = "#3C3C3C",
    hjust= - 0.11,
    fontface = "bold"
  ) +
    scale_color_manual(values = clade_palette) +
  facet_grid(pathway_label ~ ., scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Winning represented fraction among positive branches",
    y = "Gene",
    color = "Winning clade",
    size = "Positive\nbranches",
    # title = paste0(
    #   "B. Winning clade for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-ω positive branch)"
    # )
  ) +
  scale_x_continuous(
  limits = c(0, max(winner_clade_per_gene$prop_obs) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
  #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    strip.text.y = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    panel.spacing.y = unit(0.45, "lines"),
    plot.margin = margin(5.5, 140, 5.5, 5.5),
    legend.position = "right"
  )

p2 <- p2 +
theme_bold_large 

ggsave(file.path(output_dir,"Fig4Ortho.C.P1_Clade.tiff"), plot = p1, height = 8, width = 12)
ggsave(file.path(output_dir,"Fig4Ortho.D.P2_Clade.tiff"), plot = p2, height = 8, width = 12)


p1 / p2 + plot_layout(heights = c(1, 2))

pathway_panel_labels


# plain-text quick-read version of the facet summaries
panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")

panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs Functional Category opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")


summary_text_plot <- patchwork::wrap_elements(
  grid::textGrob(
    panel_B_summary_text,
    x = 0, hjust = 0,
    y = 1, vjust = 1,
    gp = grid::gpar(fontsize = 9)
  )
)
summary_text_plot

writeLines(
  panel_B_summary_text,
  con = file.path(output_dir, "panel_B_summaryCLADE.txt"),
  useBytes = TRUE
)


ggplot2::ggsave(file.path(output_dir,
  filename = "panel_B_summary_plot_clade.png"),
  plot = summary_text_plot,
  width = 9,
  height = 1,
  dpi = 300
)


```



### FIGURE 4C: diet


```{r}
#diet-group companion, winner-style and positive-branch-only

diet_pallette <- c ("Omnivore" = "blue", "Folivore" = "green" , "", "Frugivore" = "orange", "Carnivore" = "pink")

BASELINE_CAP   <- 100
SIG_Q_THRESH   <- 0.10
TOP_N_PER_PATH <- 10

omni_species <- c(
  "MACACA_MULATTA",
  "MACACA_FASCICULARIS",
  "MACACA_NEMESTRINA",
  "PAPIO_ANUBIS",
  "CERCOCEBUS_ATYS",
  "CHLOROCEBUS_SABAEUS",
  "MICROCEBUS_MURINUS",
  "SAIMIRI_BOLIVIENSIS_BOLIVIENSIS",
  "CALLITHRIX_JACCHUS",
  "HOMO_SAPIENS"
)

foli_species <- c(
  "COLOBUS_ANGOLENSIS_PALLIATUS",
  "GORILLA_GORILLA",
  "PROPITHECUS_COQUERELI",
  "PROLEMUR_SIMUS",
  "THEROPITHECUS_GELADA"
)

frugi_species <- c(
  "PONGO_ABELII",
  "AOTUS_NANCYMAAE",
  "NOMASCUS_LEUCOGENYS",
  "PAN_TROGLODYTES",
  "PAN_PANISCUS"
)

carni_species <- c(
  "CARLITO_SYRICHTA"
)

diet_sizes <- tibble(
  diet_group = c("Omnivore", "Folivore", "Frugivore", "Carnivore"),
  n_species_group = c(
    length(omni_species),
    length(foli_species),
    length(frugi_species),
    length(carni_species)
  )
)

absrel_clean <- filter_one2manyID(absrel_clean, "gene")

absrel_clean <- exorcise(absrel_clean, "gene")

##choose coorexcted qval column
absrel_work <- absrel_clean

if ("q_global" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$q_global
} else if ("aBSREL_pvalue_corrected" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$aBSREL_pvalue_corrected
} else {
  stop("Need either q_global or aBSREL_pvalue_corrected in absrel_clean.")
}


#### parse retained observations ONCE and define POSITIVE branches only
branch_dat <- absrel_work %>%
  filter(
    !is.na(branch_id),
    !str_detect(branch_id, "^NODE"),
    !is.na(aBSREL_baseline_omega),
    !is.na(pathway_label)
  ) %>%
  extract(
    col   = branch_id,
    into  = c("branch_gene", "species", "ensembl_tx", "ensembl_pep", "isoform_class"),
    regex = "^([^_]+)_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_([^_]+)_([^_]+)_(CANONICAL|ALTERNATIVE)$",
    remove = FALSE
  ) %>%
  mutate(
    species = coalesce(
      species,
      sub(".*_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_.*", "\\1", branch_id)
    ),
    species = toupper(species),
    diet_group = case_when(
      species %in% omni_species  ~ "Omnivore",
      species %in% foli_species  ~ "Folivore",
      species %in% frugi_species ~ "Frugivore",
      species %in% carni_species ~ "Carnivore",
      TRUE ~ "Other"
    ),
    sig = !is.na(q_use) & q_use <= SIG_Q_THRESH,
    omega_pos = aBSREL_baseline_omega,
    omega_cap = pmin(omega_pos, BASELINE_CAP),
    pos_branch = sig & !is.na(omega_pos) & omega_pos > 1
  ) %>%
  filter(
    diet_group != "Other",
    pos_branch
  )

# PANEL A
# pathway-level winner using SUMMED capped branch omega by diet group
# among positive branches only

pathway_diet_sum <- branch_dat %>%
  group_by(pathway_label, diet_group) %>%
  summarise(
    group_contrib = sum(omega_cap, na.rm = TRUE),
    median_omega_cap = median(omega_cap, na.rm = TRUE),
    max_omega_cap = max(omega_cap, na.rm = TRUE),
    n_pos_branches = n(),
    min_q_group = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_group = ifelse(is.infinite(min_q_group), NA_real_, min_q_group)
  )

winner_diet_by_pathway <- pathway_diet_sum %>%
  group_by(pathway_label) %>%
  slice_max(order_by = group_contrib, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(group_contrib) %>%
  mutate(
    pathway_label = factor(pathway_label, levels = pathway_label),
    label = if_else(
      !is.na(min_q_group) & min_q_group < 0.05,
      paste0(diet_group, " q=", signif(min_q_group, 2)),
      diet_group
    )
  )

#removes label next to q value

# winner_diet_by_pathway <- winner_diet_by_pathway %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_group) & min_q_group < 0.05,
#       paste0("q=", signif(min_q_group, 2)),
#       ""
#     )
#   )


# remove q scores

winner_diet_by_pathway <- winner_diet_by_pathway %>%
  mutate(
    label = diet_group  
  )
 
#reorder labels

winner_diet_by_pathway$pathway_label <- factor(winner_diet_by_pathway$pathway_label, levels = pathway_order)

#title

  # title = "A. Winning diet group per pathway"


p1 <- ggplot(
  winner_diet_by_pathway,
  aes(x = group_contrib, y = pathway_label, color = diet_group, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = group_contrib, yend = pathway_label),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 3.4,
    color = "#3C3C3C",
    hjust= - 0.15,
    fontface = "bold"
  ) +
  coord_cartesian(clip = "off") +
   scale_color_manual(values = diet_pallette) +
  labs(
    x = paste0("Summed capped branch \u03C9 by diet group (positive branches only; cap = ", BASELINE_CAP, ")"),
    y = "Functional Category",
    color = "Winning diet group",
    size = "Positive\nbranches",
    # title = "A. Winning diet group per pathway"
  ) +
  scale_x_continuous(
  limits = c(0, max(winner_diet_by_pathway$group_contrib) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
   #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    plot.margin = margin(5.5, 120, 5.5, 5.5),
    legend.position = "right"
  )

p1 <- p1 +
 theme_bold_large1

# PANEL B
# winning diet group for each top-branch gene within each pathway
# integrated with pathway-level concentration summary
# top genes per pathway based on positive branches only
top_genes <- branch_dat %>%
  group_by(pathway_label) %>%
  slice_max(order_by = omega_cap, n = TOP_N_PER_PATH, with_ties = TRUE) %>%
  ungroup() %>%
  distinct(pathway_label, gene)

# species-level collapse within gene first
gene_species_diet <- branch_dat %>%
  semi_join(top_genes, by = c("pathway_label", "gene")) %>%
  group_by(pathway_label, gene, diet_group, species) %>%
  summarise(
    species_max_omega = max(omega_cap, na.rm = TRUE),
    n_pos_branches_species = n(),
    min_q_species = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_species = ifelse(is.infinite(min_q_species), NA_real_, min_q_species)
  )

# normalized diet-group representation within gene
gene_diet_norm <- gene_species_diet %>%
  group_by(pathway_label, gene, diet_group) %>%
  summarise(
    n_species_obs = n_distinct(species),
    n_pos_branches = sum(n_pos_branches_species, na.rm = TRUE),
    median_omega_cap = median(species_max_omega, na.rm = TRUE),
    max_omega_cap = max(species_max_omega, na.rm = TRUE),
    min_q_group = suppressWarnings(min(min_q_species, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_group = ifelse(is.infinite(min_q_group), NA_real_, min_q_group)
  ) %>%
  left_join(diet_sizes, by = "diet_group") %>%
  mutate(
    prop_obs = n_species_obs / n_species_group,
    gene_pathway = paste(pathway_label, gene, sep = "___")
  )

# winning diet group per gene
winner_diet_per_gene <- gene_diet_norm %>%
  group_by(pathway_label, gene, gene_pathway) %>%
  arrange(
    desc(prop_obs),
    desc(n_pos_branches),
    diet_group,
    .by_group = TRUE
  ) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    label = if_else(
      !is.na(min_q_group) & min_q_group < 0.05,
      paste0(diet_group, " q=", signif(min_q_group, 2)),
      diet_group
    )
  )

# pathway-level positive-branch opportunity by diet group
pathway_diet_opportunity <- branch_dat %>%
  group_by(pathway_label, diet_group) %>%
  summarise(
    n_pos_branches_path_group = n(),
    .groups = "drop"
  ) %>%
  group_by(pathway_label) %>%
  mutate(
    expected_share = n_pos_branches_path_group / sum(n_pos_branches_path_group)
  ) %>%
  ungroup()

# concentration summary for winner assignments within each pathway
pathway_winner_summary <- winner_diet_per_gene %>%
  count(pathway_label, diet_group, name = "winner_n") %>%
  group_by(pathway_label) %>%
  mutate(
    winner_share = winner_n / sum(winner_n)
  ) %>%
  ungroup() %>%
  left_join(
    pathway_diet_opportunity,
    by = c("pathway_label", "diet_group")
  ) %>%
  mutate(
    winner_enrichment = winner_share / expected_share
  )

pathway_panel_labels <- pathway_winner_summary %>%
  group_by(pathway_label) %>%
  arrange(desc(winner_share), diet_group, .by_group = TRUE) %>%
  summarise(
    top_winner = first(diet_group),
    top_winner_share = first(winner_share),
    top_winner_enrichment = first(winner_enrichment),
    top2_winner_share = sum(head(winner_share, 2)),
    effective_n_winners = 1 / sum(winner_share^2),
    .groups = "drop"
  ) %>%
  mutate(
    pathway_panel = paste0(
      pathway_label,
      "\nTop winner: ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity)",
      "\nTop2 share: ", scales::percent(top2_winner_share, accuracy = 1),
      "; Effective winners: ", round(effective_n_winners, 1)
    )
  )

winner_diet_per_gene <- winner_diet_per_gene %>%
  left_join(
    pathway_panel_labels %>% select(pathway_label, pathway_panel),
    by = "pathway_label"
  )

gene_levels <- winner_diet_per_gene %>%
  group_by(pathway_label) %>%
  arrange(prop_obs, .by_group = TRUE) %>%
  ungroup() %>%
  pull(gene_pathway)

winner_diet_per_gene <- winner_diet_per_gene %>%
  mutate(
    gene_pathway = factor(gene_pathway, levels = unique(gene_levels)),
    pathway_panel = factor(pathway_panel, levels = unique(pathway_panel_labels$pathway_panel))
  )

# this removes diet label next to q value

# winner_diet_per_gene <- winner_diet_per_gene %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_group) & min_q_group < 0.05,
#       paste0("q=", signif(min_q_group, 2)),
#       ""
#     )
#   )



# remove q scores

winner_diet_per_gene <- winner_diet_per_gene %>%
  mutate(
    label = diet_group  
  )
 
#reorder labels

winner_diet_per_gene$pathway_label <- factor(winner_diet_per_gene$pathway_label, levels = pathway_orderb)

#title

    # title = paste0(
    #   "B. Winning diet group for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-\u03C9 positive branch)"
    # )


p2 <- ggplot(
  winner_diet_per_gene,
  aes(x = prop_obs, y = gene_pathway, color = diet_group, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = prop_obs, yend = gene_pathway),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 3.2,
    color = "#3C3C3C",
    hjust= - 0.15,
    fontface = "bold"
  ) +
  facet_grid(pathway_label ~ ., scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
  coord_cartesian(clip = "off") +
  scale_color_manual(values = diet_pallette) +
  labs(
    x = "Winning share of positive branches within gene",
    y = "Gene",
    color = "Winning diet group",
    size = "Positive\nbranches",
    # title = paste0(
    #   "B. Winning diet group for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-\u03C9 positive branch)"
    # )
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
  limits = c(0, max(winner_diet_per_gene$prop_obs) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
    #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    strip.text.y = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    panel.spacing.y = unit(0.45, "lines"),
    plot.margin = margin(5.5, 140, 5.5, 5.5),
    legend.position = "right"
  )

p2 <- p2 +
  theme_bold_large

ggsave(file.path(output_dir,"Fig4Ortho.E.P1_Diet.tiff"), plot = p1, height = 8, width = 12)
ggsave(file.path(output_dir,"Fig4Ortho.F.P2_Diet.tiff"), plot = p2, height = 8, width = 12)

p1 / p2 + plot_layout(heights = c(1, 2))

pathway_panel_labels



# plain-text quick-read version of the facet summaries
panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")

panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs Functional Category opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")


summary_text_plot <- patchwork::wrap_elements(
  grid::textGrob(
    panel_B_summary_text,
    x = 0, hjust = 0,
    y = 1, vjust = 1,
    gp = grid::gpar(fontsize = 9)
  )
)
summary_text_plot

writeLines(
  panel_B_summary_text,
  con = file.path(output_dir, "panel_B_summaryDIET.txt"),
  useBytes = TRUE
)


ggplot2::ggsave(file.path(output_dir,
  filename = "panel_B_summary_plot_diet.png"),
  plot = summary_text_plot,
  width = 9,
  height = 1,
  dpi = 300
)


```


### FIGURE 4D: EQ



```{r}

BASELINE_CAP   <- 100
SIG_Q_THRESH   <- 0.10
TOP_N_PER_PATH <- 10

EQ_pallette <- c("EQlow" = "red", "EQmed" = "blue", "EQhigh" = "green")

EQlow <- c(
  "COLOBUS_ANGOLENSIS_PALLIATUS",
  "MICROCEBUS_MURINUS",
  "AOTUS_NANCYMAAE",
  "THEROPITHECUS_GELADA",
  "CALLITHRIX_JACCHUS",
  "CARLITO_SYRICHTA",
  "PROPITHECUS_COQUERELI"
)

EQmed <- c(
  "MACACA_MULATTA",
  "MACACA_FASCICULARIS",
  "MACACA_NEMESTRINA",
  "PROLEMUR_SIMUS",
  "GORILLA_GORILLA",
  "CERCOCEBUS_ATYS",
  "CHLOROCEBUS_SABAEUS"
)

EQhigh <- c(
  "HOMO_SAPIENS",
  "SAIMIRI_BOLIVIENSIS_BOLIVIENSIS",
  "NOMASCUS_LEUCOGENYS",
  "PAN_TROGLODYTES",
  "PAN_PANISCUS",
  "PONGO_ABELII",
  "PAPIO_ANUBIS"
)


eq_sizes <- tibble(
  eq_group = c("EQlow", "EQmed", "EQhigh"),
  n_species_group = c(length(EQlow), length(EQmed), length(EQhigh))
)

absrel_clean <- filter_one2manyID(absrel_clean, "gene")

absrel_clean <- exorcise(absrel_clean, "gene")

##choose corrected q column once
absrel_work <- absrel_clean

if ("q_global" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$q_global
} else if ("aBSREL_pvalue_corrected" %in% names(absrel_work)) {
  absrel_work$q_use <- absrel_work$aBSREL_pvalue_corrected
} else {
  stop("Need either q_global or aBSREL_pvalue_corrected in absrel_clean.")
}


###parse retained observations ONCE and define POSITIVE branches only
branch_dat <- absrel_work %>%
  filter(
    !is.na(branch_id),
    !str_detect(branch_id, "^NODE"),
    !is.na(aBSREL_baseline_omega),
    !is.na(pathway_label)
  ) %>%
  extract(
    col   = branch_id,
    into  = c("branch_gene", "species", "ensembl_tx", "ensembl_pep", "isoform_class"),
    regex = "^([^_]+)_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_([^_]+)_([^_]+)_(CANONICAL|ALTERNATIVE)$",
    remove = FALSE
  ) %>%
  mutate(
    species = coalesce(
      species,
      sub(".*_([A-Za-z]+_[A-Za-z]+(?:_[A-Za-z]+)?)_.*", "\\1", branch_id)
    ),
    species = toupper(species),
    species_tail = sub("^[^_]+_", "", species),
    eq_group = case_when(
      species %in% EQlow  | species_tail %in% EQlow  ~ "EQlow",
      species %in% EQmed  | species_tail %in% EQmed  ~ "EQmed",
      species %in% EQhigh | species_tail %in% EQhigh ~ "EQhigh",
      TRUE ~ "Other"
    ),
    sig = !is.na(q_use) & q_use <= SIG_Q_THRESH,
    omega_pos = aBSREL_baseline_omega,
    omega_cap = pmin(omega_pos, BASELINE_CAP),
    pos_branch = sig & !is.na(omega_pos) & omega_pos > 1
  ) %>%
  filter(
    eq_group != "Other",
    pos_branch
  )

#######pANEL A
#pathway-level winner using SUMMED capped branch omega by EQ group
#among positive branches only


pathway_eq_sum <- branch_dat %>%
  group_by(pathway_label, eq_group) %>%
  summarise(
    group_contrib = sum(omega_cap, na.rm = TRUE),
    median_omega_cap = median(omega_cap, na.rm = TRUE),
    max_omega_cap = max(omega_cap, na.rm = TRUE),
    n_pos_branches = n(),
    min_q_group = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_group = ifelse(is.infinite(min_q_group), NA_real_, min_q_group)
  )

winner_eq_by_pathway <- pathway_eq_sum %>%
  group_by(pathway_label) %>%
  slice_max(order_by = group_contrib, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(group_contrib) %>%
  mutate(
    pathway_label = factor(pathway_label, levels = pathway_label),
    label = if_else(
      !is.na(min_q_group) & min_q_group < 0.05,
      paste0(eq_group, " q=", signif(min_q_group, 2)),
      eq_group
    )
  )

#removes label next to q value

# winner_eq_by_pathway <- winner_eq_by_pathway %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_group) & min_q_group < 0.05,
#       paste0("q=", signif(min_q_group, 2)),
#       ""
#     )
#   )


# remove q scores

winner_eq_by_pathway <- winner_eq_by_pathway %>%
  mutate(
    label = eq_group  
  )
 
#reorder labels

winner_eq_by_pathway$pathway_label <- factor(winner_eq_by_pathway$pathway_label, levels = pathway_order)

#title

  # title = "A. Winning EQ group per pathway"

p1 <- ggplot(
  winner_eq_by_pathway,
  aes(x = group_contrib, y = pathway_label, color = eq_group, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = group_contrib, yend = pathway_label),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 3.5,
    color = "#3C3C3C",
    hjust= - 0.2,
    fontface = "bold"
  ) +
  coord_cartesian(clip = "off") +
     scale_color_manual(values = EQ_pallette) +
  labs(
    x = paste0("Summed capped branch ω by EQ group (positive branches only; cap = ", BASELINE_CAP, ")"),
    y = "Functional Category",
    color = "Winning EQ group",
    size = "Positive\nbranches",
    # title = "A. Winning EQ group per pathway"
  ) +
  scale_x_continuous(
  limits = c(0, max(winner_eq_by_pathway$group_contrib) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
   #remove the color legend 
  guides(color = "none") +
  theme_bw() +
  theme(
    plot.margin = margin(5.5, 120, 5.5, 5.5),
    legend.position = "right"
  )

p1 <- p1 +
  theme_bold_large1

######PANEL B
##### winning () EQ group for each top-branch gene within each pathway
#integrated with pathway-level concentration summary


#####top genes per pathway based on positive branches only
top_genes <- branch_dat %>%
  group_by(pathway_label) %>%
  slice_max(order_by = omega_cap, n = TOP_N_PER_PATH, with_ties = TRUE) %>%
  ungroup() %>%
  distinct(pathway_label, gene)

####species-level collapse within gene first
gene_species_eq <- branch_dat %>%
  semi_join(top_genes, by = c("pathway_label", "gene")) %>%
  group_by(pathway_label, gene, eq_group, species) %>%
  summarise(
    species_max_omega = max(omega_cap, na.rm = TRUE),
    n_pos_branches_species = n(),
    min_q_species = suppressWarnings(min(q_use, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_species = ifelse(is.infinite(min_q_species), NA_real_, min_q_species)
  )

#####normalized EQ-group representation within gene
gene_eq_norm <- gene_species_eq %>%
  group_by(pathway_label, gene, eq_group) %>%
  summarise(
    n_species_obs = n_distinct(species),
    n_pos_branches = sum(n_pos_branches_species, na.rm = TRUE),
    median_omega_cap = median(species_max_omega, na.rm = TRUE),
    max_omega_cap = max(species_max_omega, na.rm = TRUE),
    min_q_group = suppressWarnings(min(min_q_species, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    min_q_group = ifelse(is.infinite(min_q_group), NA_real_, min_q_group)
  ) %>%
  left_join(eq_sizes, by = "eq_group") %>%
  mutate(
    prop_obs = n_species_obs / n_species_group,
    gene_pathway = paste(pathway_label, gene, sep = "___")
  )

#######winning EQ group per gene
winner_eq_per_gene <- gene_eq_norm %>%
  group_by(pathway_label, gene, gene_pathway) %>%
  arrange(
    desc(prop_obs),
    desc(n_pos_branches),
    eq_group,
    .by_group = TRUE
  ) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    label = if_else(
      !is.na(min_q_group) & min_q_group < 0.05,
      paste0(eq_group, " q=", signif(min_q_group, 2)),
      eq_group
    )
  )

#####pathway-level positive-branch opportunity by EQ group
pathway_eq_opportunity <- branch_dat %>%
  group_by(pathway_label, eq_group) %>%
  summarise(
    n_pos_branches_path_group = n(),
    .groups = "drop"
  ) %>%
  group_by(pathway_label) %>%
  mutate(
    expected_share = n_pos_branches_path_group / sum(n_pos_branches_path_group)
  ) %>%
  ungroup()

####concentration summary for winner assignments within each pathway
pathway_winner_summary <- winner_eq_per_gene %>%
  count(pathway_label, eq_group, name = "winner_n") %>%
  group_by(pathway_label) %>%
  mutate(
    winner_share = winner_n / sum(winner_n)
  ) %>%
  ungroup() %>%
  left_join(
    pathway_eq_opportunity,
    by = c("pathway_label", "eq_group")
  ) %>%
  mutate(
    winner_enrichment = winner_share / expected_share
  )

pathway_panel_labels <- pathway_winner_summary %>%
  group_by(pathway_label) %>%
  arrange(desc(winner_share), eq_group, .by_group = TRUE) %>%
  summarise(
    top_winner = first(eq_group),
    top_winner_share = first(winner_share),
    top_winner_enrichment = first(winner_enrichment),
    top2_winner_share = sum(head(winner_share, 2)),
    effective_n_winners = 1 / sum(winner_share^2),
    .groups = "drop"
  ) %>%
  mutate(
    pathway_panel = paste0(
      pathway_label,
      "\nTop winner: ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "× vs pathway opportunity)",
      "\nTop2 share: ", scales::percent(top2_winner_share, accuracy = 1),
      "; Effective winners: ", round(effective_n_winners, 1)
    )
  )

winner_eq_per_gene <- winner_eq_per_gene %>%
  left_join(
    pathway_panel_labels %>% select(pathway_label, pathway_panel),
    by = "pathway_label"
  )

gene_levels <- winner_eq_per_gene %>%
  group_by(pathway_label) %>%
  arrange(prop_obs, .by_group = TRUE) %>%
  ungroup() %>%
  pull(gene_pathway)

winner_eq_per_gene <- winner_eq_per_gene %>%
  mutate(
    gene_pathway = factor(gene_pathway, levels = unique(gene_levels)),
    pathway_panel = factor(pathway_panel, levels = unique(pathway_panel_labels$pathway_panel))
  )

#this removes eq label next to q value

# winner_eq_per_gene <- winner_eq_per_gene %>%
#   mutate(
#     label = if_else(
#       !is.na(min_q_group) & min_q_group < 0.05,
#       paste0("q=", signif(min_q_group, 2)),
#       ""
#     )
#   )

# remove q scores

winner_eq_per_gene <- winner_eq_per_gene %>%
  mutate(
    label = eq_group  
  )
 
#reorder labels

winner_eq_per_gene$pathway_label <- factor(winner_eq_per_gene$pathway_label, levels = pathway_orderb)


#title

    # title = paste0(
    #   "B. Winning EQ group for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-ω positive branch)"
    # )



p2 <- ggplot(
  winner_eq_per_gene,
  aes(x = prop_obs, y = gene_pathway, color = eq_group, size = n_pos_branches)
) +
  geom_segment(
    aes(x = 0, xend = prop_obs, yend = gene_pathway),
    color = "grey70",
    linewidth = 0.4
  ) +
  geom_point() +
  geom_text(
    aes(label = label),
    size = 2.9,
    color = "#3C3C3C",
    hjust= - 0.13,
    fontface = "bold"
  ) +
  facet_grid(pathway_label ~ ., scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
  coord_cartesian(clip = "off") +
      scale_color_manual(values = EQ_pallette) +
  labs(
    x = "Winning share of positive branches within gene",
    y = "Gene",
    color = "Winning EQ group",
    size = "Positive\nbranches",
    # title = paste0(
    #   "B. Winning EQ group for each top-branch gene within each pathway",
    #   " (genes containing a top ", TOP_N_PER_PATH, " capped-ω positive branch)"
    # )
  ) +
  theme_bw() +
  scale_x_continuous(
  limits = c(0, max(winner_eq_per_gene$prop_obs) * 1.25),
  expand = expansion(mult = c(0, 0.02))
) +
    #remove the color legend 
  guides(color = "none") +
  theme(
    strip.text.y = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    panel.spacing.y = unit(0.45, "lines"),
    plot.margin = margin(5.5, 140, 5.5, 5.5),
    legend.position = "right"
  )

p2 <- p2 +
 theme_bold_large

ggsave(file.path(output_dir,"Fig4Ortho.G.P1_EQ.tiff"), plot = p1, height = 8, width = 12)
ggsave(file.path(output_dir,"Fig4Ortho.H.P2_EQ.tiff"), plot = p2, height = 8, width = 12)

p1 / p2 + plot_layout(heights = c(1, 2))

pathway_panel_labels


# SUMMARY TABLE 

# plain-text quick-read version of the facet summaries
panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs pathway opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")



panel_B_summary_text <- pathway_panel_labels %>%
  arrange(pathway_label) %>%
  transmute(
    line = paste0(
      pathway_label, ": ",
      "Top winner = ", top_winner,
      " (", scales::percent(top_winner_share, accuracy = 1), " of genes; ",
      round(top_winner_enrichment, 2), "\u00D7 vs Functional Category opportunity); ",
      "Top2 share = ", scales::percent(top2_winner_share, accuracy = 1), "; ",
      "Effective winners = ", round(effective_n_winners, 1)
    )
  ) %>%
  pull(line) %>%
  paste(collapse = "\n")


summary_text_plot <- patchwork::wrap_elements(
  grid::textGrob(
    panel_B_summary_text,
    x = 0, hjust = 0,
    y = 1, vjust = 1,
    gp = grid::gpar(fontsize = 9)
  )
)

summary_text_plot
 

writeLines(
  panel_B_summary_text,
  con = file.path(output_dir, "panel_B_summaryEQ.txt"),
  useBytes = TRUE
)


ggplot2::ggsave(file.path(output_dir,
  filename = "panel_B_summary_plot_EQ.png"),
  plot = summary_text_plot,
  width = 9,
  height = 1,
  dpi = 300
)



```


